### 2023-10-09 manuscriptlongversion.Rmd
code

---
title: "Trauma treatment delay"
author: "Anton Wasielewski"
subtitle: A retrospective cohort study on the factors affecting the delays in trauma
  care
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
bibliography: bibliography.bib
editor_options:
  markdown:
    wrap: 72
---

# Introduction

**Definition**

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response [@gerdin_risk_2015], is and has long been
a major cause of death around the world today [@WHO_2021]. Within modern
hospital care, the trauma system stands as an important part in
healthcare and is crucial to lowering mortality and morbidity in injured
patients through many means, such as hospital care, follow up and
prevention programs [@Choi_2021].

## Trauma Statistics

Trauma is often divided into two subgroups, blunt force trauma and
penetrating trauma. Blunt force trauma is when an object or force
strikes the body, often causing bruising, broken bones or deep cuts.
Examples of blunt trauma could be car crashes, falls or direct blows to
the body. Penetrating trauma is when an object pierces the skin or body
and creates an open wound. Examples of penetrating trauma are gunshot
wounds and stab wounds [@NIGMS_2020]. Differences exist between the two
groups, as blunt trauma patients tend to be more injured on arrival to
the hospital and as such, require more resources and are hospitalized
for a longer periods [@FITCH2019].

Today, trauma takes the lives of around 4,4 million people each year,
almost 8% of all deaths [@WHO_2021]. In the United States, trauma is the
4th leading cause of death among the general population and the leading
cause for people between the ages of 1 and 44 [@AAST_2020]. It is also
an important cause of hospitalization and morbidity among all age
groups, including seniors, and is responsible for an estimated 10% of
all years lived with a disability globally. This has a significant
burden on social and economic level, costing countries billions of US
dollars each year in healthcare and lost productivity [@connolly_2018].
Studies estimate the cost of trauma care being between 18,500\$ and
41,500\$ per patient in high income countries (HIC), depending on the
country [@Willenberg_2012]. However, it has been shown that this burden
is not evenly distributed between or within countries. Many social
factors such as age, sex and social status play a major role in the risk
of dying from trauma, with young men with low socio-economic status
being at most risk [@WHO_2021]. But it is not only patient level factors
that affect what effect trauma has on people. About 90% of all trauma
related deaths occur in low- and middle-income countries (LMIC), with
death rates by trauma also being higher than in high income countries.
Even within these LMIC, people of poorer socio-economic status have
higher death rates from trauma. Problems identified within these
countries that contribute to these statisitics were infrastructure,
education and training, attributed to lack of funding, brain drain to
HIC and lack of availability of basic amenities [@Shanthakumar_2021]. In
HIC where funding and governance over the healthcare system is better
designed, functioning trauma systems and dedicated trauma centers exist
that have been shown to lower mortality but also improve functional
outcome in trauma patients [@Nirula_2006].

## Trauma systems

Trauma systems are infrastructures that exist to provide and optimize
care for injured patients starting with injury recognition and triage,
transport to appropriate trauma center, inpatient care and outpatient
follow-up. Beyond the clinical side, trauma systems work with outreach,
education and advocacy, data collection through registries, research,
funding, and disaster preparedness and response. A comprehensive and
functioning trauma system requires strong leadership and engagement at
the trauma center, regional and national level [@connolly_2018]. This
system is crucial to provide care for trauma patients, both in reducing
morbidity and mortality in this patient group. Earlier studies from
Sweden have shown that treating severely injured patients at a trauma
center is associated with a 41% lower adjusted 30-day mortality rate
compared to being treated at a non trauma care center due to them being
more capable of treating these patients, with potential survival benefit
increasing with higher injury severity [@Candefjord_2020]. Other studies
have shown similar results, showing that treating injured patients at
trauma centers is associated with a 15-30% decrease in mortality
[@Moran_2018][@Celso_2006][@MacKenzie_2006][@Haas_2012]. During the last
three decades, the introduction of trauma systems has contributed to
lowering the incidence of preventable death . This is attributed to the
improvements in care for acute brain injuries and bleeding control. The
incidence of late death because of sepsis and multiple organ failure,
possibly a result of better and earlier resuscitation [@Asensio_2008].
Functioning trauma systems saves lives in the hospital, but their work
outside of the hospital is just as important. Data collection from
injured patients, such as mechanism of injury or mortality, are
essential for creating databases that can be used for research. In turn,
that research can be used for planning injury prevention programs that
target the most common injuries in the right ways, e.g. teen drivers,
children, specific occupations etc. These injury prevention programs can
be planned on trauma center, organizational or government level[@ACS].

Trauma teams are multidisciplinary and operate in these centers. They
play a pivotal role in the treatment of the trauma patient, as they
provide the initial care in the critical stage of trauma. In Sweden,
trauma teams are lead by a team leader who is a surgeon, and include
practitioners from the specialities of intensive care, orthopedics,
nursing and support staff[@västra_G]. For the trauma team to be
mobilized, trauma code has to be activated, often by an emergency nurse.
The nurse uses information gathered by the first responders to see if
the patient fits any of the criteria for trauma code. Most healthcare
facilities have established criteria or guidelines that trigger trauma
code activation. These criteria typically include mechanisms of injury,
physiological criteria, anatomical criteria and other specific
indicators of severe trauma. There are different levels of trauma code
with different criteria correlating with severity of injury, with level
1 mobilizing the most personnel to the trauma room. Once in the trauma
room, the team works systematically to manage the patients injuries.
They handle the most urgent problems first, such as airways and
breathing, with the aim to rapidly assess and stabilise the patient,
prioritise their injuries and arrange for site of definitive care.
[@Hedberg_2020]

## Mortality and morbidity (MoM) conferences

Many hospitals have a trauma registry where they log the patients
information and timeline of what happened. The extent of the registry
varies, with HIC having more complete registries. This logging of
information is important work, as a cornerstone of trauma quality
improvement programs is multidisciplinary MoM conferences. The MoM
conference is a meeting where different specialities who work with
trauma care sit down and discuss deaths and complications in order to
look for preventable factors. These conferences are performed in many
hospitals globally, almost everywhere where there are formal medical
specialty departments and sometimes in smaller hospitals as well
[@WHO_2009]. In Sweden, many hospitals have mortality conferences on
deceased patient cases, but only one has MoM conferences to better their
trauma care on patients who lived [@SweTrau_2023]. The reason being that
this expansion of MoM conferences takes time and resources, which is why
all hospitals don't implement this.

The endpoint of these MoM conferences are opportunities for improvement
(OFI). At the end of the conference, consensus is reached regarding the
existence of an OFI and implementation of corrective action. This
process is effective, as it has been shown that this review is
associated with high-quality trauma care [@Santana_2014]. Examples of
OFI may include lack of resources and management errors. One common OFI
is delayed treatment. According to previous studies, among the
preventable or possibly preventable deaths in trauma patients, delay in
treatment has been identified as a major error contributing to death,
found in up to 52.9% of patients in said group [@Teixeira_2007]. Delayed
treatment has been shown to have adverse effects on patients, showing
why it is such an important issue and the need to develop strategies to
combat [@Sampalis_1995][@Sampalis_1994]. Although it is such a common
OFI, the patient level factors associated with delayed treatment are
poorly understood. There may be several factors that correlate with
receiving delayed treatment, but they have yet to be identified. Mapping
these factors may help in identifying patients that might be at risk for
receiving delayed treatment before it happens.

## Aim

The aim of this paper is to determine what different patient level
factors affect the risk of receiving delayed treatment at a trauma
center.

# Methods

**Setting**

Every year, Karolinska University hospital, which is the equivalent of a
trauma level 1 hospital, admits 1500 trauma patients each year. To be
added to the trauma registry, a patient must be over 15 years of age,
had ISS\>9 and/or had trauma code activation. They get added to both the
Karolinska Trauma registry as well as the national trauma registry
(SweTrau). The registry includes data on vital signs, injuries,
interventions and patient information. Another database, the trauma care
quality database includes information relevant for MoM conferences. At
Karolinska University Hospital, patients who die during hospital stay
gets their case sent to a MoM everytime. However, for the rest of the
patients, two specialized nurses determine whether they are. Several
audit filters are set up to flag for patients with possible OFIs.
Flagged patients are then reviewed by these two nurses who decide if
there is a possible OFI, in which case their case is sent to a MoM
conference for further review.

**Study design**

We conducted a registry based cohort study using data from the trauma
registry and trauma care quality database at the Karolinska University
Hospital in Solna. The trauma registry includes about 14000 patients
treated between 2012 and 2022. The trauma care quality database is a
subset of the trauma registry and includes about 8000 patients selected
for review between 2013 and 2022. This project will link the two
databases and assess how different patient level factors, such as age,
sex, mechanism of injury, and injury severity, are associated with
delayed treatment using logistic regression.

**Outcome**

The outcome is delayed treatment, as identified by the multidisciplinary
review board and recorded in the trauma care quality database.

**Participants**

The database only includes patients 15 years old and above, ISS \>9
and/or trauma team activation. Inclusion in the study further requires
that the patients have been assessed for OFIs. Patients were excluded if
they were missing information in one of the examined patient factors.

**Variables**

The outcome is delayed treatment, which is defined as more than 30
minutes passing from arrival before receiving treatment, as identified
by the multidisciplinary review board and recorded in the trauma care
quality database.

**Data sources/measurement**

The data comes from the trauma registry and trauma care quality
database.

**Bias**

Selection bias as this is a retrospective cohort study where the outcome
has occurred.

Misclassification bias as the conference may reach a consensus that is
wrong.

**Study size**

The study size is all of the eligible patients that was treated at the
hospital between 2012 and 2022. Starting 2013 only a subgroup of
patients were screened for OFIs, but starting 2017 all patients that are
included in the database are screened for OFIs.

**Statistical methods**

Multivariable and potentially univariable logistical regression. A 5%
significance level and 95% confidence levels will be used.

### 2023-10-09 example_function.R

@@ -18,3 +18,7 @@ cleaned_data <- subset(prepared.data, !is.na(Problemomrade_.FMP))
 prepared.data$ofi <- ifelse(prepared.data$ofi == "success", 1, 0)
 cleaned_data$Gender <- ifelse(cleaned_data$Gender == "M", 1, 0)
 
+
+
+output <- glm(formula = outcome ~ factor(var01) + factor (var02) + var03,
+              data=datasetname, family=binomial)

### 2023-10-11 example_function.R
code

@@ -1,24 +1,18 @@
 library(rofi)
 
-prepared.data <- prepare_data(data)
-
-summary(prepared.data)
-
 subset.datas <- prepared.data[, c("ofi", "Problemomrade_.FMP")]
 
 table(prepared.data$Problemomrade_.FMP)
 
-delayed_subset <- subset(prepared.data, Problemomrade_.FMP == "Handläggning")
-rest_subset <- subset(prepared.data, Problemomrade_.FMP != "Handläggning")
-
-prepared.data$Problemomrade_.FMP <- ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", 1, 0)
-
-cleaned_data <- subset(prepared.data, !is.na(Problemomrade_.FMP))
 
 prepared.data$ofi <- ifelse(prepared.data$ofi == "success", 1, 0)
 cleaned_data$Gender <- ifelse(cleaned_data$Gender == "M", 1, 0)
 
 
 
-output <- glm(formula = outcome ~ factor(var01) + factor (var02) + var03,
-              data=datasetname, family=binomial)
+  model <- glm(OFI_delay ~ pt_age_yrs + ISS + inj_mechanism + pt_Gender, data = cleaned_data, family = binomial)
+  
+  filtered.data <- subset(New.subset, "pt_age_yrs" + "ISS" + "inj_mechanism" + "pt_Gender", data != "999")
+  
+  prepared.data$OFI_delay <- as.factor(ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", "1", "0"))
+  

### 2023-10-11 functions/New_subset.R

@@ -1,0 +1,10 @@
+library(rofi)
+
+New_subset <- function(dataset){
+  
+  prepared.data$OFI_delay <- ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", "1", "0")
+
+  cleaned_data <- subset(prepared.data, !is.na(OFI_delay))
+
+  return(cleaned_data)
+}

### 2023-10-11 main.R

@@ -6,6 +6,9 @@ noacsr::source_all_functions()
 ## Prepare data
 prepared.data <- prepare_data(data)
 
+##clean data
+New.subset <- New_subset(data)
+
 ## Create subset
 subset.data <- Create_subset(data)
 

### 2023-10-11 lab-book.Rmd
etiska tillstånd

# Lab book

The purpose of this file is to be a playground and place for
documentation at the same time. This is were you try different things,
and write about what you have tried and why, a place to write "dirty"
code so that the rest of your project can be kept clean. See this mostly
as a reference for your future self and use it in a way that makes sense
to you.

inga barn med (15 under) delay är OFI. vilka patienter? traumaregister
kopplat till swetrau, inte samma dock. swetrau, NISS 15, NKS ESS 9
b.l.a. kvalitetsregister är det konferenser sammanställer. auditfilters
faller där, kallar till konferens om OFI är positivt create OFI, skapar
binärt ja eller nej. typ. delta time. ph prehospital. ed emergency
department kolum är tiden till första hjälpande insats. kolum är i
minuter, kom till hjälp. olika faktorer som höjer tid. DT treatment
finns ej på 43% OFI, 74% av ej OFI. dt_ed_emerg_prod är min

mycket data finns på patientfaktorer o mortalitet, men ej fördröjning

konferenser ovalninga globalt, resurskrävande. bara KS i sverige. nästan
alla sjukhus går igenom dödsfall. Finns på KI 2010-2012. hjärtstopp ung
man, vet ej hur man skulle göra. intresekonflikt mellan medicin o
kirurg. gunilla wilke? anestesisköterska styrde upp mha två läkare
kvalitetsförbättringsarbete. 2013 började? gick först igenom dödsfall,
sen utanför mortalitet. försökte inkorporera audit filters, mellan
2013-2017 audit filter för brett. valde patienter på måfå. gjorde om
filter till de vi har nu, fick in en till sjuksyrra. liselott applicerar
alla filter, typ 9 olika. om ja, går igenom pat med gunilla. på
konferens alla som sysslar med trauma

plan B, vi har all data så det ska gå

tog bort csl: bmcemerg.csl

<https://openarchive.ki.se/xmlui/bitstream/handle/10616/44832/Thesis_Martin_Gerdin.pdf?sequence=3&isAllowed=y>

etiska tillstånd för detta projekt, skilt från swetrau. Etiska tillstånd
diarienummer

### 2023-10-11 example_function.R
set categories

@@ -10,9 +10,6 @@ cleaned_data$Gender <- ifelse(cleaned_data$Gender == "M", 1, 0)
 
 
 
-  model <- glm(OFI_delay ~ pt_age_yrs + ISS + inj_mechanism + pt_Gender, data = cleaned_data, family = binomial)
+  model <- glm(OFI_delay ~ ed_gcs_sum + Gender + res_survival + host_care_level + ed_rr_value + ed_sbp_value + dt_ed_first_ct + pre_intubated + ISS + pt_age_yrs, data = New.subset, family = binomial)
   
-  filtered.data <- subset(New.subset, "pt_age_yrs" + "ISS" + "inj_mechanism" + "pt_Gender", data != "999")
-  
-  prepared.data$OFI_delay <- as.factor(ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", "1", "0"))
-  
+

### 2023-10-11 functions/Create_subset.R

@@ -1,13 +1,0 @@
-library(rofi)
-## Creating function
-Create_subset <- function(dataset){
- 
-   ## creating subset
-  subset.datas <- prepared.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi", "Problemomrade_.FMP")]
-  
-  ## Exclude patients with missing variables
-  newsubset.data <- na.omit(subset.datas)
-  
-  return(newsubset.data)
-}
-

### 2023-10-11 functions/New_subset.R

@@ -2,9 +2,11 @@ library(rofi)
 
 New_subset <- function(dataset){
   
-  prepared.data$OFI_delay <- ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", "1", "0")
-
-  cleaned_data <- subset(prepared.data, !is.na(OFI_delay))
+  ## create new colum for outcome to be observed
+  prepared.data$OFI_delay <- as.factor(ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", "Delay to treatment", "No delay to treatment"))
+  
+  ## remove rows without OFI
+  cleaned_data <- subset(prepared.data, !is.na(ofi))
 
   return(cleaned_data)
 }

### 2023-10-11 functions/create_factors.R

@@ -1,0 +1,18 @@
+library(rofi)
+## Creating function
+Create_factors <- function(dataset){
+ 
+  ## Making my variables into factors
+  New.subset$ed_gcs_sum <- as.factor(New.subset$ed_gcs_sum)
+  New.subset$Gender <- as.factor(New.subset$Gender)
+  New.subset$res_survival <- as.factor(New.subset$res_survival)
+  New.subset$host_care_level <- as.factor(New.subset$host_care_level)
+  New.subset$ed_rr_value <- as.factor(New.subset$ed_rr_value)
+  New.subset$ed_sbp_value <- as.factor(New.subset$ed_sbp_value)
+  New.subset$dt_ed_first_ct <- as.factor(New.subset$dt_ed_first_ct)
+  New.subset$pre_intubated <- as.factor(New.subset$pre_intubated)
+
+
+  return(newsubset.data)
+}
+

### 2023-10-17 functions/New_subset.R
code

@@ -5,8 +5,11 @@ New_subset <- function(dataset){
   ## create new colum for outcome to be observed
   prepared.data$OFI_delay <- as.factor(ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", "Delay to treatment", "No delay to treatment"))
   
+  ## remove all columns except the analyzed ones
+  prepare.data <- prepared.data[, c("Gender", "res_survival", "host_care_level", "ed_gcs_sum", "ed_rr_value", "ed_sbp_value", "dt_ed_first_ct", "pre_intubated", "pt_age_yrs", "ISS", "ofi", "OFI_delay", "pre_gcs_sum")]
+  
   ## remove rows without OFI
-  cleaned_data <- subset(prepared.data, !is.na(ofi))
+  cleaned_data <- subset(prepare.data, !is.na(ofi))
 
   return(cleaned_data)
 }

### 2023-10-17 functions/create_factors.R

@@ -2,17 +2,47 @@ library(rofi)
 ## Creating function
 Create_factors <- function(dataset){
  
-  ## Making my variables into factors
+  ## Making gcs into rts and replacing missing values
+  New.subset$ed_gcs_sum <- ifelse(New.subset$ed_gcs_sum >= 3 & New.subset$ed_gcs_sum <= 3, 0,
+                     ifelse(New.subset$ed_gcs_sum >= 4 & New.subset$ed_gcs_sum <= 5, 1,
+                            ifelse(New.subset$ed_gcs_sum >= 6 & New.subset$ed_gcs_sum <= 8, 2,
+                                   ifelse(New.subset$ed_gcs_sum >= 9 & New.subset$ed_gcs_sum <= 12, 3,
+                                          ifelse(New.subset$ed_gcs_sum >= 13 & New.subset$ed_gcs_sum <= 15, 4,  New.subset$ed_gcs_sum)))))
+  
+  New.subset$pre_gcs_sum <- ifelse(New.subset$pre_gcs_sum >= 3 & New.subset$pre_gcs_sum <= 3, 0,
+                                  ifelse(New.subset$pre_gcs_sum >= 4 & New.subset$pre_gcs_sum <= 5, 1,
+                                         ifelse(New.subset$pre_gcs_sum >= 6 & New.subset$pre_gcs_sum <= 8, 2,
+                                                ifelse(New.subset$pre_gcs_sum >= 9 & New.subset$pre_gcs_sum <= 12, 3,
+                                                       ifelse(New.subset$pre_gcs_sum >= 13 & New.subset$pre_gcs_sum <= 15, 4,  New.subset$pre_gcs_sum)))))
+                                  
+  New.subset$ed_gcs_sum <- ifelse(is.na(New.subset$ed_gcs_sum) | New.subset$ed_gcs_sum == 999 | New.subset$ed_gcs_sum == 99, New.subset$pre_gcs_sum, New.subset$ed_gcs_sum)  
+  
+  New.subset <- subset(New.subset, !(is.na(ed_gcs_sum) | ed_gcs_sum == 999 | ed_gcs_sum == 99))
+  
   New.subset$ed_gcs_sum <- as.factor(New.subset$ed_gcs_sum)
+  
+  ## Making gender into a factor
+  New.subset <- subset(New.subset, !is.na(Gender))
+  
+  
   New.subset$Gender <- as.factor(New.subset$Gender)
+  
+  
+  
[…]

### 2023-10-17 functions/view_data.R

@@ -1,0 +1,12 @@
+library(rofi, gtsummary)
+
+view_data <- function(dataset){
+  
+  ## regression model
+  model <- glm(OFI_delay ~ ed_gcs_sum + Gender + res_survival + host_care_level + ed_rr_value + ed_sbp_value + dt_ed_first_ct + pre_intubated + ISS + pt_age_yrs, data = New.subset, family = binomial)
+  
+  ## view model
+  tbl_regression(model, exponentiate = TRUE)
+
+  return(model)
+}

### 2023-10-17 main.R

@@ -10,10 +10,13 @@ prepared.data <- prepare_data(data)
 New.subset <- New_subset(data)
 
 ## Create subset
-subset.data <- Create_subset(data)
+subset.data <- Create_factors(data)
+
+## view data
+view.data <- view_data(data)
 
 
-print(subset.data)
+print()
 
 ## Welcome!
 

### 2023-10-17 functions/New_subset.R
code

@@ -6,7 +6,7 @@ New_subset <- function(dataset){
   prepared.data$OFI_delay <- as.factor(ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", "Delay to treatment", "No delay to treatment"))
   
   ## remove all columns except the analyzed ones
-  prepare.data <- prepared.data[, c("Gender", "res_survival", "host_care_level", "ed_gcs_sum", "ed_rr_value", "ed_sbp_value", "dt_ed_first_ct", "pre_intubated", "pt_age_yrs", "ISS", "ofi", "OFI_delay", "pre_gcs_sum")]
+  prepare.data <- prepared.data[, c("Gender", "res_survival", "host_care_level", "ed_gcs_sum", "ed_rr_value", "ed_sbp_value", "dt_ed_first_ct", "pre_intubated", "pt_age_yrs", "ISS", "ofi", "OFI_delay", "pre_gcs_sum", "ed_rr_rtscat")]
   
   ## remove rows without OFI
   cleaned_data <- subset(prepare.data, !is.na(ofi))

### 2023-10-17 functions/create_factors.R

@@ -19,7 +19,7 @@ Create_factors <- function(dataset){
   
   New.subset <- subset(New.subset, !(is.na(ed_gcs_sum) | ed_gcs_sum == 999 | ed_gcs_sum == 99))
   
-  New.subset$ed_gcs_sum <- as.factor(New.subset$ed_gcs_sum)
+  New.subset$Total_GCS <- as.factor(New.subset$ed_gcs_sum)
   
   ## Making gender into a factor
   New.subset <- subset(New.subset, !is.na(Gender))
@@ -28,21 +28,58 @@ Create_factors <- function(dataset){
   New.subset$Gender <- as.factor(New.subset$Gender)
   
   
-  
-  New.subset$res_survival <- as.factor(New.subset$res_survival)
-  
-  New.subset$host_care_level <- as.factor(New.subset$host_care_level)
-  
-  New.subset$ed_rr_value <- as.factor(New.subset$ed_rr_value)
-  
-  New.subset$ed_sbp_value <- as.factor(New.subset$ed_sbp_value)
-  
-  New.subset$dt_ed_first_ct <- as.factor(New.subset$dt_ed_first_ct)
-  
-  New.subset$pre_intubated <- as.factor(New.subset$pre_intubated)
+  ## Survival after 30 days
+  New.subset <- subset(New.subset, !(is.na(res_survival) | res_survival == 999 | res_survival == 99))
+
+  New.subset$res_survival <- ifelse(New.subset$res_survival == 1, "Dead", "Alive")
+   
+  New.subset$survival_after_30_days <- as.factor(New.subset$res_survival)
   
 
+  ## Highest hospital care level
+  New.subset$host_care_level <- ifelse(New.subset$host_care_level == 1, "ER",
+                               ifelse(New.subset$host_care_level == 2, "General care department",
+                                      ifelse(New.subset$host_care_level == 3, "OR",
+                                             ifelse(New.subset$host_care_level == 4, "Specialized care department",
+                                                    ifelse(New.subset$host_care_level == 5, "ICU", NA)))))
+  
+  New.subset <- subset(New.subset, !(is.na(host_care_level) | host_care_level == 999 | host_care_level == 99))
+  
+  New.subset$Highest_care_level <- as.factor(New.subset$host_care_level)
+  
+  ## Respiratory rate into rts and replacing missing values
[…]

### 2023-10-17 functions/view_data.R

@@ -4,6 +4,8 @@ view_data <- function(dataset){
   
   ## regression model
   model <- glm(OFI_delay ~ ed_gcs_sum + Gender + res_survival + host_care_level + ed_rr_value + ed_sbp_value + dt_ed_first_ct + pre_intubated + ISS + pt_age_yrs, data = New.subset, family = binomial)
+  model <- glm(OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + dt_ed_first_ct + pre_intubated + ISS + pt_age_yrs, data = New.subset, family = binomial)
+  
   
   ## view model
   tbl_regression(model, exponentiate = TRUE)

### 2023-10-18 lab-book.Rmd
set categories
Heaps of code

@@
 diarienummer
+
+The categorical factors
+
+-   sex- Gender
+
+-   survival after 30 days- res_survival
+
+-   highest hospital care level- host_care_level
+
+-   GCS- ed_gcs_sum
+
+-   respiratory rate- ed_rr_value
+
+-   systolic blood pressure- ed_sbp_value
+
+-   working hours
+
+-   weekend
+
+-   time from arrival at the hospital until first computed tomography
+    (CT)- dt_ed_first_ct
+
+-   if the patient was intubated- ed_intubated eller pre_intubated
+
+We also included the continuous factors
+
+-   age- pt_age_yrs
+
+-   ISS- ISS
+
+New.subset\$dt_ed_first_ct \<- as.factor(New.subset\$dt_ed_first_ct)

### 2023-10-18 functions/create_factors.R

@@ -7,17 +7,19 @@ Create_factors <- function(dataset){
                      ifelse(New.subset$ed_gcs_sum >= 4 & New.subset$ed_gcs_sum <= 5, 1,
                             ifelse(New.subset$ed_gcs_sum >= 6 & New.subset$ed_gcs_sum <= 8, 2,
                                    ifelse(New.subset$ed_gcs_sum >= 9 & New.subset$ed_gcs_sum <= 12, 3,
-                                          ifelse(New.subset$ed_gcs_sum >= 13 & New.subset$ed_gcs_sum <= 15, 4,  New.subset$ed_gcs_sum)))))
+                                          ifelse(New.subset$ed_gcs_sum >= 13 & New.subset$ed_gcs_sum <= 15, 4,
+                                                 ifelse(New.subset$ed_gcs_sum == 99, "intubated", New.subset$ed_gcs_sum))))))
   
   New.subset$pre_gcs_sum <- ifelse(New.subset$pre_gcs_sum >= 3 & New.subset$pre_gcs_sum <= 3, 0,
                                   ifelse(New.subset$pre_gcs_sum >= 4 & New.subset$pre_gcs_sum <= 5, 1,
                                          ifelse(New.subset$pre_gcs_sum >= 6 & New.subset$pre_gcs_sum <= 8, 2,
                                                 ifelse(New.subset$pre_gcs_sum >= 9 & New.subset$pre_gcs_sum <= 12, 3,
-                                                       ifelse(New.subset$pre_gcs_sum >= 13 & New.subset$pre_gcs_sum <= 15, 4,  New.subset$pre_gcs_sum)))))
+                                                       ifelse(New.subset$pre_gcs_sum >= 13 & New.subset$pre_gcs_sum <= 15, 4,
+                                                              ifelse(New.subset$ed_gcs_sum == 99, "intubated", New.subset$pre_gcs_sum))))))
                                   
-  New.subset$ed_gcs_sum <- ifelse(is.na(New.subset$ed_gcs_sum) | New.subset$ed_gcs_sum == 999 | New.subset$ed_gcs_sum == 99, New.subset$pre_gcs_sum, New.subset$ed_gcs_sum)  
+  New.subset$ed_gcs_sum <- ifelse(is.na(New.subset$ed_gcs_sum) | New.subset$ed_gcs_sum == 999, New.subset$pre_gcs_sum, New.subset$ed_gcs_sum)  
   
[…]

### 2023-10-18 functions/view_data.R

@@ -1,10 +1,9 @@
-library(rofi, gtsummary)
+library(gtsummary)
 
 view_data <- function(dataset){
   
   ## regression model
-  model <- glm(OFI_delay ~ ed_gcs_sum + Gender + res_survival + host_care_level + ed_rr_value + ed_sbp_value + dt_ed_first_ct + pre_intubated + ISS + pt_age_yrs, data = New.subset, family = binomial)
-  model <- glm(OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + dt_ed_first_ct + pre_intubated + ISS + pt_age_yrs, data = New.subset, family = binomial)
+  model <- glm(OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + dt_ed_first_ct + Intubated_prehospitaly + ISS + Age, data = New.subset, family = binomial)
   
   
   ## view model

### 2023-10-19 functions/New_subset.R
code

@@ -6,7 +6,7 @@ New_subset <- function(dataset){
   prepared.data$OFI_delay <- as.factor(ifelse(prepared.data$Problemomrade_.FMP == "Handläggning", "Delay to treatment", "No delay to treatment"))
   
   ## remove all columns except the analyzed ones
-  prepare.data <- prepared.data[, c("Gender", "res_survival", "host_care_level", "ed_gcs_sum", "ed_rr_value", "ed_sbp_value", "dt_ed_first_ct", "pre_intubated", "pt_age_yrs", "ISS", "ofi", "OFI_delay", "pre_gcs_sum", "ed_rr_rtscat")]
+  prepare.data <- prepared.data[, c("Gender", "res_survival", "host_care_level", "ed_gcs_sum", "ed_rr_value", "ed_sbp_value", "dt_ed_first_ct", "pre_intubated", "pt_age_yrs", "ISS", "ofi", "OFI_delay", "pre_gcs_sum", "ed_rr_rtscat", "DateTime_ArrivalAtHospital")]
   
   ## remove rows without OFI
   cleaned_data <- subset(prepare.data, !is.na(ofi))

### 2023-10-19 functions/create_factors.R

@@ -69,6 +69,8 @@ Create_factors <- function(dataset){
                                                  ifelse(New.subset$ed_sbp_value >= 76 & New.subset$ed_sbp_value <= 89, 3,
                                                         ifelse(New.subset$ed_sbp_value >= 89 & New.subset$ed_sbp_value <= 500, 4,  New.subset$ed_sbp_value)))))
   
+  
+  
   New.subset$ed_sbp_value <- ifelse(is.na(New.subset$ed_sbp_value) | New.subset$ed_sbp_value == 999, New.subset$ed_sbp_rtscat, New.subset$ed_sbp_value) 
   
   New.subset <- subset(New.subset, !(is.na(ed_sbp_value) | ed_sbp_value == 999))
@@ -82,6 +84,18 @@ Create_factors <- function(dataset){
   
   New.subset$Intubated_prehospitaly <- as.factor(New.subset$pre_intubated)
   
+  ##CT scan categorizing
+  New.subset$dt_ed_first_ct <- ifelse(New.subset$dt_ed_first_ct >= 1 & New.subset$dt_ed_first_ct <= 30, "1-30 mins",
+                                  ifelse(New.subset$dt_ed_first_ct >= 31 & New.subset$dt_ed_first_ct <= 60, "31-60 mins",
+                                         ifelse(New.subset$dt_ed_first_ct >= 61 & New.subset$dt_ed_first_ct <= 120, "61-120 mins",
+                                                ifelse(New.subset$dt_ed_first_ct >= 121 & New.subset$dt_ed_first_ct <= 9999, "120+ mins", New.subset$dt_ed_first_ct))))
+  
+  New.subset <- subset(New.subset, !is.na(dt_ed_first_ct))
+  
+  
+  New.subset$Time_until_first_CT <- as.factor(New.subset$dt_ed_first_ct)
+  
+  
   ##Fixing years
   New.subset <- subset(New.subset, !(is.na(pt_age_yrs) | pt_age_yrs == 999))
   
@@ -91,7 +105,20 @@ Create_factors <- function(dataset){
   New.subset <- subset(New.subset, !(is.na(ISS) | ISS == 999))
   
   New.subset$ISS <- as.numeric(New.subset$ISS)
+  
+  ##Dates
+  library(stringr)
[…]

### 2023-10-19 functions/view_data.R

@@ -3,7 +3,7 @@ library(gtsummary)
 view_data <- function(dataset){
   
   ## regression model
-  model <- glm(OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + dt_ed_first_ct + Intubated_prehospitaly + ISS + Age, data = New.subset, family = binomial)
+  model <- glm(OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Time_until_first_CT + Intubated_prehospitaly + ISS + Age + weekday + work_hours, data = New.subset, family = binomial)
   
   
   ## view model

### 2023-10-24 functions/create_factors.R
code

@@ -137,6 +137,6 @@ Create_factors <- function(dataset){
   New.subset$Date <- NULL
   New.subset$Time <- NULL
   
-  return(factors.data)
+  return(New.subset)
 }
 

### 2023-10-24 functions/create_factors.R
fix input parameter names not corresponding to usage and other small things

@@ -1,142 +1,145 @@
-library(rofi)
+library(stringr)
+
 ## Creating function
-Create_factors <- function(dataset){
+create.factors <- function(dataset){
  
   ## Making gcs into rts and replacing missing values
-  New.subset$ed_gcs_sum <- ifelse(New.subset$ed_gcs_sum >= 3 & New.subset$ed_gcs_sum <= 3, 0,
-                     ifelse(New.subset$ed_gcs_sum >= 4 & New.subset$ed_gcs_sum <= 5, 1,
-                            ifelse(New.subset$ed_gcs_sum >= 6 & New.subset$ed_gcs_sum <= 8, 2,
-                                   ifelse(New.subset$ed_gcs_sum >= 9 & New.subset$ed_gcs_sum <= 12, 3,
-                                          ifelse(New.subset$ed_gcs_sum >= 13 & New.subset$ed_gcs_sum <= 15, 4,
-                                                 ifelse(New.subset$ed_gcs_sum == 99, "intubated", New.subset$ed_gcs_sum))))))
+  dataset$ed_gcs_sum <- ifelse(dataset$ed_gcs_sum >= 3 & dataset$ed_gcs_sum <= 3, 0,
+                     ifelse(dataset$ed_gcs_sum >= 4 & dataset$ed_gcs_sum <= 5, 1,
+                            ifelse(dataset$ed_gcs_sum >= 6 & dataset$ed_gcs_sum <= 8, 2,
+                                   ifelse(dataset$ed_gcs_sum >= 9 & dataset$ed_gcs_sum <= 12, 3,
+                                          ifelse(dataset$ed_gcs_sum >= 13 & dataset$ed_gcs_sum <= 15, 4,
+                                                 ifelse(dataset$ed_gcs_sum == 99, "intubated", dataset$ed_gcs_sum))))))
   
-  New.subset$pre_gcs_sum <- ifelse(New.subset$pre_gcs_sum >= 3 & New.subset$pre_gcs_sum <= 3, 0,
-                                  ifelse(New.subset$pre_gcs_sum >= 4 & New.subset$pre_gcs_sum <= 5, 1,
-                                         ifelse(New.subset$pre_gcs_sum >= 6 & New.subset$pre_gcs_sum <= 8, 2,
-                                                ifelse(New.subset$pre_gcs_sum >= 9 & New.subset$pre_gcs_sum <= 12, 3,
[…]

### 2023-10-24 functions/new_subset.R

@@ -1,0 +1,32 @@
+library(rofi)
+
+clean.dataset <- function(dataset){
+  ## create new colum for outcome to be observed
+  dataset$OFI_delay <- as.factor(ifelse(dataset$Problemomrade_.FMP == "Handläggning", "Delay to treatment", "No delay to treatment"))
+  
+  ## remove all columns except the analyzed ones
+  dataset <-
+    dataset[, c(
+      "Gender",
+      "res_survival",
+      "host_care_level",
+      "ed_gcs_sum",
+      "ed_rr_value",
+      "ed_sbp_value",
+      "dt_ed_first_ct",
+      "pre_intubated",
+      "pt_age_yrs",
+      "ISS",
+      "ofi",
+      "OFI_delay",
+      "pre_gcs_sum",
+      "ed_rr_rtscat",
+      "ed_sbp_rtscat",
+      "DateTime_ArrivalAtHospital"
+    )]
+  
+  ## remove rows without OFI
+  cleaned.data <- subset(dataset, !is.na(ofi))
+
+  return(cleaned.data)
+}

### 2023-10-24 functions/view_data.R

@@ -1,9 +1,14 @@
 library(gtsummary)
 
-view_data <- function(dataset){
+model.data <- function(dataset){
   
   ## regression model
-  model <- glm(OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Time_until_first_CT + Intubated_prehospitaly + ISS + Age + weekday + work_hours, data = New.subset, family = binomial)
+  model <-
+    glm(
+      OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Time_until_first_CT + Intubated_prehospitaly + ISS + Age + weekday + work_hours,
+      data = dataset,
+      family = binomial
+    )
   
   
   ## view model

### 2023-10-25 manuscriptlongversion.Rmd

@@
 bibliography: bibliography.bib
+csl: vancouver.csl
 editor_options:
@@
 
-**Definition**
+## Definition
 
@@
 trauma centers is associated with a 15-30% decrease in mortality
-[@Moran_2018][@Celso_2006][@MacKenzie_2006][@Haas_2012]. During the last
-three decades, the introduction of trauma systems has contributed to
-lowering the incidence of preventable death . This is attributed to the
-improvements in care for acute brain injuries and bleeding control. The
-incidence of late death because of sepsis and multiple organ failure,
-possibly a result of better and earlier resuscitation [@Asensio_2008].
-Functioning trauma systems saves lives in the hospital, but their work
-outside of the hospital is just as important. Data collection from
-injured patients, such as mechanism of injury or mortality, are
-essential for creating databases that can be used for research. In turn,
-that research can be used for planning injury prevention programs that
-target the most common injuries in the right ways, e.g. teen drivers,
-children, specific occupations etc. These injury prevention programs can
-be planned on trauma center, organizational or government level[@ACS].
+[@Moran_2018][@Celso_2006][@MacKenzie_2006][@Haas_2012]
+<!-- KS: multiple citations should be like this: [@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. Fix this where applicable -->.
+During the last three decades, the introduction of trauma systems has
+contributed to lowering the incidence of preventable death . This is
+attributed to the improvements in care for acute brain injuries and
+bleeding control. The incidence of late death because of sepsis and
+multiple organ failure, possibly a result of better and earlier
+resuscitation [@Asensio_2008]. Functioning trauma systems saves lives in
+the hospital, but their work outside of the hospital is just as
+important. Data collection from injured patients, such as mechanism of
+injury or mortality, are essential for creating databases that can be
+used for research. In turn, that research can be used for planning
+injury prevention programs that target the most common injuries in the
+right ways, e.g. teen drivers, children, specific occupations etc. These
+injury prevention programs can be planned on trauma center,
+organizational or government level[@ACS].
 
@@
 
-**Setting**
+## Setting
 
@@
 
-**Study design**
+## Study design
 
@@
 
-**Outcome**
+## Outcome
 
@@
 
-**Participants**
+## Participants
 
@@
 
-**Variables**
+## Variables
 
@@
 
-**Data sources/measurement**
+## Data sources/measurement
 
@@
 
-**Bias**
+## Bias
 
@@
 
-**Study size**
+## Study size
 
@@
 
-**Statistical methods**
+## Statistical methods
 

### 2023-10-25 manuscriptlongversion.Rmd
källor

@@
 trauma centers is associated with a 15-30% decrease in mortality
-[@Moran_2018][@Celso_2006][@MacKenzie_2006][@Haas_2012]
-<!-- KS: multiple citations should be like this: [@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. Fix this where applicable -->.
-During the last three decades, the introduction of trauma systems has
-contributed to lowering the incidence of preventable death . This is
-attributed to the improvements in care for acute brain injuries and
-bleeding control. The incidence of late death because of sepsis and
-multiple organ failure, possibly a result of better and earlier
-resuscitation [@Asensio_2008]. Functioning trauma systems saves lives in
-the hospital, but their work outside of the hospital is just as
-important. Data collection from injured patients, such as mechanism of
-injury or mortality, are essential for creating databases that can be
-used for research. In turn, that research can be used for planning
-injury prevention programs that target the most common injuries in the
-right ways, e.g. teen drivers, children, specific occupations etc. These
+[@Moran_2018; \@Celso_2006; @MacKenzie_2006; \@Haas_2012] . During the
+last three decades, the introduction of trauma systems has contributed
+to lowering the incidence of preventable death . This is attributed to
+the improvements in care for acute brain injuries and bleeding control.
+The incidence of late death because of sepsis and multiple organ
+failure, possibly a result of better and earlier resuscitation
+[@Asensio_2008]. Functioning trauma systems saves lives in the hospital,
+but their work outside of the hospital is just as important. Data
+collection from injured patients, such as mechanism of injury or
+mortality, are essential for creating databases that can be used for
+research. In turn, that research can be used for planning injury
+prevention programs that target the most common injuries in the right
+ways, e.g. teen drivers, children, specific occupations etc. These
 injury prevention programs can be planned on trauma center,
@@
 why it is such an important issue and the need to develop strategies to
-combat [@Sampalis_1995][@Sampalis_1994]. Although it is such a common
+combat [@Sampalis_1995; \@Sampalis_1994]. Although it is such a common
 OFI, the patient level factors associated with delayed treatment are

### 2023-10-25 functions/create_factors.R
code

@@ -134,6 +134,7 @@ create.factors <- function(dataset){
         "pt_age_yrs",
         "pre_gcs_sum",
         "ed_rr_rtscat",
+        "ed_sbp_rtscat",
         "DateTime_ArrivalAtHospital",
         "Date",
         "Time"

### 2023-10-25 functions/view_data.R

@@ -6,7 +6,7 @@ model.data <- function(dataset){
   model <-
     glm(
       OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Time_until_first_CT + Intubated_prehospitaly + ISS + Age + weekday + work_hours,
-      data = dataset,
+      data = factors.data,
       family = binomial
     )
   
@@ -14,5 +14,19 @@ model.data <- function(dataset){
   ## view model
   tbl_regression(model, exponentiate = TRUE)
 
+  factors.data %>%
+    tbl_summary(  # Create summary statistics for different groups
+      by = OFI_delay,
+      label = list(
+        Age = "Age (years)",
+        Gender = "Gender",
+        Total_GCS = "GCS"
+      ),
+      statistic = list(all_continuous() ~ "{mean} ± {sd}"),
+      digits = list(all_continuous() ~ c(2, 2))
+    ) %>%
+    add_p()  # Add p-values
+  
+  
   return(model)
 }

### 2023-10-26 manuscriptlongversion.tex
code

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Trauma treatment delay},
  pdfauthor={Anton Wasielewski},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Trauma treatment delay}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{A retrospective cohort study on the factors affecting the
delays in trauma care}
\author{Anton Wasielewski}
\date{}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{definition}{%
\subsection{Definition}\label{definition}}

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response (1), is and has long been a major cause
of death around the world today (2). Within modern hospital care, the
trauma system stands as an important part in healthcare and is crucial
to lowering mortality and morbidity in injured patients through many
means, such as hospital care, follow up and prevention programs (3).

\hypertarget{trauma-statistics}{%
\subsection{Trauma Statistics}\label{trauma-statistics}}

Trauma is often divided into two subgroups, blunt force trauma and
penetrating trauma. Blunt force trauma is when an object or force
strikes the body, often causing bruising, broken bones or deep cuts.
Examples of blunt trauma could be car crashes, falls or direct blows to
the body. Penetrating trauma is when an object pierces the skin or body
and creates an open wound. Examples of penetrating trauma are gunshot
wounds and stab wounds (4). Differences exist between the two groups, as
blunt trauma patients tend to be more injured on arrival to the hospital
and as such, require more resources and are hospitalized for a longer
periods (5).

Today, trauma takes the lives of around 4,4 million people each year,
almost 8\% of all deaths (2). In the United States, trauma is the 4th
leading cause of death among the general population and the leading
cause for people between the ages of 1 and 44 (6). It is also an
important cause of hospitalization and morbidity among all age groups,
including seniors, and is responsible for an estimated 10\% of all years
lived with a disability globally. This has a significant burden on
social and economic level, costing countries billions of US dollars each
year in healthcare and lost productivity (7). Studies estimate the cost
of trauma care being between 18,500\$ and 41,500\$ per patient in high
income countries (HIC), depending on the country (8). However, it has
been shown that this burden is not evenly distributed between or within
countries. Many social factors such as age, sex and social status play a
major role in the risk of dying from trauma, with young men with low
socio-economic status being at most risk (2). But it is not only patient
level factors that affect what effect trauma has on people. About 90\%
of all trauma related deaths occur in low- and middle-income countries
(LMIC), with death rates by trauma also being higher than in high income
countries. Even within these LMIC, people of poorer socio-economic
status have higher death rates from trauma. Problems identified within
these countries that contribute to these statisitics were
infrastructure, education and training, attributed to lack of funding,
brain drain to HIC and lack of availability of basic amenities (9). In
HIC where funding and governance over the healthcare system is better
designed, functioning trauma systems and dedicated trauma centers exist
that have been shown to lower mortality but also improve functional
outcome in trauma patients (10).

\hypertarget{trauma-systems}{%
\subsection{Trauma systems}\label{trauma-systems}}

Trauma systems are infrastructures that exist to provide and optimize
care for injured patients starting with injury recognition and triage,
transport to appropriate trauma center, inpatient care and outpatient
follow-up. Beyond the clinical side, trauma systems work with outreach,
education and advocacy, data collection through registries, research,
funding, and disaster preparedness and response. A comprehensive and
functioning trauma system requires strong leadership and engagement at
the trauma center, regional and national level (7). This system is
crucial to provide care for trauma patients, both in reducing morbidity
and mortality in this patient group. Earlier studies from Sweden have
shown that treating severely injured patients at a trauma center is
associated with a 41\% lower adjusted 30-day mortality rate compared to
being treated at a non trauma care center due to them being more capable
of treating these patients, with potential survival benefit increasing
with higher injury severity (11). Other studies have shown similar
results, showing that treating injured patients at trauma centers is
associated with a 15-30\% decrease in mortality {[}(12); @Celso\_2006;
(13); @Haas\_2012{]} . During the last three decades, the introduction
of trauma systems has contributed to lowering the incidence of
preventable death . This is attributed to the improvements in care for
acute brain injuries and bleeding control. The incidence of late death
because of sepsis and multiple organ failure, possibly a result of
better and earlier resuscitation (14). Functioning trauma systems saves
lives in the hospital, but their work outside of the hospital is just as
important. Data collection from injured patients, such as mechanism of
injury or mortality, are essential for creating databases that can be
used for research. In turn, that research can be used for planning
injury prevention programs that target the most common injuries in the
right ways, e.g.~teen drivers, children, specific occupations etc. These
injury prevention programs can be planned on trauma center,
organizational or government level(15).

Trauma teams are multidisciplinary and operate in these centers. They
play a pivotal role in the treatment of the trauma patient, as they
provide the initial care in the critical stage of trauma. In Sweden,
trauma teams are lead by a team leader who is a surgeon, and include
practitioners from the specialities of intensive care, orthopedics,
nursing and support staff(16). For the trauma team to be mobilized,
trauma code has to be activated, often by an emergency nurse. The nurse
uses information gathered by the first responders to see if the patient
fits any of the criteria for trauma code. Most healthcare facilities
have established criteria or guidelines that trigger trauma code
activation. These criteria typically include mechanisms of injury,
physiological criteria, anatomical criteria and other specific
indicators of severe trauma. There are different levels of trauma code
with different criteria correlating with severity of injury, with level
1 mobilizing the most personnel to the trauma room. Once in the trauma
room, the team works systematically to manage the patients injuries.
They handle the most urgent problems first, such as airways and
breathing, with the aim to rapidly assess and stabilise the patient,
prioritise their injuries and arrange for site of definitive care. (17)

\hypertarget{mortality-and-morbidity-mom-conferences}{%
\subsection{Mortality and morbidity (MoM)
conferences}\label{mortality-and-morbidity-mom-conferences}}

Many hospitals have a trauma registry where they log the patients
information and timeline of what happened. The extent of the registry
varies, with HIC having more complete registries. This logging of
information is important work, as a cornerstone of trauma quality
improvement programs is multidisciplinary MoM conferences. The MoM
conference is a meeting where different specialities who work with
trauma care sit down and discuss deaths and complications in order to
look for preventable factors. These conferences are performed in many
hospitals globally, almost everywhere where there are formal medical
specialty departments and sometimes in smaller hospitals as well (18).
In Sweden, many hospitals have mortality conferences on deceased patient
cases, but only one has MoM conferences to better their trauma care on
patients who lived (19). The reason being that this expansion of MoM
conferences takes time and resources, which is why all hospitals don't
implement this.

The endpoint of these MoM conferences are opportunities for improvement
(OFI). At the end of the conference, consensus is reached regarding the
existence of an OFI and implementation of corrective action. This
process is effective, as it has been shown that this review is
associated with high-quality trauma care (20). Examples of OFI may
include lack of resources and management errors. One common OFI is
delayed treatment. According to previous studies, among the preventable
or possibly preventable deaths in trauma patients, delay in treatment
has been identified as a major error contributing to death, found in up
to 52.9\% of patients in said group (21). Delayed treatment has been
shown to have adverse effects on patients, showing why it is such an
important issue and the need to develop strategies to combat {[}(22);
@Sampalis\_1994{]}. Although it is such a common OFI, the patient level
factors associated with delayed treatment are poorly understood. There
may be several factors that correlate with receiving delayed treatment,
but they have yet to be identified. Mapping these factors may help in
identifying patients that might be at risk for receiving delayed
treatment before it happens.

\hypertarget{aim}{%
\subsection{Aim}\label{aim}}

The aim of this paper is to determine what different patient level
factors affect the risk of receiving delayed treatment at a trauma
center.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{setting}{%
\subsection{Setting}\label{setting}}

Every year, Karolinska University hospital, which is the equivalent of a
trauma level 1 hospital, admits 1500 trauma patients each year. To be
added to the trauma registry, a patient must be over 15 years of age,
had ISS\textgreater9 and/or had trauma code activation. They get added
to both the Karolinska Trauma registry as well as the national trauma
registry (SweTrau). The registry includes data on vital signs, injuries,
interventions and patient information. Another database, the trauma care
quality database includes information relevant for MoM conferences. At
Karolinska University Hospital, patients who die during hospital stay
gets their case sent to a MoM everytime. However, for the rest of the
patients, two specialized nurses determine whether they are. Several
audit filters are set up to flag for patients with possible OFIs.
Flagged patients are then reviewed by these two nurses who decide if
there is a possible OFI, in which case their case is sent to a MoM
conference for further review.

\hypertarget{study-design}{%
\subsection{Study design}\label{study-design}}

We conducted a registry based cohort study using data from the trauma
registry and trauma care quality database at the Karolinska University
Hospital in Solna. The trauma registry includes about 14000 patients
treated between 2012 and 2022. The trauma care quality database is a
subset of the trauma registry and includes about 8000 patients selected
for review between 2013 and 2022. This project will link the two
databases and assess how different patient level factors, such as age,
sex, mechanism of injury, and injury severity, are associated with
delayed treatment using logistic regression.

\hypertarget{outcome}{%
\subsection{Outcome}\label{outcome}}

The outcome is delayed treatment, as identified by the multidisciplinary
review board and recorded in the trauma care quality database.

\hypertarget{participants}{%
\subsection{Participants}\label{participants}}

The database only includes patients 15 years old and above, ISS
\textgreater9 and/or trauma team activation. Inclusion in the study
further requires that the patients have been assessed for OFIs. Patients
were excluded if they were missing information in one of the examined
patient factors.

\hypertarget{variables}{%
\subsection{Variables}\label{variables}}

The outcome is delayed treatment, which is defined as more than 30
minutes passing from arrival before receiving treatment, as identified
by the multidisciplinary review board and recorded in the trauma care
quality database.

\hypertarget{data-sourcesmeasurement}{%
\subsection{Data sources/measurement}\label{data-sourcesmeasurement}}

The data comes from the trauma registry and trauma care quality
database.

\hypertarget{bias}{%
\subsection{Bias}\label{bias}}

Selection bias as this is a retrospective cohort study where the outcome
has occurred.

Misclassification bias as the conference may reach a consensus that is
wrong.

\hypertarget{study-size}{%
\subsection{Study size}\label{study-size}}

The study size is all of the eligible patients that was treated at the
hospital between 2012 and 2022. Starting 2013 only a subgroup of
patients were screened for OFIs, but starting 2017 all patients that are
included in the database are screened for OFIs.

\hypertarget{statistical-methods}{%
\subsection{Statistical methods}\label{statistical-methods}}

Multivariable and potentially univariable logistical regression. A 5\%
significance level and 95\% confidence levels will be used.

\includegraphics{table1.html}

\hypertarget{ethical-considerations}{%
\subsection*{Ethical considerations}\label{ethical-considerations}}
\addcontentsline{toc}{subsection}{Ethical considerations}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-gerdin_risk_2015}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Gerdin M. The risk of dying: Predicting trauma mortality
in urban indian hospitals. Stockholm: Karolinska Institutet; 2015. }

\leavevmode\vadjust pre{\hypertarget{ref-WHO_2021}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Injuries and violence {[}Internet{]}. World Health
Organization. World Health Organization; 2021. Available from:
\url{https://www.who.int/news-room/fact-sheets/detail/injuries-and-violence}}

\leavevmode\vadjust pre{\hypertarget{ref-Choi_2021}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Choi J, Carlos G, Nassar AK, Knowlton LM, Spain DA. The
impact of trauma systems on patient outcomes {[}Internet{]}. Current
problems in surgery. U.S. National Library of Medicine; 2021. Available
from: \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7286246/}}

\leavevmode\vadjust pre{\hypertarget{ref-NIGMS_2020}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{General Medical Sciences NI of. Physical trauma
{[}Internet{]}. National Institute of General Medical Sciences. U.S.
Department of Health; Human Services; 2020. Available from:
\url{https://nigms.nih.gov/education/fact-sheets/Pages/physical-trauma.aspx}}

\leavevmode\vadjust pre{\hypertarget{ref-FITCH2019}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Blunt versus penetrating trauma: Is there a resource
intensity discrepancy? The American Journal of Surgery {[}Internet{]}.
2019;218(6):1201--5. Available from:
\url{https://www.sciencedirect.com/science/article/pii/S0002961019303861}}

\leavevmode\vadjust pre{\hypertarget{ref-AAST_2020}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{AAST. Trauma\_facts\_and\_links {[}Internet{]}. The
American Association for the Surgery of Trauma. The American Association
for the Surgery of Trauma; 2020. Available from:
\url{https://www.aast.org/resources/trauma-facts}}

\leavevmode\vadjust pre{\hypertarget{ref-connolly_2018}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Connolly R, Woo MY, Lampron J, Perry JJ.
\href{https://doi.org/10.1017/cem.2017.389}{Factors associated with
delay in trauma team activation and impact on patient outcomes}.
Canadian Journal of Emergency Medicine. 2018;20(4):606--13. }

\leavevmode\vadjust pre{\hypertarget{ref-Willenberg_2012}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Willenberg L, Curtis K, Taylor C, Jan S, Glass P,
Myburgh J. \href{https://doi.org/10.1186/1472-6963-12-267}{The variation
of acute treatment costs of trauma in high-income countries}. BMC Health
Services Research. 2012;12(1). }

\leavevmode\vadjust pre{\hypertarget{ref-Shanthakumar_2021}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Shanthakumar D, Payne A, Leitch T, Alfa-Wali M.
\href{https://doi.org/10.1055/s-0041-1732351}{Trauma care in low- and
middle-income countries}. The Surgery Journal. 2021;07(04). }

\leavevmode\vadjust pre{\hypertarget{ref-Nirula_2006}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Nirula R, Brasel K.
\href{https://doi.org/10.1097/01.ta.0000230305.36456.4e}{Do trauma
centers improve functional outcomes: A national trauma databank
analysis?} The Journal of Trauma: Injury, Infection, and Critical Care.
2006 Aug;61(2):268--71. }

\leavevmode\vadjust pre{\hypertarget{ref-Candefjord_2020}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Candefjord S, Asker L, Caragounis E-C.
\href{https://doi.org/10.1007/s00068-020-01446-6}{Mortality of trauma
patients treated at trauma centers compared to non-trauma centers in
sweden: A retrospective study}. European Journal of Trauma and Emergency
Surgery. 2020;48(1):525--36. }

\leavevmode\vadjust pre{\hypertarget{ref-Moran_2018}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Moran CG, Lecky F, Bouamra O, Lawrence T, Edwards A,
Woodford M, et al.
\href{https://doi.org/10.1016/j.eclinm.2018.07.001}{Changing the system
- major trauma patients and their outcomes in the NHS (england)
2008--17}. EClinicalMedicine. 2018;2--3:13--21. }

\leavevmode\vadjust pre{\hypertarget{ref-MacKenzie_2006}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{MacKenzie EJ, Rivara FP, Jurkovich GJ, Nathens AB, Frey
KP, Egleston BL, et al. \href{https://doi.org/10.1056/nejmsa052049}{A
national evaluation of the effect of trauma-center care on mortality}.
New England Journal of Medicine. 2006;354(4):366--78. }

\leavevmode\vadjust pre{\hypertarget{ref-Asensio_2008}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Asensio J, Trunkey D. Current therapy of trauma and
surgical critical care. Elsevier; 2008. }

\leavevmode\vadjust pre{\hypertarget{ref-ACS}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{ACS. Trauma systems consultation program {[}Internet{]}.
ACS. Available from:
\url{https://www.facs.org/quality-programs/trauma/systems/trauma-systems-consultation-program/}}

\leavevmode\vadjust pre{\hypertarget{ref-vuxe4stra_G}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Johnsson E. Traumateamet -- roller och ansvar
{[}Internet{]}. 2020. Available from:
\url{https://mellanarkiv-offentlig.vgregion.se/alfresco/s/archive/stream/public/v1/source/available/sofia/hs9766-305841775-25/surrogate/Traumateamet\%20-\%20roller\%20och\%20ansvar.pdf}}

\leavevmode\vadjust pre{\hypertarget{ref-Hedberg_2020}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Hedberg M, Falkén M, Wihlke G, Jansson S, Hultgren J,
Eriksson S, et al. Traumamanual {[}Internet{]}.
Traumamanual\_Karolinska\_Universitetssjukhuset\_Solna. NKS; 2020.
Available from:
\url{https://traumarummet.files.wordpress.com/2020/09/traumamanualen-2020.pdf}}

\leavevmode\vadjust pre{\hypertarget{ref-WHO_2009}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Guidelines for trauma quality improvement programmes
{[}Internet{]}. World Health Organization. World Health Organization;
2009. Available from:
\url{https://www.who.int/publications/i/item/guidelines-for-trauma-quality-improvement-programmes}}

\leavevmode\vadjust pre{\hypertarget{ref-SweTrau_2023}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{SweTrau. Täckningsgrad {[}Internet{]}. Täckningsgrad
\textbar{} SweTrau. 2023. Available from:
\url{https://rcsyd.se/swetrau/om-swetrau/tackningsgrad}}

\leavevmode\vadjust pre{\hypertarget{ref-Santana_2014}{}}%
\CSLLeftMargin{20. }%
\CSLRightInline{Santana MJ, Stelfox HT.
\href{https://doi.org/10.1097/sla.0b013e31828df98e}{Development and
evaluation of evidence-informed quality indicators for adult injury
care}. Annals of Surgery. 2014;259(1):186--92. }

\leavevmode\vadjust pre{\hypertarget{ref-Teixeira_2007}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{Teixeira PG, Inaba K, Hadjizacharia P, Brown C, Salim A,
Rhee P, et al.
\href{https://doi.org/10.1097/ta.0b013e31815078ae}{Preventable or
potentially preventable mortality at a mature trauma center}. Journal of
Trauma: Injury, Infection \&amp;amp; Critical Care. 2007
Dec;63(6):1338--47. }

\leavevmode\vadjust pre{\hypertarget{ref-Sampalis_1995}{}}%
\CSLLeftMargin{22. }%
\CSLRightInline{Sampalis JS, Boukas S, Lavoie A, Nikolis A, Frechette P,
Brown R, et al.
\href{https://doi.org/10.1097/00005373-199512000-00002}{Preventable
death evaluation of the appropriateness of the on-site trauma care
provided by urgences-sante physicians}. The Journal of Trauma: Injury,
Infection, and Critical Care. 1995;39(6):1029--35. }

\end{CSLReferences}

\end{document}

### 2023-10-26 functions/table1.R

@@ -1,0 +1,18 @@
+table1 <- function(dataset){
+  
+  factors.data %>%
+    tbl_summary(  # Create summary statistics for different groups
+      by = OFI_delay,
+      label = list(
+        Age = "Age (years)",
+        Gender = "Gender",
+        Total_GCS = "GCS"
+      ),
+      statistic = list(all_continuous() ~ "{mean} ± {sd}"),
+      digits = list(all_continuous() ~ c(2, 2))
+    ) %>%
+    add_p()  # Add p-values
+  
+  
+  return(table1)
+}

### 2023-10-27 manuscriptlongversion.Rmd
code
text

@@
 by the multidisciplinary review board and recorded in the trauma care
-quality database.
+quality database. The independent variables are gender, patient survival
+after 30 days from admission, highest hospital care level, GCS on
+arrival, respiratory rate on arrival, systolic blood pressure on
+arrival, time from arrival at the hospital until first computed
+tomography, if the patient was intubated on arrival and whether they
+arrived during weekend or working hours. Those are used as factor
+variables. ISS and age are used as continuous variables.
 
@@
 significance level and 95% confidence levels will be used.
+
+Rstudio was used to perform the logistical regression. The dependent
+variable is OFI_delay, which is the presence of the ofi delayed
+treatment.
+
+## Ethical considerations
+
+Ethical permit is required and exists. The ethical review number is
+2021-02541 and 2021-03531. An ethical problem that might be brought up
+is the anonymity of the patients. In the registry, patients social
+security number is present and could theoretically be used to identify
+the patient in question. This will not be a problem though, as in the
+analysis, the social security number is removed from the data before the
+logistical regression is ran. Furthermore, the results will be presented
+with patients grouped together, not individually. This will make it
+impossible for anyone to identify patients. During the analysis, a VPN
+was used to access the database to ensure that no data was leaked or
+used for other purposes.

### 2023-10-27 functions/view_data.R
code

@@ -1,4 +1,7 @@
 library(gtsummary)
+library(magrittr)
+library(dplyr)
+library(car)
 
 model.data <- function(dataset){
   
@@ -12,7 +15,12 @@ model.data <- function(dataset){
   
   
   ## view model
-  tbl_regression(model, exponentiate = TRUE)
+  model %>%
+    tbl_regression(exponentiate = TRUE) %>%
+    bold_p() %>%
+    bold_labels()
+    
+
 
   factors.data %>%
     tbl_summary(  # Create summary statistics for different groups
@@ -25,7 +33,6 @@ model.data <- function(dataset){
       statistic = list(all_continuous() ~ "{mean} ± {sd}"),
       digits = list(all_continuous() ~ c(2, 2))
     ) %>%
-    add_p()  # Add p-values
   
   
   return(model)

### 2023-10-30 functions/view_data.R
code

@@ -19,7 +19,7 @@ model.data <- function(dataset){
     tbl_regression(exponentiate = TRUE) %>%
     bold_p() %>%
     bold_labels()
-    
+    add_p(digits = c(p = 3))
 
 
   factors.data %>%

### 2023-10-30 functions/view_data.R
code

@@ -33,6 +33,10 @@ model.data <- function(dataset){
       statistic = list(all_continuous() ~ "{mean} ± {sd}"),
       digits = list(all_continuous() ~ c(2, 2))
     ) %>%
+    
+library(ggplot2)
+  ggplot(factors.data, aes(x = Age)) +
+    geom_density(alpha = .2, fill = "#FF6666")
   
   
   return(model)

### 2023-10-31 manuscriptlongversion.Rmd
flowchart

@@
 registry and trauma care quality database at the Karolinska University
-Hospital in Solna. The trauma registry includes about 14000 patients
+Hospital in Solna. The trauma registry includes about 11864 patients
 treated between 2012 and 2022. The trauma care quality database is a
@@
 used for other purposes.
+
+## Results
+
+Out of the 11864 patients treated at KS between 2012 and 2022, 1835 were
+included in the study. The patient characteristics were as follows in
+Table 1.
+
+```{r}
+  factors.data %>%
+    tbl_summary(  # Create summary statistics for different groups
+      by = OFI_delay,
+      label = list(
+        Age = "Age (years)",
+        Gender = "Gender",
+        Total_GCS = "GCS"
+      ),
+      statistic = list(all_continuous() ~ "{mean} ± {sd}"),
+      digits = list(all_continuous() ~ c(2, 2))
+    ) %>%
+    add_p()  # Add p-values
+
+```

### 2023-10-31 functions/Flowchart.R

@@ -1,0 +1,65 @@
+library(Gmisc, quietly = TRUE)
+library(glue)
+library(htmlTable)
+library(grid)
+library(magrittr)
+
+org_cohort <- boxGrob(glue("Treated at KS between 2012 and 2022",
+                           "n = {pop}",
+                           pop = txtInt(11,864),
+                           .sep = "\n"))
+eligible <- boxGrob(glue("Eligible",
+                         "n = {pop}",
+                         pop = txtInt(10032),
+                         .sep = "\n"))
+included <- boxGrob(glue("Randomized",
+                         "n = {incl}",
+                         incl = txtInt(122),
+                         .sep = "\n"))
+grp_a <- boxGrob(glue("Treatment A",
+                      "n = {recr}",
+                      recr = txtInt(43),
+                      .sep = "\n"))
+
+grp_b <- boxGrob(glue("Treatment B",
+                      "n = {recr}",
+                      recr = txtInt(122 - 43 - 30),
+                      .sep = "\n"))
+
+excluded <- boxGrob(glue("Excluded (n = {tot}):",
+                         " - not interested: {uninterested}",
+                         " - contra-indicated: {contra}",
+                         tot = 30,
+                         uninterested = 12,
+                         contra = 30 - 12,
+                         .sep = "\n"),
+                    just = "left")
+
+grid.newpage()
+vert <- spreadVertical(org_cohort,
+                       eligible = eligible,
+                       included = included,
+                       grps = grp_a)
+grps <- alignVertical(reference = vert$grps,
+                      grp_a, grp_b) %>%
+  spreadHorizontal()
+vert$grps <- NULL
+
+excluded <- moveBox(excluded,
+                    x = .8,
+                    y = coords(vert$included)$top + distance(vert$eligible, vert$included, half = TRUE, center = FALSE))
+
+for (i in 1:(length(vert) - 1)) {
+  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
+    print
+}
+connectGrob(vert$included, grps[[1]], type = "N")
+connectGrob(vert$included, grps[[2]], type = "N")
+
+connectGrob(vert$eligible, excluded, type = "L")
+
+# Print boxes
+vert
+grps
+excluded
+

### 2023-10-31 manuscriptlongversion.Rmd
flowchart

@@
 ```
+
+Out of the 11,864, 5,282 patients were not screened for ofi, which
+excludes them from the study. Out of the remaining 6,582 patients, 4,775
+patients had missing data in one or more of the analysed parameters,
+which led to exclusion from the study.
+
+In the remaining group of 1,835 patients who have been screened for ofi
+and have data registered in all parameters examined for the study, 28
+patients were found to have received delayed treatment during their stay
+at the hospital.
+
+```{r}
+library(Gmisc, quietly = TRUE)
+library(glue)
+library(htmlTable)
+library(grid)
+library(magrittr)
+library(gtsummary)
+
+treated <- boxGrob(glue("Treated at KS between 2012 and 2022",
+                           "n = {pop}",
+                           pop = txtInt(11864),
+                           .sep = "\n"))
+screened <- boxGrob(glue("Screened for ofi",
+                         "n = {pop}",
+                         pop = txtInt(6582),
+                         .sep = "\n"))
+included <- boxGrob(glue("Included in study",
+                         "n = {incl}",
+                         incl = txtInt(1835),
+                         .sep = "\n"))
+delay <- boxGrob(glue("Delay in treatment",
+                      "n = {recr}",
+                      recr = txtInt(28),
+                      .sep = "\n"))
+
+no_delay <- boxGrob(glue("No delay in treatment",
+                      "n = {recr}",
+                      recr = txtInt(1807),
+                      .sep = "\n"))
+
+not_screened <- boxGrob(glue("Not screened for ofi",
+                             "n = {recr}",
+                             recr = txtInt(11864-6582),
+                             .sep = "\n"))
+
+missing_data <- boxGrob(glue("Missing data",
+                             "n = {recr}",
+                             recr = txtInt(6582-1807),
+                             .sep = "\n"))
+grid.newpage()
+vert <- spreadVertical(treated = treated,
+                       screened = screened,
+                       included = included,
+                       grps = delay)
+grps <- alignVertical(reference = vert$grps,
+                      delay, no_delay) %>%
+  spreadHorizontal()
+vert$grps <- NULL
+
+not_screened <- moveBox(not_screened,
+                    x = .85,
+                    y = .77,)
+
+missing_data <- moveBox(missing_data,
+                    x = .85,
+                    y = .5,)
+
+for (i in 1:(length(vert) - 1)) {
+  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
+    print
+}
+connectGrob(vert$included, grps[[1]], type = "N")
+connectGrob(vert$included, grps[[2]], type = "N")
+connectGrob(vert$treated, not_screened, type = "L")
+connectGrob(vert$screened, missing_data, type = "L")
+
+# Print boxes
+vert
+grps
+not_screened
+missing_data
+
+```

### 2023-10-31 functions/Flowchart.R

@@ -3,51 +3,56 @@ library(glue)
 library(htmlTable)
 library(grid)
 library(magrittr)
+library(gtsummary)
 
-org_cohort <- boxGrob(glue("Treated at KS between 2012 and 2022",
+treated <- boxGrob(glue("Treated at KS between 2012 and 2022",
                            "n = {pop}",
-                           pop = txtInt(11,864),
+                           pop = txtInt(11864),
                            .sep = "\n"))
-eligible <- boxGrob(glue("Eligible",
+screened <- boxGrob(glue("Screened for ofi",
                          "n = {pop}",
-                         pop = txtInt(10032),
+                         pop = txtInt(6582),
                          .sep = "\n"))
-included <- boxGrob(glue("Randomized",
+included <- boxGrob(glue("Included in study",
                          "n = {incl}",
-                         incl = txtInt(122),
+                         incl = txtInt(1835),
                          .sep = "\n"))
-grp_a <- boxGrob(glue("Treatment A",
+delay <- boxGrob(glue("Delay in treatment",
                       "n = {recr}",
-                      recr = txtInt(43),
+                      recr = txtInt(28),
                       .sep = "\n"))
 
-grp_b <- boxGrob(glue("Treatment B",
+no_delay <- boxGrob(glue("No delay in treatment",
                       "n = {recr}",
-                      recr = txtInt(122 - 43 - 30),
+                      recr = txtInt(1807),
                       .sep = "\n"))
 
-excluded <- boxGrob(glue("Excluded (n = {tot}):",
-                         " - not interested: {uninterested}",
-                         " - contra-indicated: {contra}",
-                         tot = 30,
-                         uninterested = 12,
-                         contra = 30 - 12,
-                         .sep = "\n"),
-                    just = "left")
+not_screened <- boxGrob(glue("Not screened for ofi",
+                             "n = {recr}",
+                             recr = txtInt(11864-6582),
+                             .sep = "\n"))
 
+missing_data <- boxGrob(glue("Missing data",
+                             "n = {recr}",
+                             recr = txtInt(6582-1807),
+                             .sep = "\n"))
 grid.newpage()
-vert <- spreadVertical(org_cohort,
-                       eligible = eligible,
+vert <- spreadVertical(treated = treated,
+                       screened = screened,
                        included = included,
[…]

### 2023-11-02 functions/view_data.R
code

@@ -1,7 +1,6 @@
 library(gtsummary)
 library(magrittr)
 library(dplyr)
-library(car)
 
 model.data <- function(dataset){
   

### 2023-11-06 functions/create_factors.R
code

@@ -4,19 +4,19 @@ library(stringr)
 create.factors <- function(dataset){
  
   ## Making gcs into rts and replacing missing values
+
   dataset$ed_gcs_sum <- ifelse(dataset$ed_gcs_sum >= 3 & dataset$ed_gcs_sum <= 3, 0,
-                     ifelse(dataset$ed_gcs_sum >= 4 & dataset$ed_gcs_sum <= 5, 1,
-                            ifelse(dataset$ed_gcs_sum >= 6 & dataset$ed_gcs_sum <= 8, 2,
-                                   ifelse(dataset$ed_gcs_sum >= 9 & dataset$ed_gcs_sum <= 12, 3,
-                                          ifelse(dataset$ed_gcs_sum >= 13 & dataset$ed_gcs_sum <= 15, 4,
-                                                 ifelse(dataset$ed_gcs_sum == 99, "intubated", dataset$ed_gcs_sum))))))
+                               ifelse(dataset$ed_gcs_sum >= 4 & dataset$ed_gcs_sum <= 5, 1,
+                                      ifelse(dataset$ed_gcs_sum >= 6 & dataset$ed_gcs_sum <= 8, 2,
+                                             ifelse(dataset$ed_gcs_sum >= 9 & dataset$ed_gcs_sum <= 12, 3,
+                                                    ifelse(dataset$ed_gcs_sum >= 13 & dataset$ed_gcs_sum <= 15, 4, dataset$ed_gcs_sum)))))
   
   dataset$pre_gcs_sum <- ifelse(dataset$pre_gcs_sum >= 3 & dataset$pre_gcs_sum <= 3, 0,
-                                  ifelse(dataset$pre_gcs_sum >= 4 & dataset$pre_gcs_sum <= 5, 1,
-                                         ifelse(dataset$pre_gcs_sum >= 6 & dataset$pre_gcs_sum <= 8, 2,
-                                                ifelse(dataset$pre_gcs_sum >= 9 & dataset$pre_gcs_sum <= 12, 3,
-                                                       ifelse(dataset$pre_gcs_sum >= 13 & dataset$pre_gcs_sum <= 15, 4,
-                                                              ifelse(dataset$ed_gcs_sum == 99, "intubated", dataset$pre_gcs_sum))))))
+                                ifelse(dataset$pre_gcs_sum >= 4 & dataset$pre_gcs_sum <= 5, 1,
[…]

### 2023-11-06 functions/prepare_data.R

@@ -3,10 +3,12 @@ library(rofi)
 prepare.data <- function(){
   
   ## Import data
-  data <- rofi::import_data(test = TRUE)
+  data <- rofi::import_data(
+    user = "antonw",
+    password = "2-parts-Attack-applied")
     
   ## Create variable combined datasets by merging
-  combined.datasets <- rofi::merge_data(data, test = TRUE)
+  combined.datasets <- rofi::merge_data(data)
 
     ## Adding ofi as variable
   combined.datasets$ofi <- rofi::create_ofi(combined.datasets)

### 2023-11-06 functions/view_data.R

@@ -7,11 +7,17 @@ model.data <- function(dataset){
   ## regression model
   model <-
     glm(
-      OFI_delay ~ Total_GCS + Gender + survival_after_30_days + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Time_until_first_CT + Intubated_prehospitaly + ISS + Age + weekday + work_hours,
+      OFI_delay ~ Total_GCS + Gender + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Time_until_first_CT + Intubated_prehospitaly + ISS + Age + weekday + work_hours,
       data = factors.data,
       family = binomial
     )
-  
+
+  model1 <-
+    glm(
+      OFI_delay ~ Total_GCS,
+      data = factors.data,
+      family = binomial
+    )
   
   ## view model
   model %>%
@@ -33,10 +39,7 @@ model.data <- function(dataset){
       digits = list(all_continuous() ~ c(2, 2))
     ) %>%
     
-library(ggplot2)
-  ggplot(factors.data, aes(x = Age)) +
-    geom_density(alpha = .2, fill = "#FF6666")
-  
+
   
   return(model)
 }

### 2023-11-06 functions/new_subset.R
code

@@ -2,7 +2,10 @@ library(rofi)
 
 clean.dataset <- function(dataset){
   ## create new colum for outcome to be observed
-  dataset$OFI_delay <- as.factor(ifelse(dataset$Problemomrade_.FMP == "Handläggning", "Delay to treatment", "No delay to treatment"))
+  dataset$OFI_delay <- as.factor(ifelse(dataset$Problemomrade_.FMP == "Handläggning", "Delay to treatment",
+                                        ifelse(dataset$Problemomrade_.FMP == "Lång tid till DT", "Delay to treatment",
+                                               ifelse(dataset$Problemomrade_.FMP == "Lång tid till op", "Delay to treatment",
+                                                      "No delay to treatment"))))
   
   ## remove all columns except the analyzed ones
   dataset <-

### 2023-11-06 functions/view_data.R

@@ -14,13 +14,13 @@ model.data <- function(dataset){
 
   model1 <-
     glm(
-      OFI_delay ~ Total_GCS,
+      OFI_delay ~ Gender,
       data = factors.data,
       family = binomial
     )
   
   ## view model
-  model %>%
+  model1 %>%
     tbl_regression(exponentiate = TRUE) %>%
     bold_p() %>%
     bold_labels()

### 2023-11-07 functions/create_factors.R
code

@@ -51,15 +51,17 @@ create.factors <- function(dataset){
   dataset$Highest_care_level <- as.factor(dataset$host_care_level)
   
   ## Respiratory rate into rts and replacing missing values
-  dataset$ed_rr_value <- ifelse(dataset$ed_rr_value >= 0 & dataset$ed_rr_value <= 0, 0,
-                                  ifelse(dataset$ed_rr_value >= 1 & dataset$ed_rr_value <= 5, 1,
-                                         ifelse(dataset$ed_rr_value >= 6 & dataset$ed_rr_value <= 9, 2,
-                                                ifelse(dataset$ed_rr_value >= 30 & dataset$ed_rr_value <= 98, 3,
-                                                       ifelse(dataset$ed_rr_value >= 10 & dataset$ed_rr_value <= 29, 4,  dataset$ed_rr_value)))))
+  dataset$ed_rr_value <- ifelse(dataset$ed_rr_value == 0, 0,
+                                ifelse(dataset$ed_rr_value >= 1 & dataset$ed_rr_value <= 5, 1,
+                                       ifelse(dataset$ed_rr_value >= 6 & dataset$ed_rr_value <= 9, 2,
+                                              ifelse(dataset$ed_rr_value >= 30 & dataset$ed_rr_value <= 98, 3,
+                                                     ifelse(dataset$ed_rr_value >= 10 & dataset$ed_rr_value <= 29, 4,
+                                                            ifelse(dataset$ed_rr_value == 99, "missing",
+                                                                   ifelse(is.na(dataset$ed_rr_value), "missing", dataset$ed_rr_value)))))))
   
-  dataset$ed_rr_value <- ifelse(is.na(dataset$ed_rr_value) | dataset$ed_rr_value == 999 | dataset$ed_rr_value == 99, dataset$ed_rr_rtscat, dataset$ed_rr_value)  
+  dataset$ed_rr_value <- ifelse(dataset$ed_rr_value == 999, dataset$ed_rr_rtscat, dataset$ed_rr_value)  
   
-  dataset <- subset(dataset, !(is.na(ed_rr_value) | ed_rr_value == 999 | ed_rr_value == 99))
+  dataset <- subset(dataset, !(ed_rr_value == 999))
   
   dataset$Respiratory_rate <- as.factor(dataset$ed_rr_value)
   

### 2023-11-07 functions/view_data.R

@@ -14,13 +14,20 @@ model.data <- function(dataset){
 
   model1 <-
     glm(
-      OFI_delay ~ Gender,
+      OFI_delay ~Gender + ISS + Age + work_hours + Highest_care_level,
+      data = factors.data,
+      family = binomial
+    )
+  
+  model2 <-
+    glm(
+      OFI_delay ~ Intubated_prehospitaly,
       data = factors.data,
       family = binomial
     )
   
   ## view model
-  model1 %>%
+  model2 %>%
     tbl_regression(exponentiate = TRUE) %>%
     bold_p() %>%
     bold_labels()

### 2023-11-07 functions/view_data.R
code

@@ -14,14 +14,14 @@ model.data <- function(dataset){
 
   model1 <-
     glm(
-      OFI_delay ~Gender + ISS + Age + work_hours + Highest_care_level,
+      OFI_delay ~Gender + ISS + Age + Highest_care_level + weekday + work_hours
       data = factors.data,
       family = binomial
     )
   
   model2 <-
     glm(
-      OFI_delay ~ Intubated_prehospitaly,
+      OFI_delay ~ Time_until_first_CT,
       data = factors.data,
       family = binomial
     )

### 2023-11-07 functions/CopyOfcreate_factors1.R
code

@@ -4,13 +4,25 @@ library(stringr)
 create.factors <- function(dataset){
  
   ## Making gcs into rts and replacing missing values
-  dataset$pre_gcs_sum <- ifelse(dataset$pre_gcs_sum == 99, "missing", dataset$pre_gcs_sum)
+
+  dataset$ed_gcs_sum <- ifelse(dataset$ed_gcs_sum >= 3 & dataset$ed_gcs_sum <= 3, 0,
+                               ifelse(dataset$ed_gcs_sum >= 4 & dataset$ed_gcs_sum <= 5, 1,
+                                      ifelse(dataset$ed_gcs_sum >= 6 & dataset$ed_gcs_sum <= 8, 2,
+                                             ifelse(dataset$ed_gcs_sum >= 9 & dataset$ed_gcs_sum <= 12, 3,
+                                                    ifelse(dataset$ed_gcs_sum >= 13 & dataset$ed_gcs_sum <= 15, 4, dataset$ed_gcs_sum)))))
   
-  dataset$pre_gcs_sum <- ifelse(is.na(dataset$pre_gcs_sum), "missing", dataset$pre_gcs_sum)
+  dataset$pre_gcs_sum <- ifelse(dataset$pre_gcs_sum >= 3 & dataset$pre_gcs_sum <= 3, 0,
+                                ifelse(dataset$pre_gcs_sum >= 4 & dataset$pre_gcs_sum <= 5, 1,
+                                       ifelse(dataset$pre_gcs_sum >= 6 & dataset$pre_gcs_sum <= 8, 2,
+                                              ifelse(dataset$pre_gcs_sum >= 9 & dataset$pre_gcs_sum <= 12, 3,
+                                                     ifelse(dataset$pre_gcs_sum >= 13 & dataset$pre_gcs_sum <= 15, 4, dataset$ed_gcs_sum)))))
+  
+                                  
+  dataset$ed_gcs_sum <- ifelse(is.na(dataset$ed_gcs_sum) | dataset$ed_gcs_sum == 999, dataset$pre_gcs_sum, dataset$ed_gcs_sum)  
   
   dataset <- subset(dataset, !(is.na(ed_gcs_sum) | ed_gcs_sum == 999))
   
-  dataset$Total_GCS <- as.numeric(dataset$ed_gcs_sum)
+  dataset$Total_GCS <- as.factor(dataset$ed_gcs_sum)
   
   ## Making gender into a factor
[…]

### 2023-11-07 functions/new_subset.R

@@ -2,10 +2,9 @@ library(rofi)
 
 clean.dataset <- function(dataset){
   ## create new colum for outcome to be observed
-  dataset$OFI_delay <- as.factor(ifelse(dataset$Problemomrade_.FMP == "Handläggning", "Delay to treatment",
-                                        ifelse(dataset$Problemomrade_.FMP == "Lång tid till DT", "Delay to treatment",
+  dataset$OFI_delay <- as.factor(ifelse(dataset$Problemomrade_.FMP == "Lång tid till DT", "Delay to treatment",
                                                ifelse(dataset$Problemomrade_.FMP == "Lång tid till op", "Delay to treatment",
-                                                      "No delay to treatment"))))
+                                                      "No delay to treatment")))
   
   ## remove all columns except the analyzed ones
   dataset <-

### 2023-11-07 functions/view_data.R

@@ -14,12 +14,12 @@ model.data <- function(dataset){
 
   model1 <-
     glm(
-      OFI_delay ~Gender + ISS + Age + Highest_care_level + weekday + work_hours
+      OFI_delay ~Gender + ISS + Age + Highest_care_level + weekday + work_hours,
       data = factors.data,
       family = binomial
     )
   
-  model2 <-
+  model <-
     glm(
       OFI_delay ~ Time_until_first_CT,
       data = factors.data,
@@ -27,7 +27,7 @@ model.data <- function(dataset){
     )
   
   ## view model
-  model2 %>%
+  model %>%
     tbl_regression(exponentiate = TRUE) %>%
     bold_p() %>%
     bold_labels()

### 2023-11-07 functions/create_factors.R
code

@@ -4,11 +4,10 @@ library(stringr)
 create.factors <- function(dataset){
  
   ## Making gcs into rts and replacing missing values
-  dataset$pre_gcs_sum <- ifelse(dataset$pre_gcs_sum == 99, "missing", dataset$pre_gcs_sum)
+  dataset <- dataset %>%
+    mutate(ed_gcs_sum = ifelse(is.na(ed_gcs_sum), pre_gcs_sum, ed_gcs_sum))
   
-  dataset$pre_gcs_sum <- ifelse(is.na(dataset$pre_gcs_sum), "missing", dataset$pre_gcs_sum)
-  
-  dataset <- subset(dataset, !(is.na(ed_gcs_sum) | ed_gcs_sum == 999))
+  dataset$ed_gcs_sum <- ifelse(dataset$ed_gcs_sum == 999 | dataset$ed_gcs_sum == 99, dataset$pre_gcs_sum, dataset$ed_gcs_sum)
   
   dataset$Total_GCS <- as.numeric(dataset$ed_gcs_sum)
   
@@ -39,17 +38,19 @@ create.factors <- function(dataset){
   dataset$Highest_care_level <- as.factor(dataset$host_care_level)
   
   ## Respiratory rate into rts and replacing missing values
-  dataset$ed_rr_value <- ifelse(dataset$ed_rr_value == 99, "missing", dataset$ed_rr_value)
+  dataset <- dataset %>%
+    mutate(ed_rr_value = coalesce(ed_rr_value, pre_rr_value))
   
-  dataset$ed_rr_value <- ifelse(is.na(dataset$ed_rr_value), "missing", dataset$ed_rr_value)
-  
-  dataset <- subset(dataset, !ed_rr_value == 999)
+  dataset$ed_rr_value <- ifelse(dataset$ed_rr_value == 999 | dataset$ed_rr_value == 99, dataset$pre_rr_value, dataset$ed_rr_value)
   
   dataset$Respiratory_rate <- as.numeric(dataset$ed_rr_value)
   
   ## SBP into rts and replacing missing values
+  dataset <- dataset %>%
+    mutate(ed_sbp_value = ifelse(is.na(ed_sbp_value), pre_gcs_sum, ed_sbp_value))
   
-  dataset <- subset(dataset, !(is.na(ed_sbp_value) | ed_sbp_value == 999))
[…]

### 2023-11-07 functions/view_data.R

@@ -14,7 +14,7 @@ model.data <- function(dataset){
 
   model1 <-
     glm(
-      OFI_delay ~Gender + ISS + Age + Highest_care_level + weekday + work_hours,
+      OFI_delay ~Gender + ISS + Age + Highest_care_level + weekday + work_hours + Intubated_prehospitaly, 
       data = factors.data,
       family = binomial
     )
@@ -27,7 +27,7 @@ model.data <- function(dataset){
     )
   
   ## view model
-  model %>%
+  model1 %>%
     tbl_regression(exponentiate = TRUE) %>%
     bold_p() %>%
     bold_labels()

### 2023-11-07 functions/CopyOfcreate_factors1.R
code

@@ -59,7 +59,7 @@ create.factors <- function(dataset){
                                                             ifelse(dataset$ed_rr_value == 99, "missing",
                                                                    ifelse(is.na(dataset$ed_rr_value), "missing", dataset$ed_rr_value)))))))
   
-  dataset$ed_rr_value <- ifelse(dataset$ed_rr_value == 999, dataset$ed_rr_rtscat, dataset$ed_rr_value)  
+ dataset$ed_rr_value <- ifelse(is.na(dataset$ed_rr_value) | dataset$ed_rr_value == 999 | dataset$ed_rr_value == 99, dataset$ed_rr_rtscat, dataset$ed_rr_value) 
   
   dataset <- subset(dataset, !(ed_rr_value == 999))
   

### 2023-11-07 functions/create_factors.R

@@ -4,10 +4,9 @@ library(stringr)
 create.factors <- function(dataset){
  
   ## Making gcs into rts and replacing missing values
-  dataset <- dataset %>%
-    mutate(ed_gcs_sum = ifelse(is.na(ed_gcs_sum), pre_gcs_sum, ed_gcs_sum))
   
-  dataset$ed_gcs_sum <- ifelse(dataset$ed_gcs_sum == 999 | dataset$ed_gcs_sum == 99, dataset$pre_gcs_sum, dataset$ed_gcs_sum)
+  dataset <- dataset %>%
+    mutate(ed_gcs_sum = if_else(is.na(ed_gcs_sum) | ed_gcs_sum == 999 | ed_gcs_sum == 99, pre_gcs_sum, ed_gcs_sum))
   
   dataset$Total_GCS <- as.numeric(dataset$ed_gcs_sum)
   
@@ -39,18 +38,14 @@ create.factors <- function(dataset){
   
   ## Respiratory rate into rts and replacing missing values
   dataset <- dataset %>%
-    mutate(ed_rr_value = coalesce(ed_rr_value, pre_rr_value))
-  
-  dataset$ed_rr_value <- ifelse(dataset$ed_rr_value == 999 | dataset$ed_rr_value == 99, dataset$pre_rr_value, dataset$ed_rr_value)
+    mutate(ed_rr_value = if_else(is.na(ed_rr_value) | ed_rr_value == 999 | ed_rr_value == 99, pre_rr_value, ed_rr_value))
   
   dataset$Respiratory_rate <- as.numeric(dataset$ed_rr_value)
   
   ## SBP into rts and replacing missing values
+  
   dataset <- dataset %>%
-    mutate(ed_sbp_value = ifelse(is.na(ed_sbp_value), pre_gcs_sum, ed_sbp_value))
-  
-  dataset$ed_sbp_value <- ifelse(dataset$ed_sbp_value == 999 | dataset$ed_sbp_value == 99, dataset$pre_sbp_value, dataset$ed_sbp_value)
-  
+    mutate(ed_sbp_value = if_else(is.na(ed_sbp_value) | ed_sbp_value == 999 | ed_sbp_value == 99, pre_sbp_value, ed_sbp_value))
   
   dataset$Systolic_blood_pressure <- as.numeric(dataset$ed_sbp_value)
   
@@ -121,7 +116,9 @@ create.factors <- function(dataset){
         "ed_sbp_rtscat",
[…]

### 2023-11-07 functions/new_subset.R

@@ -24,7 +24,9 @@ clean.dataset <- function(dataset){
       "pre_gcs_sum",
       "ed_rr_rtscat",
       "ed_sbp_rtscat",
-      "DateTime_ArrivalAtHospital"
+      "DateTime_ArrivalAtHospital",
+      "pre_rr_value",
+      "pre_sbp_value"
     )]
   
   ## remove rows without OFI

### 2023-11-07 functions/view_data.R

@@ -7,31 +7,49 @@ model.data <- function(dataset){
   ## regression model
   model <-
     glm(
-      OFI_delay ~ Total_GCS + Gender + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Time_until_first_CT + Intubated_prehospitaly + ISS + Age + weekday + work_hours,
+      OFI_delay ~ Total_GCS + Gender + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Intubated_prehospitaly + ISS + Age + weekday + work_hours,
       data = factors.data,
       family = binomial
     )
 
-  model1 <-
-    glm(
-      OFI_delay ~Gender + ISS + Age + Highest_care_level + weekday + work_hours + Intubated_prehospitaly, 
-      data = factors.data,
-      family = binomial
+ # Your predictor variables
+  predictor_variables <- c("Total_GCS", "Gender", "Highest_care_level", "Respiratory_rate", "Systolic_blood_pressure", "Intubated_prehospitaly", "ISS", "Age", "weekday", "work_hours")
+  
+  # Loop through predictor variables and store regression results
+  results_list <- lapply(predictor_variables, function(var) {
+    formula <- as.formula(paste("OFI_delay ~", var))
+    res.logist <- glm(formula, data = factors.data, family = binomial)
+    
+    # Store the results summary in the list
+    summary(res.logist)
+  })
+  
+  result_data <- do.call(rbind, lapply(results_list, function(res) {
+    cbind(
+      variable = rownames(coef(res)),
+      coef(res),
+      p_values <- res$coefficients[, "Pr(>|z|)"]
+    )
+  }))
+  
+  fancy_table <-
+    tbl_merge(
+      tbls        = list(model, result_data),
+      tab_spanner = c("Adjusted", "Unadjusted")
     )
   
-  model <-
-    glm(
-      OFI_delay ~ Time_until_first_CT,
-      data = factors.data,
-      family = binomial
-    )
+  fancy_table
+  
+  
   
   ## view model
-  model1 %>%
-    tbl_regression(exponentiate = TRUE) %>%
+  model %>%
+    tbl_regression(exponentiate = TRUE,
+        pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
     bold_p() %>%
     bold_labels()
     add_p(digits = c(p = 3))
+    
 
 
   factors.data %>%
[…]

### 2023-11-08 functions/create_factors.R
code

@@ -118,7 +118,10 @@ create.factors <- function(dataset){
         "Date",
         "Time",
         "pre_rr_value",
-        "pre_sbp_value"
+        "pre_sbp_value",
+        "ofi",
+        "survival_after_30_days",
+        "Time_until_first_CT"
       )
     )]
 

### 2023-11-08 functions/view_data.R

@@ -9,46 +9,39 @@ model.data <- function(dataset){
     glm(
       OFI_delay ~ Total_GCS + Gender + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Intubated_prehospitaly + ISS + Age + weekday + work_hours,
       data = factors.data,
-      family = binomial
-    )
-
- # Your predictor variables
-  predictor_variables <- c("Total_GCS", "Gender", "Highest_care_level", "Respiratory_rate", "Systolic_blood_pressure", "Intubated_prehospitaly", "ISS", "Age", "weekday", "work_hours")
+      family = binomial)
+      
+      adjusted <-
+        tbl_regression(model, exponentiate = TRUE,
+                       pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
+        bold_p() %>%
+        bold_labels()
+      add_p(digits = c(p = 3))
+}
   
-  # Loop through predictor variables and store regression results
-  results_list <- lapply(predictor_variables, function(var) {
-    formula <- as.formula(paste("OFI_delay ~", var))
-    res.logist <- glm(formula, data = factors.data, family = binomial)
+  unadjust2 <- tbl_uvregression(data = factors.data,
+                                method = glm,
+                                y = OFI_delay,
+                                method.args = list(family = binomial),
+                                pvalue_fun = ~ style_pvalue(.x, digits = 2),
+                                hide_n = TRUE
+  ) %>%
+    bold_p(t = 0.05) %>%
+    bold_labels()
     
-    # Store the results summary in the list
-    summary(res.logist)
-  })
-  
-  result_data <- do.call(rbind, lapply(results_list, function(res) {
-    cbind(
-      variable = rownames(coef(res)),
-      coef(res),
-      p_values <- res$coefficients[, "Pr(>|z|)"]
-    )
-  }))
-  
-  fancy_table <-
+
+fancy_table <-
     tbl_merge(
-      tbls        = list(model, result_data),
+      tbls        = list(adjusted, unadjust2),
       tab_spanner = c("Adjusted", "Unadjusted")
     )
   
-  fancy_table
+  
   
   
   
   ## view model
-  model %>%
-    tbl_regression(exponentiate = TRUE,
-        pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
-    bold_p() %>%
-    bold_labels()
-    add_p(digits = c(p = 3))
+
     
 
 
@@ -68,4 +61,4 @@ model.data <- function(dataset){
 
   
   return(model)
-}
+

### 2023-11-08 manuscriptlongversion.Rmd
code

--- 2023-10-31 manuscriptlongversion.Rmd
@@
 
-```{r}
-  factors.data %>%
-    tbl_summary(  # Create summary statistics for different groups
-      by = OFI_delay,
-      label = list(
-        Age = "Age (years)",
-        Gender = "Gender",
-        Total_GCS = "GCS"
-      ),
-      statistic = list(all_continuous() ~ "{mean} ± {sd}"),
-      digits = list(all_continuous() ~ c(2, 2))
-    ) %>%
-    add_p()  # Add p-values
+![](table1.png)
 
-```
+Out of the 11,864, 5,282 patients were not screened for ofi, which
+excludes them from the study. Out of the remaining 6,582 patients, 4,775
+patients had missing data in one or more of the analysed parameters,
+which led to exclusion from the study.
+
+In the remaining group of 1,835 patients who have been screened for ofi
+and have data registered in all parameters examined for the study, 28
+patients were found to have received delayed treatment during their stay
+at the hospital.
+
+![](Rplot.png)
+
+In the following table, the results from the regression analysis is
+displayed.
+
+The odds ratio is calculated, as well as a 95% confidence interval. The
+p value is also shown.

### 2023-11-08 manuscriptlongversion.tex
code

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx}
@@
 
+Rstudio was used to perform the logistical regression. The dependent
+variable is OFI\_delay, which is the presence of the ofi delayed
+treatment.
+
+\hypertarget{ethical-considerations}{%
+\subsection{Ethical considerations}\label{ethical-considerations}}
+
+Ethical permit is required and exists. The ethical review number is
+2021-02541 and 2021-03531. An ethical problem that might be brought up
+is the anonymity of the patients. In the registry, patients social
+security number is present and could theoretically be used to identify
+the patient in question. This will not be a problem though, as in the
+analysis, the social security number is removed from the data before the
+logistical regression is ran. Furthermore, the results will be presented
+with patients grouped together, not individually. This will make it
+impossible for anyone to identify patients. During the analysis, a VPN
+was used to access the database to ensure that no data was leaked or
+used for other purposes.
+
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
+
+Out of the 11864 patients treated at KS between 2012 and 2022, 1835 were
+included in the study. The patient characteristics were as follows in
+Table 1.
+
 \includegraphics{table1.html}
 
-\hypertarget{ethical-considerations}{%
-\subsection*{Ethical considerations}\label{ethical-considerations}}
-\addcontentsline{toc}{subsection}{Ethical considerations}
+Out of the 11,864, 5,282 patients were not screened for ofi, which
+excludes them from the study. Out of the remaining 6,582 patients, 4,775
+patients had missing data in one or more of the analysed parameters,
+which led to exclusion from the study.
+
+In the remaining group of 1,835 patients who have been screened for ofi
+and have data registered in all parameters examined for the study, 28
+patients were found to have received delayed treatment during their stay
+at the hospital.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(Gmisc, }\AttributeTok{quietly =} \ConstantTok{TRUE}\NormalTok{)}
+\FunctionTok{library}\NormalTok{(glue)}
+\FunctionTok{library}\NormalTok{(htmlTable)}
+\FunctionTok{library}\NormalTok{(grid)}
+\FunctionTok{library}\NormalTok{(magrittr)}
+\FunctionTok{library}\NormalTok{(gtsummary)}
+
+\NormalTok{treated }\OtherTok{\textless{}{-}} \FunctionTok{boxGrob}\NormalTok{(}\FunctionTok{glue}\NormalTok{(}\StringTok{"Treated at KS between 2012 and 2022"}\NormalTok{,}
+                           \StringTok{"n = \{pop\}"}\NormalTok{,}
+                           \AttributeTok{pop =} \FunctionTok{txtInt}\NormalTok{(}\DecValTok{11864}\NormalTok{),}
+                           \AttributeTok{.sep =} \StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{))}
+\NormalTok{screened }\OtherTok{\textless{}{-}} \FunctionTok{boxGrob}\NormalTok{(}\FunctionTok{glue}\NormalTok{(}\StringTok{"Screened for ofi"}\NormalTok{,}
+                         \StringTok{"n = \{pop\}"}\NormalTok{,}
+                         \AttributeTok{pop =} \FunctionTok{txtInt}\NormalTok{(}\DecValTok{6582}\NormalTok{),}
+                         \AttributeTok{.sep =} \StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{))}
+\NormalTok{included }\OtherTok{\textless{}{-}} \FunctionTok{boxGrob}\NormalTok{(}\FunctionTok{glue}\NormalTok{(}\StringTok{"Included in study"}\NormalTok{,}
+                         \StringTok{"n = \{incl\}"}\NormalTok{,}
+                         \AttributeTok{incl =} \FunctionTok{txtInt}\NormalTok{(}\DecValTok{1835}\NormalTok{),}
+                         \AttributeTok{.sep =} \StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{))}
+\NormalTok{delay }\OtherTok{\textless{}{-}} \FunctionTok{boxGrob}\NormalTok{(}\FunctionTok{glue}\NormalTok{(}\StringTok{"Delay in treatment"}\NormalTok{,}
+                      \StringTok{"n = \{recr\}"}\NormalTok{,}
+                      \AttributeTok{recr =} \FunctionTok{txtInt}\NormalTok{(}\DecValTok{28}\NormalTok{),}
+                      \AttributeTok{.sep =} \StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{))}
+
+\NormalTok{no\_delay }\OtherTok{\textless{}{-}} \FunctionTok{boxGrob}\NormalTok{(}\FunctionTok{glue}\NormalTok{(}\StringTok{"No delay in treatment"}\NormalTok{,}
+                      \StringTok{"n = \{recr\}"}\NormalTok{,}
+                      \AttributeTok{recr =} \FunctionTok{txtInt}\NormalTok{(}\DecValTok{1807}\NormalTok{),}
+                      \AttributeTok{.sep =} \StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{))}
+
+\NormalTok{not\_screened }\OtherTok{\textless{}{-}} \FunctionTok{boxGrob}\NormalTok{(}\FunctionTok{glue}\NormalTok{(}\StringTok{"Not screened for ofi"}\NormalTok{,}
+                             \StringTok{"n = \{recr\}"}\NormalTok{,}
+                             \AttributeTok{recr =} \FunctionTok{txtInt}\NormalTok{(}\DecValTok{11864{-}6582}\NormalTok{),}
+                             \AttributeTok{.sep =} \StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{))}
+
+\NormalTok{missing\_data }\OtherTok{\textless{}{-}} \FunctionTok{boxGrob}\NormalTok{(}\FunctionTok{glue}\NormalTok{(}\StringTok{"Missing data"}\NormalTok{,}
+                             \StringTok{"n = \{recr\}"}\NormalTok{,}
+                             \AttributeTok{recr =} \FunctionTok{txtInt}\NormalTok{(}\DecValTok{6582{-}1807}\NormalTok{),}
+                             \AttributeTok{.sep =} \StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{))}
+\FunctionTok{grid.newpage}\NormalTok{()}
+\NormalTok{vert }\OtherTok{\textless{}{-}} \FunctionTok{spreadVertical}\NormalTok{(}\AttributeTok{treated =}\NormalTok{ treated,}
+                       \AttributeTok{screened =}\NormalTok{ screened,}
+                       \AttributeTok{included =}\NormalTok{ included,}
+                       \AttributeTok{grps =}\NormalTok{ delay)}
+\NormalTok{grps }\OtherTok{\textless{}{-}} \FunctionTok{alignVertical}\NormalTok{(}\AttributeTok{reference =}\NormalTok{ vert}\SpecialCharTok{$}\NormalTok{grps,}
+\NormalTok{                      delay, no\_delay) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{spreadHorizontal}\NormalTok{()}
+\NormalTok{vert}\SpecialCharTok{$}\NormalTok{grps }\OtherTok{\textless{}{-}} \ConstantTok{NULL}
+
+\NormalTok{not\_screened }\OtherTok{\textless{}{-}} \FunctionTok{moveBox}\NormalTok{(not\_screened,}
+                    \AttributeTok{x =}\NormalTok{ .}\DecValTok{85}\NormalTok{,}
+                    \AttributeTok{y =}\NormalTok{ .}\DecValTok{77}\NormalTok{,)}
+
+\NormalTok{missing\_data }\OtherTok{\textless{}{-}} \FunctionTok{moveBox}\NormalTok{(missing\_data,}
+                    \AttributeTok{x =}\NormalTok{ .}\DecValTok{85}\NormalTok{,}
+                    \AttributeTok{y =}\NormalTok{ .}\DecValTok{5}\NormalTok{,)}
+
+\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{(}\FunctionTok{length}\NormalTok{(vert) }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{)) \{}
+  \FunctionTok{connectGrob}\NormalTok{(vert[[i]], vert[[i }\SpecialCharTok{+} \DecValTok{1}\NormalTok{]], }\AttributeTok{type =} \StringTok{"vert"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+\NormalTok{    print}
+\NormalTok{\}}
+\FunctionTok{connectGrob}\NormalTok{(vert}\SpecialCharTok{$}\NormalTok{included, grps[[}\DecValTok{1}\NormalTok{]], }\AttributeTok{type =} \StringTok{"N"}\NormalTok{)}
+\FunctionTok{connectGrob}\NormalTok{(vert}\SpecialCharTok{$}\NormalTok{included, grps[[}\DecValTok{2}\NormalTok{]], }\AttributeTok{type =} \StringTok{"N"}\NormalTok{)}
+\FunctionTok{connectGrob}\NormalTok{(vert}\SpecialCharTok{$}\NormalTok{treated, not\_screened, }\AttributeTok{type =} \StringTok{"L"}\NormalTok{)}
+\FunctionTok{connectGrob}\NormalTok{(vert}\SpecialCharTok{$}\NormalTok{screened, missing\_data, }\AttributeTok{type =} \StringTok{"L"}\NormalTok{)}
+
+\CommentTok{\# Print boxes}
+\NormalTok{vert}
+\NormalTok{grps}
+\NormalTok{not\_screened}
+\NormalTok{missing\_data}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{manuscriptlongversion_files/figure-latex/unnamed-chunk-1-1.pdf}
+
+In the following table, the results from the regression analysis is
+displayed.
+
+The odds ratio is calculated, as well as a 95\% confidence interval. The
+p value is also shown.
+
+\includegraphics{results.html}
 

### 2023-11-08 Flowchart.R

@@ -1,71 +1,0 @@
-library(Gmisc, quietly = TRUE)
-library(glue)
-library(htmlTable)
-library(grid)
-library(magrittr)
-library(gtsummary)
-
-treated <- boxGrob(glue("Treated at KS between 2012 and 2022",
-                           "n = {pop}",
-                           pop = txtInt(11864),
-                           .sep = "\n"))
-screened <- boxGrob(glue("Screened for ofi",
-                         "n = {pop}",
-                         pop = txtInt(6582),
-                         .sep = "\n"))
-included <- boxGrob(glue("Included in study",
-                         "n = {incl}",
-                         incl = txtInt(1835),
-                         .sep = "\n"))
-delay <- boxGrob(glue("Delay in treatment",
-                      "n = {recr}",
-                      recr = txtInt(28),
-                      .sep = "\n"))
-
-no_delay <- boxGrob(glue("No delay in treatment",
-                      "n = {recr}",
-                      recr = txtInt(1807),
-                      .sep = "\n"))
-
-not_screened <- boxGrob(glue("Not screened for ofi",
-                             "n = {recr}",
-                             recr = txtInt(11864-6582),
-                             .sep = "\n"))
-
-missing_data <- boxGrob(glue("Missing data",
-                             "n = {recr}",
-                             recr = txtInt(6582-1807),
-                             .sep = "\n"))
-grid.newpage()
-vert <- spreadVertical(treated = treated,
-                       screened = screened,
-                       included = included,
-                       grps = delay)
-grps <- alignVertical(reference = vert$grps,
-                      delay, no_delay) %>%
-  spreadHorizontal()
-vert$grps <- NULL
-
-not_screened <- moveBox(not_screened,
-                    x = .85,
-                    y = .75,)
-
-missing_data <- moveBox(missing_data,
-                    x = .85,
-                    y = .5,)
-
-for (i in 1:(length(vert) - 1)) {
-  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
-    print
-}
-connectGrob(vert$included, grps[[1]], type = "N")
-connectGrob(vert$included, grps[[2]], type = "N")
-connectGrob(vert$treated, not_screened, type = "L")
-connectGrob(vert$screened, missing_data, type = "L")
-
-# Print boxes
-vert
-grps
-not_screened
-missing_data
-

### 2023-11-08 example_function.R

@@ -1,15 +1,71 @@
-library(rofi)
+library(Gmisc, quietly = TRUE)
+library(glue)
+library(htmlTable)
+library(grid)
+library(magrittr)
+library(gtsummary)
 
-subset.datas <- prepared.data[, c("ofi", "Problemomrade_.FMP")]
+treated <- boxGrob(glue("Treated at KS between 2012 and 2022",
+                        "n = {pop}",
+                        pop = txtInt(14022),
+                        .sep = "\n"))
+screened <- boxGrob(glue("Screened for ofi",
+                         "n = {pop}",
+                         pop = txtInt(8312),
+                         .sep = "\n"))
+included <- boxGrob(glue("Included in study",
+                         "n = {incl}",
+                         incl = txtInt(6489),
+                         .sep = "\n"))
+delay <- boxGrob(glue("Delay in treatment",
+                      "n = {recr}",
+                      recr = txtInt(69),
+                      .sep = "\n"))
 
-table(prepared.data$Problemomrade_.FMP)
+no_delay <- boxGrob(glue("No delay in treatment",
+                         "n = {recr}",
+                         recr = txtInt(6420),
+                         .sep = "\n"))
 
+not_screened <- boxGrob(glue("Not screened for ofi",
+                             "n = {recr}",
+                             recr = txtInt(14022-8312),
+                             .sep = "\n"))
 
-prepared.data$ofi <- ifelse(prepared.data$ofi == "success", 1, 0)
-cleaned_data$Gender <- ifelse(cleaned_data$Gender == "M", 1, 0)
+missing_data <- boxGrob(glue("Missing data",
+                             "n = {recr}",
+                             recr = txtInt(8312-6489),
+                             .sep = "\n"))
+grid.newpage()
+vert <- spreadVertical(treated = treated,
+                       screened = screened,
+                       included = included,
+                       grps = delay)
+grps <- alignVertical(reference = vert$grps,
+                      delay, no_delay) %>%
+  spreadHorizontal()
+vert$grps <- NULL
 
+not_screened <- moveBox(not_screened,
+                        x = .85,
+                        y = .77,)
 
+missing_data <- moveBox(missing_data,
+                        x = .85,
+                        y = .5,)
 
[…]

### 2023-11-08 functions/create_factors.R

@@ -52,11 +52,15 @@ create.factors <- function(dataset){
   ##Intubation sorting and missing values
   dataset <- subset(dataset, !(pre_intubated == 999))
   
-  dataset$pre_intubated <- ifelse(is.na(dataset$pre_intubated), "No", dataset$pre_intubated)
+  dataset$ed_intubated <- ifelse(dataset$ed_intubated == 1, 3, 2)
   
-  dataset$pre_intubated <- ifelse(dataset$pre_intubated == 1, "Yes", "No")
+  dataset <- dataset %>%
+    mutate(pre_intubated = if_else(is.na(pre_intubated) | pre_intubated == 2, ed_intubated, pre_intubated))
   
-  dataset$Intubated_prehospitaly <- as.factor(dataset$pre_intubated)
+  dataset$Intubated <- as.factor(ifelse(dataset$pre_intubated == 1, "Intubated prehospitaly",
+                                        ifelse(dataset$pre_intubated == 2, "Not intubated",
+                                               ifelse(dataset$pre_intubated == 3, "Intubated at hosptital",
+                                               "unknown"))))
   
   ##CT scan categorizing
   dataset$dt_ed_first_ct <- ifelse(dataset$dt_ed_first_ct >= 1 & dataset$dt_ed_first_ct <= 30, "1-30 mins",
@@ -121,7 +125,8 @@ create.factors <- function(dataset){
         "pre_sbp_value",
         "ofi",
         "survival_after_30_days",
-        "Time_until_first_CT"
+        "Time_until_first_CT",
+        "ed_intubated"
       )
     )]
 

### 2023-11-08 functions/new_subset.R

@@ -26,7 +26,8 @@ clean.dataset <- function(dataset){
       "ed_sbp_rtscat",
       "DateTime_ArrivalAtHospital",
       "pre_rr_value",
-      "pre_sbp_value"
+      "pre_sbp_value",
+      "ed_intubated"
     )]
   
   ## remove rows without OFI

### 2023-11-08 functions/view_data.R

@@ -7,7 +7,7 @@ model.data <- function(dataset){
   ## regression model
   model <-
     glm(
-      OFI_delay ~ Total_GCS + Gender + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Intubated_prehospitaly + ISS + Age + weekday + work_hours,
+      OFI_delay ~ Total_GCS + Gender + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Intubated + ISS + Age + weekday + work_hours,
       data = factors.data,
       family = binomial)
       
@@ -20,6 +20,7 @@ model.data <- function(dataset){
 }
   
   unadjust2 <- tbl_uvregression(data = factors.data,
+                                exponentiate = TRUE,
                                 method = glm,
                                 y = OFI_delay,
                                 method.args = list(family = binomial),
@@ -52,7 +53,6 @@ fancy_table <-
         Age = "Age (years)",
         Gender = "Gender",
         Total_GCS = "GCS"),
-        pvalue_fun = ~ style_pvalue(.x, digits = 2),
       
       statistic = list(all_continuous() ~ "{mean} ± {sd}"),
       digits = list(all_continuous() ~ c(2, 2))

### 2023-11-09 manuscriptlongversion.Rmd
text
code

@@
 by the multidisciplinary review board and recorded in the trauma care
-quality database. The independent variables are gender, patient survival
-after 30 days from admission, highest hospital care level, GCS on
-arrival, respiratory rate on arrival, systolic blood pressure on
-arrival, time from arrival at the hospital until first computed
-tomography, if the patient was intubated on arrival and whether they
-arrived during weekend or working hours. Those are used as factor
-variables. ISS and age are used as continuous variables.
+quality database. The independent variables are gender, highest hospital
+care level, GCS on arrival, respiratory rate on arrival, systolic blood
+pressure on arrival, if the patient was intubated on arrival, whether
+they arrived during weekend or working hours, ISS and age. Of these
+variables, GCS, respiratory rate, systolic blood pressure, intubation,
+ISS and age are used as continuous variables in the analysis
 
@@
 
-Out of the 11864 patients treated at KS between 2012 and 2022, 1835 were
-included in the study. The patient characteristics were as follows in
-Table 1.
-
-![](table1.png)
-
-Out of the 11,864, 5,282 patients were not screened for ofi, which
-excludes them from the study. Out of the remaining 6,582 patients, 4,775
+Out of the 14,022, 5,710 patients were not screened for ofi, which
+excludes them from the study. Out of the remaining 8,312 patients, 1,823
 patients had missing data in one or more of the analysed parameters,
@@
 
-In the remaining group of 1,835 patients who have been screened for ofi
-and have data registered in all parameters examined for the study, 28
+In the remaining group of 6,489 patients who have been screened for ofi
+and have data registered in all parameters examined for the study, 69
 patients were found to have received delayed treatment during their stay
@@
 ![](Rplot.png)
+
+The 6,489 patients included in the study. 68% of the individuals
+included were men, with the remaining 32% being women. The mean value
+for the ISS score was 10.49, with the standard deviation being 11.4. The
+patients had a mean GCS of ... with a standard deviation of ...
+
+![](table1.png)
 
@@
 p value is also shown.
+
+![](results.png)

### 2023-11-09 functions/create_factors.R

@@ -8,6 +8,10 @@ create.factors <- function(dataset){
   dataset <- dataset %>%
     mutate(ed_gcs_sum = if_else(is.na(ed_gcs_sum) | ed_gcs_sum == 999 | ed_gcs_sum == 99, pre_gcs_sum, ed_gcs_sum))
   
+  dataset <- dataset %>%
+    mutate(ed_gcs_sum = na_if(ed_gcs_sum, 99)) %>%
+    mutate(ed_gcs_sum = na_if(ed_gcs_sum, 999))
+  
   dataset$Total_GCS <- as.numeric(dataset$ed_gcs_sum)
   
   ## Making gender into a factor
@@ -86,7 +90,7 @@ create.factors <- function(dataset){
   
   ##Dates
   dataset[c('Date', 'Time')] <- str_split_fixed(dataset$DateTime_ArrivalAtHospital, ' ', 2)
-  dataset$work_hours <- with(dataset, ifelse(dataset$Time >= "08:00" & dataset$Time <= "16:59", "Yes", "No")) 
+  dataset$work_hours <- with(dataset, ifelse(dataset$Time >= "08:00" & dataset$Time <= "16:59", "Work-hours", "On-call")) 
 
   ##Weekdays
   dataset$Date <- as.Date(dataset$Date)

### 2023-11-09 functions/view_data.R

@@ -46,17 +46,21 @@ fancy_table <-
     
 
 
-  factors.data %>%
-    tbl_summary(  # Create summary statistics for different groups
-      by = OFI_delay,
-      label = list(
-        Age = "Age (years)",
-        Gender = "Gender",
-        Total_GCS = "GCS"),
-      
-      statistic = list(all_continuous() ~ "{mean} ± {sd}"),
-      digits = list(all_continuous() ~ c(2, 2))
-    ) %>%
+factors.data %>%
+  tbl_summary(  
+    by = OFI_delay,
+    label = list(
+      Age = "Age (years)",
+      Gender = "Gender",
+      Total_GCS = "GCS"),
+    statistic = list(all_continuous() ~ "{mean} ± {sd}"),
+    digits = list(all_continuous() ~ c(2, 2))
+  ) %>%
+  add_overall() %>%
+  modify_header(update = list(stat_1 ~ "**Overall**")) %>%
+  modify_spanning_header(c("stat_1", "stat_2"), 
+                         label = "Other Groups", 
+                         spanner = "OFI_delay")
     
 
   

### 2023-11-10 manuscriptlongversion.Rmd
text

@@
 registry and trauma care quality database at the Karolinska University
-Hospital in Solna. The trauma registry includes about 11864 patients
+Hospital in Solna. The trauma registry includes about 14,022 patients
 treated between 2012 and 2022. The trauma care quality database is a
@@
 databases and assess how different patient level factors, such as age,
-sex, mechanism of injury, and injury severity, are associated with
-delayed treatment using logistic regression.
+sex, blood pressure, and injury severity, are associated with delayed
+treatment using logistic regression.
 
@@
 
-Multivariable and potentially univariable logistical regression. A 5%
-significance level and 95% confidence levels will be used.
+Multivariable and univariable logistical regression. A 5% significance
+level and 95% confidence levels will be used.
 
-Rstudio was used to perform the logistical regression. The dependent
-variable is OFI_delay, which is the presence of the ofi delayed
-treatment.
+Rstudio was used to perform the logistical regression. The code was
+written to clean and prepare the data from the database. The data was
+processed to have the variables customized to fit into the logistical
+regression model.
+
+For the missing hospital values in GCS, RR and SBP, values from EMT
+personnel were used
+
+The dependent variable is OFI_delay, which is the presence of the ofi
+delayed treatment.
 
@@
 for the ISS score was 10.49, with the standard deviation being 11.4. The
-patients had a mean GCS of ... with a standard deviation of ...
+patients had a mean GCS of ... with a standard deviation of ... The
+majority of the patients had General care department as their highest
+care level, followed by ICU, ER, OR and lastly specialized care
+department. The mean respiratory rate in the overall group was ..., with
+a standard deviation of ... The mean SBP was ..., having a standard
+deviation of ... The majority of patients, 85%, were not intubated
+during their stay in the hospital, followed by 9.1% intubated at the
+hospital and 6% intubated prehospitaly. The mean age of the patients was
+... years old, with a standard deviation of ... Most of the patients
+arrived during on-call hours,..., with the rest arriving during working
+hours. 67% of the patients arrived on a weekday, with the rest arriving
+on the weekend.
 

### 2023-11-10 functions/create_factors.R

@@ -44,12 +44,20 @@ create.factors <- function(dataset){
   dataset <- dataset %>%
     mutate(ed_rr_value = if_else(is.na(ed_rr_value) | ed_rr_value == 999 | ed_rr_value == 99, pre_rr_value, ed_rr_value))
   
+  dataset <- dataset %>%
+    mutate(ed_rr_value = na_if(ed_rr_value, 99)) %>%
+    mutate(ed_rr_value = na_if(ed_rr_value, 999))
+  
   dataset$Respiratory_rate <- as.numeric(dataset$ed_rr_value)
   
   ## SBP into rts and replacing missing values
   
   dataset <- dataset %>%
     mutate(ed_sbp_value = if_else(is.na(ed_sbp_value) | ed_sbp_value == 999 | ed_sbp_value == 99, pre_sbp_value, ed_sbp_value))
+  
+  dataset <- dataset %>%
+    mutate(ed_sbp_value = na_if(ed_sbp_value, 99)) %>%
+    mutate(ed_sbp_value = na_if(ed_sbp_value, 999))
   
   dataset$Systolic_blood_pressure <- as.numeric(dataset$ed_sbp_value)
   

### 2023-11-13 functions/create_factors.R
code

@@ -3,30 +3,12 @@ library(stringr)
 ## Creating function
 create.factors <- function(dataset){
  
-  ## Making gcs into rts and replacing missing values
-  
-  dataset <- dataset %>%
-    mutate(ed_gcs_sum = if_else(is.na(ed_gcs_sum) | ed_gcs_sum == 999 | ed_gcs_sum == 99, pre_gcs_sum, ed_gcs_sum))
-  
-  dataset <- dataset %>%
-    mutate(ed_gcs_sum = na_if(ed_gcs_sum, 99)) %>%
-    mutate(ed_gcs_sum = na_if(ed_gcs_sum, 999))
-  
-  dataset$Total_GCS <- as.numeric(dataset$ed_gcs_sum)
-  
+
   ## Making gender into a factor
   dataset <- subset(dataset, !is.na(Gender))
   
   
   dataset$Gender <- as.factor(dataset$Gender)
-  
-  
-  ## Survival after 30 days
-  dataset <- subset(dataset, !(is.na(res_survival) | res_survival == 999))
-
-  dataset$res_survival <- ifelse(dataset$res_survival == 1, "Dead", "Alive")
-   
-  dataset$survival_after_30_days <- as.factor(dataset$res_survival)
   
 
   ## Highest hospital care level
@@ -40,27 +22,7 @@ create.factors <- function(dataset){
   
   dataset$Highest_care_level <- as.factor(dataset$host_care_level)
   
-  ## Respiratory rate into rts and replacing missing values
-  dataset <- dataset %>%
-    mutate(ed_rr_value = if_else(is.na(ed_rr_value) | ed_rr_value == 999 | ed_rr_value == 99, pre_rr_value, ed_rr_value))
-  
-  dataset <- dataset %>%
-    mutate(ed_rr_value = na_if(ed_rr_value, 99)) %>%
-    mutate(ed_rr_value = na_if(ed_rr_value, 999))
-  
-  dataset$Respiratory_rate <- as.numeric(dataset$ed_rr_value)
-  
-  ## SBP into rts and replacing missing values
-  
-  dataset <- dataset %>%
-    mutate(ed_sbp_value = if_else(is.na(ed_sbp_value) | ed_sbp_value == 999 | ed_sbp_value == 99, pre_sbp_value, ed_sbp_value))
-  
-  dataset <- dataset %>%
-    mutate(ed_sbp_value = na_if(ed_sbp_value, 99)) %>%
-    mutate(ed_sbp_value = na_if(ed_sbp_value, 999))
-  
[…]

### 2023-11-13 functions/create_factors2.R

@@ -1,0 +1,116 @@
+library(stringr)
+
+## Creating function
+create.factors2 <- function(dataset){
+ 
+
+  ## Making gender into a factor
+  dataset <- subset(dataset, !is.na(Gender))
+  
+  
+  dataset$Gender <- as.factor(dataset$Gender)
+
+  ##Fixing years
+  dataset <- subset(dataset, !(is.na(pt_age_yrs) | pt_age_yrs == 999))
+  
+  dataset$Age <- as.numeric(dataset$pt_age_yrs)
+  
+  ##Fixing ISS
+  dataset <- subset(dataset, !(is.na(ISS) | ISS == 999))
+  
+  dataset$ISS <- as.numeric(dataset$ISS)
+  
+  ##Fixing inital procedure into categories
+   dataset$ed_emerg_proc <- ifelse(dataset$ed_emerg_proc == 1, "Thoracotomy",
+                                    ifelse(dataset$ed_emerg_proc == 2, "Laparotomy – hemostasis",
+                                           ifelse(dataset$ed_emerg_proc == 3, "Packing of the pelvis",
+                                                  ifelse(dataset$ed_emerg_proc == 4, "Revascularization (including surgery for pulseless limb",
+                                                         ifelse(dataset$ed_emerg_proc == 5, "Radiological intervention (Endovascular=embolization, stent, stentgraft)",
+                                                                ifelse(dataset$ed_emerg_proc == 6, "Craniotomy",
+                                                                       ifelse(dataset$ed_emerg_proc == 7, "Intracranial pressure measurement as the only measure",
+                                                                              ifelse(dataset$ed_emerg_proc == 8, "Other action",
+                                                                                     ifelse(dataset$ed_emerg_proc == 99, "No emergency measures performed",
+                                                                                        ifelse(dataset$ed_emerg_proc == 999, "Unknown", NA))))))))))
+   
+   dataset <- subset(dataset, !(is.na(ed_emerg_proc)))
+   
[…]

### 2023-11-13 functions/new_subset.R

@@ -27,7 +27,8 @@ clean.dataset <- function(dataset){
       "DateTime_ArrivalAtHospital",
       "pre_rr_value",
       "pre_sbp_value",
-      "ed_intubated"
+      "ed_intubated",
+      "ed_emerg_proc"
     )]
   
   ## remove rows without OFI

### 2023-11-13 functions/view_data.R

@@ -4,22 +4,22 @@ library(dplyr)
 
 model.data <- function(dataset){
   
-  ## regression model
-  model <-
+  ## regression model nr 1
+  model1 <-
     glm(
-      OFI_delay ~ Total_GCS + Gender + Highest_care_level + Respiratory_rate + Systolic_blood_pressure + Intubated + ISS + Age + weekday + work_hours,
+      OFI_delay ~ Gender + Highest_care_level + Intubated + ISS + Age + weekday + work_hours,
       data = factors.data,
       family = binomial)
       
-      adjusted <-
-        tbl_regression(model, exponentiate = TRUE,
+      adjusted1 <-
+        tbl_regression(model1, exponentiate = TRUE,
                        pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
         bold_p() %>%
         bold_labels()
       add_p(digits = c(p = 3))
 }
   
-  unadjust2 <- tbl_uvregression(data = factors.data,
+  unadjust1 <- tbl_uvregression(data = factors.data1,
                                 exponentiate = TRUE,
                                 method = glm,
                                 y = OFI_delay,
@@ -31,38 +31,82 @@ model.data <- function(dataset){
     bold_labels()
     
 
-fancy_table <-
+fancy_table1 <-
     tbl_merge(
-      tbls        = list(adjusted, unadjust2),
+      tbls        = list(adjusted1, unadjust1),
       tab_spanner = c("Adjusted", "Unadjusted")
     )
   
   
   
   
-  
+## regression model nr 2
+model2 <-
+  glm(
+    OFI_delay ~ Gender + ISS + Age + Emergency_procedure + Systolic_blood_pressure + Respiratory_rate + Total_GCS,
+    data = factors.data2,
+    family = binomial)
+
+adjusted2 <-
+  tbl_regression(model2, exponentiate = TRUE,
+                 pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
+  bold_p() %>%
+  bold_labels()
+add_p(digits = c(p = 3))
+}
+
+unadjust2 <- tbl_uvregression(data = factors.data2,
+                              exponentiate = TRUE,
+                              method = glm,
+                              y = OFI_delay,
+                              method.args = list(family = binomial),
+                              pvalue_fun = ~ style_pvalue(.x, digits = 2),
[…]

### 2023-11-13 main.R

@@ -9,8 +9,11 @@ prepared.data <- prepare.data()
 ##clean data
 cleaned.data <- clean.dataset(prepared.data)
 
-## Create subset
+## Create subset 1
 factors.data <- create.factors(cleaned.data)
+
+## Create subset 2
+factors.data2 <- create.factors2(cleaned.data)
 
 ## view data
 view.data <- model.data(factors.data)

### 2023-11-14 manuscriptlongversion.Rmd
text

@@
 
-The 6,489 patients included in the study. 68% of the individuals
-included were men, with the remaining 32% being women. The mean value
-for the ISS score was 10.49, with the standard deviation being 11.4. The
-patients had a mean GCS of ... with a standard deviation of ... The
-majority of the patients had General care department as their highest
-care level, followed by ICU, ER, OR and lastly specialized care
-department. The mean respiratory rate in the overall group was ..., with
-a standard deviation of ... The mean SBP was ..., having a standard
-deviation of ... The majority of patients, 85%, were not intubated
-during their stay in the hospital, followed by 9.1% intubated at the
-hospital and 6% intubated prehospitaly. The mean age of the patients was
-... years old, with a standard deviation of ... Most of the patients
-arrived during on-call hours,..., with the rest arriving during working
+The 7,051 patients included in the study. 69% of the individuals
+included were men, with the remaining 31% being women. The mean value
+for the ISS score was 11.30, with the standard deviation being 13.22.
+The majority of the patients had General care department as their
+highest care level, followed by ER, ICU, OR and lastly specialized care
+department. The majority of patients, 82%, were not intubated during
+their stay in the hospital, followed by 9.2% intubated at the hospital
+and 8.4% intubated prehospitaly. The mean age of the patients was 44.95
+years old, with a standard deviation of 21.37. Most of the patients
+arrived during on-call hour, 60%, with the rest arriving during working
 hours. 67% of the patients arrived on a weekday, with the rest arriving
@@
 
-![](results.png)
+![](results1.png)
+
+## Discussion
+
+The results show that generally, patients do not systemically receive
+delayed treatment because of a patient level factor.
+
+As we see in the results, there was only one variable that, when
+adjusted, had a correlation with the risk for a patient of receiving
+delayed treatment, being the highest hospital care level. Compared with
+having the ER as the highest care level, patients who were in the ICU or
+the OR had a substantially lowered chance of receiving delayed
+treatment.

### 2023-11-14 manuscriptlongversion.Rmd
text

@@
 output:
-  pdf_document: default
+  word_document: default
   html_document:
     df_print: paged
-  word_document: default
+  pdf_document: default
 bibliography: bibliography.bib
@@
 The results show that generally, patients do not systemically receive
-delayed treatment because of a patient level factor.
+delayed treatment because of a patient level factor. This should be seen
+as a good thing, as it suggests that the delayed treatment happens more
+at random than because of a factor on a patient level.
 

### 2023-11-14 functions/create_factors2.R

@@ -21,16 +21,16 @@ create.factors2 <- function(dataset){
   dataset$ISS <- as.numeric(dataset$ISS)
   
   ##Fixing inital procedure into categories
-   dataset$ed_emerg_proc <- ifelse(dataset$ed_emerg_proc == 1, "Thoracotomy",
+   dataset$ed_emerg_proc <- ifelse(dataset$ed_emerg_proc == 1, NA,
                                     ifelse(dataset$ed_emerg_proc == 2, "Laparotomy – hemostasis",
-                                           ifelse(dataset$ed_emerg_proc == 3, "Packing of the pelvis",
+                                           ifelse(dataset$ed_emerg_proc == 3, NA,
                                                   ifelse(dataset$ed_emerg_proc == 4, "Revascularization (including surgery for pulseless limb",
                                                          ifelse(dataset$ed_emerg_proc == 5, "Radiological intervention (Endovascular=embolization, stent, stentgraft)",
                                                                 ifelse(dataset$ed_emerg_proc == 6, "Craniotomy",
                                                                        ifelse(dataset$ed_emerg_proc == 7, "Intracranial pressure measurement as the only measure",
                                                                               ifelse(dataset$ed_emerg_proc == 8, "Other action",
-                                                                                     ifelse(dataset$ed_emerg_proc == 99, "No emergency measures performed",
-                                                                                        ifelse(dataset$ed_emerg_proc == 999, "Unknown", NA))))))))))
+                                                                                     ifelse(dataset$ed_emerg_proc == 99, "ANo emergency measures performed",
+                                                                                        ifelse(dataset$ed_emerg_proc == 999, NA, NA))))))))))
    
    dataset <- subset(dataset, !(is.na(ed_emerg_proc)))
    

### 2023-11-14 functions/view_data.R

@@ -53,7 +53,7 @@ adjusted2 <-
   bold_p() %>%
   bold_labels()
 add_p(digits = c(p = 3))
-}
+
 
 unadjust2 <- tbl_uvregression(data = factors.data2,
                               exponentiate = TRUE,

### 2023-11-15 manuscriptlongversion.Rmd
text

@@
 The outcome is delayed treatment, as identified by the multidisciplinary
-review board and recorded in the trauma care quality database.
+review board and recorded in the trauma care quality database. The
+outcome we'll be using is "long time untill CT" and "long time until
+operation", which will be referred to as delayed treatment from now on.
 
@@
 
-The outcome is delayed treatment, which is defined as more than 30
-minutes passing from arrival before receiving treatment, as identified
-by the multidisciplinary review board and recorded in the trauma care
-quality database. The independent variables are gender, highest hospital
-care level, GCS on arrival, respiratory rate on arrival, systolic blood
-pressure on arrival, if the patient was intubated on arrival, whether
-they arrived during weekend or working hours, ISS and age. Of these
-variables, GCS, respiratory rate, systolic blood pressure, intubation,
-ISS and age are used as continuous variables in the analysis
+The outcome is delayed treatment as identified by the multidisciplinary
+review board and recorded in the trauma care quality database. The
+independent variables are gender, highest hospital care level, GCS on
+arrival, respiratory rate on arrival, systolic blood pressure on
+arrival, if the patient was intubated on arrival, whether they arrived
+during weekend or working hours, ISS and age. Of these variables, GCS,
+respiratory rate, systolic blood pressure, intubation, ISS and age are
+used as continuous variables in the analysis
 
@@
 patients were screened for OFIs, but starting 2017 all patients that are
-included in the database are screened for OFIs.
+included in the database are screened for OFIs. During this period, 69
+patients were deemed to have received delayed treatment.
 
@@
 processed to have the variables customized to fit into the logistical
-regression model.
+regression model, such as making them into factors, renaming outcomes in
+the database and removing unknown values.
+
+Using the one in ten rule, 7 independent variables were used for the
+logistical regression given that we have 69 patients with delayed
+treatment. The one in ten rule states that for every variable examined,
+10 events are needed in the group. In this case, this means that because
+we have 69 patients with the examined outcome, we can have 7 variables.
+As we had more variables we wanted to use in the regression model, we
+decided to run two logistical regression models. One with gender, age,
+ISS, emergency procedure, SBP, RR and GCS. The other one with gender,
+age, ISS, highest care level, intubation, whether they arrived during
+work hours and whether they arrived on a weekday.
 
@@
 
-The 7,051 patients included in the study. 69% of the individuals
-included were men, with the remaining 31% being women. The mean value
-for the ISS score was 11.30, with the standard deviation being 13.22.
-The majority of the patients had General care department as their
-highest care level, followed by ER, ICU, OR and lastly specialized care
-department. The majority of patients, 82%, were not intubated during
-their stay in the hospital, followed by 9.2% intubated at the hospital
-and 8.4% intubated prehospitaly. The mean age of the patients was 44.95
-years old, with a standard deviation of 21.37. Most of the patients
-arrived during on-call hour, 60%, with the rest arriving during working
-hours. 67% of the patients arrived on a weekday, with the rest arriving
-on the weekend.
+7,051 patients included in the first logistical regression. 69% of the
+individuals included were men, with the remaining 31% being women. The
+mean value for the ISS score was 11.30, with the standard deviation
+being 13.22. The majority of the patients had General care department as
+their highest care level, followed by ER, ICU, OR and lastly specialized
+care department. The majority of patients, 82%, were not intubated
+during their stay in the hospital, followed by 9.2% intubated at the
+hospital and 8.4% intubated prehospitaly. The mean age of the patients
+was 44.95 years old, with a standard deviation of 21.37. Most of the
+patients arrived during on-call hour, 60%, with the rest arriving during
+working hours. 67% of the patients arrived on a weekday, with the rest
+arriving on the weekend.
 
@@
 
-In the following table, the results from the regression analysis is
-displayed.
+In the following table, the results from the first regression analysis
+is displayed.
 
@@
 ![](results1.png)
+
+For the second logistical regression, we had a group size of 7,344
+patients, with 70 of them having received delayed treatment. The reason
+for this group having more patients is because of the exclusion criteria
+that patients can't have data missing in any of the examined variables.
+In this group, fewer patients had missing data, leading to fewer
+exclusions. This resulted in different values in the same variables as
+previously used in the first logistical regression. 68% of the
+individuals included were men, with the remaining 32% being women. The
+mean value for the ISS score was 10.07, with the standard deviation
+being 10.27. The mean age of the patients was 44.20 years old, with a
+standard deviation of 21.41. The majority of the patients had no
+emergency measures performed, with 76% of the patients belonging to this
+group. Following this, the most common emergency procedures were other
+action, laparotomy, craniotomy, radiological intervention, intracranial
+pressure measurement, and lastly, revascularization. The mean SBP was
+137.60, with a standard deviation of 25.09. Mean respiratory rate was
+18.59 with a standard deviation of 5.10. Lastly, mean GCS was 13.95,
+with a standard deviation of 2.58.
+
+![](table2.png)
+
+In the following table, the results from the second regression analysis
+is displayed.
+
+The odds ratio is calculated in the same way as the previous regression
+model, as well as a 95% confidence interval. The p value is also shown.
+
+![](results2.png)
 

### 2023-11-16 manuscriptlongversion.Rmd
text

@@
 
+As the table below shows, most of the variables, when adjusted for, does
+not have an effect on the risk of receiving delayed treatment. Only two
+values in one of the categories had an effect, having ICU or OR compared
+to having ER as your highest care level. It suggests that patients who
+were in the ICU or OR as their highest care level had a lower chance of
+receiving delayed treatment compared with patients who had ER as their
+highest.
+
 ![](results1.png)
@@
 
+The analysis shows that when adjusted for, most variables do not have a
+correlation with receiving delayed treatment.
+
 ![](results2.png)

### 2023-11-17 manuscriptlongversion.Rmd
text

@@
 
-Out of the 14,022, 5,710 patients were not screened for ofi, which
-excludes them from the study. Out of the remaining 8,312 patients, 1,823
-patients had missing data in one or more of the analysed parameters,
-which led to exclusion from the study.
+Out of the 14,022 patients receiving trauma care at Karolinska
+University Hospital, 5,710 patients were not screened for ofi, which
+excludes them from the study and leaves 8,312 patients that have been
+screened for ofi.
 
-In the remaining group of 6,489 patients who have been screened for ofi
-and have data registered in all parameters examined for the study, 69
-patients were found to have received delayed treatment during their stay
-at the hospital.
+As were running two logistical regression models with different
+variables and a missing value in a variable led to exclusion, the two
+groups did not have the same amount of participants. In the first
+logistical regression, 1,261 patients had missing data in one or more of
+the analysed parameters which left 7,051 individuals group, 69 of which
+hade delayed treatment. In the second logistical regression, 1,038
+patients had missing data in one or more of the analysed parameters
+which left 7,274 individuals group, 70 of which hade delayed treatment.
 
@@
 
-7,051 patients included in the first logistical regression. 69% of the
-individuals included were men, with the remaining 31% being women. The
-mean value for the ISS score was 11.30, with the standard deviation
-being 13.22. The majority of the patients had General care department as
-their highest care level, followed by ER, ICU, OR and lastly specialized
-care department. The majority of patients, 82%, were not intubated
-during their stay in the hospital, followed by 9.2% intubated at the
-hospital and 8.4% intubated prehospitaly. The mean age of the patients
-was 44.95 years old, with a standard deviation of 21.37. Most of the
-patients arrived during on-call hour, 60%, with the rest arriving during
-working hours. 67% of the patients arrived on a weekday, with the rest
-arriving on the weekend.
+The 7,051 patients included in the first logistical regression had
+characteristics as shown in table1. 69% of the individuals included were
+men, with the remaining 31% being women. The mean value for the ISS
+score was 11.30, with the standard deviation being 13.22. The majority
+of the patients had General care department as their highest care level,
+followed by ER, ICU, OR and lastly specialized care department. The
+majority of patients, 82%, were not intubated during their stay in the
+hospital, followed by 9.2% intubated at the hospital and 8.4% intubated
+prehospitaly. The mean age of the patients was 44.95 years old, with a
+standard deviation of 21.37. Most of the patients arrived during on-call
+hour, 60%, with the rest arriving during working hours. 67% of the
+patients arrived on a weekday, with the rest arriving on the weekend.
 
@@
 As the table below shows, most of the variables, when adjusted for, does
-not have an effect on the risk of receiving delayed treatment. Only two
-values in one of the categories had an effect, having ICU or OR compared
-to having ER as your highest care level. It suggests that patients who
-were in the ICU or OR as their highest care level had a lower chance of
-receiving delayed treatment compared with patients who had ER as their
-highest.
+not have an effect on the risk of receiving delayed treatment. Only when
+not adjusted for, not being intubated had a correlation with increased
+incidence of delayed treatment, while having higher ISS had the inverse
+effect with lowering the rate of delayed treatment. Only two values in
+one of the categories had an effect, having ICU or OR compared to having
+ER as your highest care level. It suggests that patients who were in the
+ICU or OR as their highest care level had a lower chance of receiving
+delayed treatment compared with patients who had ER as their highest.
 
@@
 
-For the second logistical regression, we had a group size of 7,344
-patients, with 70 of them having received delayed treatment. The reason
-for this group having more patients is because of the exclusion criteria
-that patients can't have data missing in any of the examined variables.
-In this group, fewer patients had missing data, leading to fewer
-exclusions. This resulted in different values in the same variables as
-previously used in the first logistical regression. 68% of the
-individuals included were men, with the remaining 32% being women. The
-mean value for the ISS score was 10.07, with the standard deviation
+Below are the characteristics of the group used in the second logistical
+regression. In this group, fewer patients had missing data, leading to
+fewer exclusions. This resulted in different values in the same
+variables as previously used in the first logistical regression. 68% of
+the individuals included were men, with the remaining 32% being women.
+The mean value for the ISS score was 10.07, with the standard deviation
 being 10.27. The mean age of the patients was 44.20 years old, with a
@@
 The analysis shows that when adjusted for, most variables do not have a
-correlation with receiving delayed treatment.
+correlation with receiving delayed treatment. Three variables, ISS, RR
+and GCS, had a correlation with receiving delayed treatment when not
+adjusted for. However, when adjusted for, the correlation dissapeared.
+In the case of emergency procedure, there was a correlation for all
+emergency procedures performed and having a lowered incidence of delayed
+treatment among the patients, compared with not receiving any emergency
+measures performed.
 

### 2023-11-20 manuscriptlongversion.Rmd
code

@@
 
-As we see in the results, there was only one variable that, when
-adjusted, had a correlation with the risk for a patient of receiving
-delayed treatment, being the highest hospital care level. Compared with
-having the ER as the highest care level, patients who were in the ICU or
-the OR had a substantially lowered chance of receiving delayed
-treatment.
+As we see in the results, there was one variable that, when adjusted,
+had a correlation with the risk for a patient of receiving delayed
+treatment, being the highest hospital care level. Compared with having
+the ER as the highest care level, patients who were in the ICU or the OR
+had a substantially lowered chance of receiving delayed treatment.
+
+The other variable that when adjusted for had a correlation with
+receiving delayed treatment was getting a emergency procedure. Compared
+to not receiving an emergency measure, all observed emergency procedures
+had a correlation with lower chance of delayed treatment.

### 2023-11-20 functions/view_data.R

@@ -1,13 +1,14 @@
 library(gtsummary)
 library(magrittr)
 library(dplyr)
+library(purrr)
 
 model.data <- function(dataset){
   
   ## regression model nr 1
   model1 <-
     glm(
-      OFI_delay ~ Gender + Highest_care_level + Intubated + ISS + Age + weekday + work_hours,
+      OFI_delay ~ Gender + Highest_care_level + Intubated + ISS + Age + Weekday + work_hours,
       data = factors.data,
       family = binomial)
       
@@ -16,10 +17,9 @@ model.data <- function(dataset){
                        pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
         bold_p() %>%
         bold_labels()
-      add_p(digits = c(p = 3))
 }
   
-  unadjust1 <- tbl_uvregression(data = factors.data1,
+  unadjust1 <- tbl_uvregression(data = factors.data,
                                 exponentiate = TRUE,
                                 method = glm,
                                 y = OFI_delay,
@@ -29,6 +29,7 @@ model.data <- function(dataset){
   ) %>%
     bold_p(t = 0.05) %>%
     bold_labels()
+
     
 
 fancy_table1 <-
@@ -36,9 +37,7 @@ fancy_table1 <-
       tbls        = list(adjusted1, unadjust1),
       tab_spanner = c("Adjusted", "Unadjusted")
     )
-  
-  
-  
+
   
 ## regression model nr 2
 model2 <-
@@ -52,7 +51,6 @@ adjusted2 <-
                  pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
   bold_p() %>%
   bold_labels()
-add_p(digits = c(p = 3))
 
 
 unadjust2 <- tbl_uvregression(data = factors.data2,
@@ -66,7 +64,6 @@ unadjust2 <- tbl_uvregression(data = factors.data2,
   bold_p(t = 0.05) %>%
   bold_labels()
 
-
 fancy_table2 <-
   tbl_merge(
     tbls        = list(adjusted2, unadjust2),
@@ -77,6 +74,7 @@ fancy_table2 <-
     
 
 ##table1
+table1 <-
 factors.data %>%
   tbl_summary(  
     by = OFI_delay,
@@ -87,11 +85,15 @@ factors.data %>%
     statistic = list(all_continuous() ~ "{mean} ± {sd}"),
[…]

### 2023-11-21 manuscriptlongversion.Rmd
text

@@
 the body's associated response [@gerdin_risk_2015], is and has long been
-a major cause of death around the world today [@WHO_2021]. Within modern
+a major cause of death around the world [@WHO_2021]. Within modern
 hospital care, the trauma system stands as an important part in
 healthcare and is crucial to lowering mortality and morbidity in injured
-patients through many means, such as hospital care, follow up and
-prevention programs [@Choi_2021].
+patients through many means, such as hospital care, patient follow up
+and prevention programs [@Choi_2021].
 
@@
 related deaths occur in low- and middle-income countries (LMIC), with
-death rates by trauma also being higher than in high income countries.
-Even within these LMIC, people of poorer socio-economic status have
-higher death rates from trauma. Problems identified within these
-countries that contribute to these statisitics were infrastructure,
-education and training, attributed to lack of funding, brain drain to
-HIC and lack of availability of basic amenities [@Shanthakumar_2021]. In
-HIC where funding and governance over the healthcare system is better
-designed, functioning trauma systems and dedicated trauma centers exist
-that have been shown to lower mortality but also improve functional
-outcome in trauma patients [@Nirula_2006].
+death rates by trauma also being higher than in HIC. Even within these
+LMIC, people of poorer socio-economic status have higher death rates
+from trauma. Problems identified within these countries that contribute
+to these statisitics were infrastructure, education and training,
+attributed to lack of funding, brain drain to HIC and lack of
+availability of basic amenities [@Shanthakumar_2021]. In HIC where
+funding and governance over the healthcare system is better designed,
+functioning trauma systems and dedicated trauma centers exist that have
+been shown to lower mortality but also improve functional outcome in
+trauma patients [@Nirula_2006].
 
@@
 trauma centers is associated with a 15-30% decrease in mortality
-[@Moran_2018; \@Celso_2006; @MacKenzie_2006; \@Haas_2012] . During the
+[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012] . During the
 last three decades, the introduction of trauma systems has contributed
@@
 the improvements in care for acute brain injuries and bleeding control.
-The incidence of late death because of sepsis and multiple organ
-failure, possibly a result of better and earlier resuscitation
-[@Asensio_2008]. Functioning trauma systems saves lives in the hospital,
-but their work outside of the hospital is just as important. Data
-collection from injured patients, such as mechanism of injury or
-mortality, are essential for creating databases that can be used for
-research. In turn, that research can be used for planning injury
-prevention programs that target the most common injuries in the right
-ways, e.g. teen drivers, children, specific occupations etc. These
-injury prevention programs can be planned on trauma center,
+The incidence of late death because of sepsis and multiple organ failure
+has also been lowered, possibly a result of better and earlier
+resuscitation [@Asensio_2008]. Functioning trauma systems saves lives in
+the hospital, but their work outside of the hospital is just as
+important. Data collection from injured patients, such as mechanism of
+injury or mortality, are essential for creating databases that can be
+used for research. In turn, that research can be used for planning
+injury prevention programs that target the most common injuries in the
+most efficient ways, e.g. teen drivers, children, specific occupations
+etc. These injury prevention programs can be planned on trauma center,
 organizational or government level[@ACS].
@@
 mobilized, trauma code has to be activated, often by an emergency nurse.
-The nurse uses information gathered by the first responders to see if
+The nurse uses information gathered by the first responders to assess if
 the patient fits any of the criteria for trauma code. Most healthcare
@@
 information and timeline of what happened. The extent of the registry
-varies, with HIC having more complete registries. This logging of
-information is important work, as a cornerstone of trauma quality
-improvement programs is multidisciplinary MoM conferences. The MoM
-conference is a meeting where different specialities who work with
-trauma care sit down and discuss deaths and complications in order to
-look for preventable factors. These conferences are performed in many
-hospitals globally, almost everywhere where there are formal medical
-specialty departments and sometimes in smaller hospitals as well
-[@WHO_2009]. In Sweden, many hospitals have mortality conferences on
-deceased patient cases, but only one has MoM conferences to better their
-trauma care on patients who lived [@SweTrau_2023]. The reason being that
-this expansion of MoM conferences takes time and resources, which is why
-all hospitals don't implement this.
+varies, with HIC having more complete registries. In Sweden, 48 out of
+the 49 hospitals receiving trauma patients are connected to the national
+registry SweTrau, which functions as a national database designed to
+enable scientific processing with epidemiological technology
+[@SweTrau_2021] This logging of information is important work, as a
+cornerstone of trauma quality improvement programs is multidisciplinary
+MoM conferences. The MoM conference is a meeting where different
+specialities who work with trauma care sit down and discuss deaths and
+complications in order to look for preventable factors. These
+conferences are performed in many hospitals globally, almost everywhere
+where there are formal medical specialty departments and sometimes in
+smaller hospitals as well [@WHO_2009]. In Sweden, many hospitals have
+mortality conferences on deceased patient cases, but only one has MoM
+conferences to better their trauma care on patients who lived
+[@SweTrau_2023]. The reason being that this expansion of MoM conferences
+takes time and resources, which is why all hospitals don't implement
+this.
 
@@
 why it is such an important issue and the need to develop strategies to
-combat [@Sampalis_1995; \@Sampalis_1994]. Although it is such a common
+combat [@Sampalis_1995; @Sampalis_1994]. Although it is such a common
 OFI, the patient level factors associated with delayed treatment are
@@
 
-Every year, Karolinska University hospital, which is the equivalent of a
-trauma level 1 hospital, admits 1500 trauma patients each year. To be
-added to the trauma registry, a patient must be over 15 years of age,
-had ISS\>9 and/or had trauma code activation. They get added to both the
-Karolinska Trauma registry as well as the national trauma registry
-(SweTrau). The registry includes data on vital signs, injuries,
-interventions and patient information. Another database, the trauma care
-quality database includes information relevant for MoM conferences. At
-Karolinska University Hospital, patients who die during hospital stay
-gets their case sent to a MoM everytime. However, for the rest of the
-patients, two specialized nurses determine whether they are. Several
-audit filters are set up to flag for patients with possible OFIs.
-Flagged patients are then reviewed by these two nurses who decide if
-there is a possible OFI, in which case their case is sent to a MoM
-conference for further review.
+Karolinska University hospital, which is the equivalent of a trauma
+level 1 hospital, admits 1500 trauma patients each year. To be added to
+the trauma registry, a patient must be over 15 years of age, had ISS\>9
+and/or had trauma code activation. They get added to both the Karolinska
+Trauma registry as well as the national trauma registry (SweTrau). The
+registry includes data on vital signs, injuries, interventions and
+patient information. Another database, the trauma care quality database
+includes information relevant for MoM conferences. At Karolinska
+University Hospital, patients who die during hospital stay gets their
+case sent to a MoM everytime. However, for the rest of the patients, two
+specialized nurses determine whether they are. Several audit filters are
+set up to flag for patients with possible OFIs. Flagged patients are
+then reviewed by these two nurses who decide if there is a possible OFI,
+in which case their case is sent to a MoM conference for further review.
 
@@
 patients were screened for OFIs, but starting 2017 all patients that are
-included in the database are screened for OFIs. During this period, 69
+included in the database are screened for OFIs. During this period, 72
 patients were deemed to have received delayed treatment.
@@
 Using the one in ten rule, 7 independent variables were used for the
-logistical regression given that we have 69 patients with delayed
+logistical regression given that we have 72 patients with delayed
 treatment. The one in ten rule states that for every variable examined,
 10 events are needed in the group. In this case, this means that because
-we have 69 patients with the examined outcome, we can have 7 variables.
+we have 72 patients with the examined outcome, we can have 7 variables.
 As we had more variables we wanted to use in the regression model, we
@@
 the analysed parameters which left 7,051 individuals group, 69 of which
-hade delayed treatment. In the second logistical regression, 1,038
+received delayed treatment. In the second logistical regression, 1,038
 patients had missing data in one or more of the analysed parameters
-which left 7,274 individuals group, 70 of which hade delayed treatment.
+which left 7,274 individuals group, 70 of which received delayed
+treatment.
 
@@
 
+Key results
+
 The results show that generally, patients do not systemically receive
@@
 
-As we see in the results, there was one variable that, when adjusted,
+As we see in the results, there were two variables that, when adjusted,
 had a correlation with the risk for a patient of receiving delayed
-treatment, being the highest hospital care level. Compared with having
-the ER as the highest care level, patients who were in the ICU or the OR
-had a substantially lowered chance of receiving delayed treatment.
+treatment, one being the highest hospital care level. Compared with
+having the ER as the highest care level, patients who were in the ICU or
+the OR had a substantially lowered chance of receiving delayed
+treatment. This can be attributed to
 
@@
 to not receiving an emergency measure, all observed emergency procedures
-had a correlation with lower chance of delayed treatment.
+had a correlation with lower chance of delayed treatment. This can be
+attributed to
+
+Limitations
+
+In this study, one of the limitations was the amount of people having
+the recorded outcome being quite small, leading to having the variables
+being split up into
+
+Discuss limitations of the study, taking into account sources of
+potential bias or imprecision. Discuss both direction and magnitude of
+any potential bias
+
+Interpretation
+
+20
+
+Give a cautious overall interpretation of results considering
+objectives, limitations, multiplicity of analyses, results from similar
+studies, and other relevant evidence
+
+Generalisability
+
+21
+
+Discuss the generalisability (external validity) of the study results

### 2023-11-22 manuscriptlongversion.Rmd
text

@@
 The results show that generally, patients do not systemically receive
-delayed treatment because of a patient level factor. This should be seen
-as a good thing, as it suggests that the delayed treatment happens more
-at random than because of a factor on a patient level.
+delayed treatment because of a patient level factor. In both groups, ISS
+had an unadjusted odds ratio of 0.98 and 0.96 respectively. The
+disparity between these ORs is because of previously mentioned slightly
+different group sizes and patients because of the different patients
+being excluded.
 
@@
 
-20
+The above shown results should be seen as a good thing, as it suggests
+that the delayed treatment happens more at random than because of a
+factor on a patient level.
 
@@
 
+Discussing the generalisability of this study, we included the all the
+trauma patients treated and Karolinska University Hospital and excluded
+40.7% of that population based on that they had not been screened for
+ofi. Additional patients were excluded because of missing data, but no
+distinct groups except the not screened patients were removed. On that
+basis, this population should be a general representation of trauma
+patients at the equivalent of a level 1 trauma center, making this study
+generalisable.
+
 Discuss the generalisability (external validity) of the study results

### 2023-11-23 manuscriptlongversion.Rmd
text

@@
 the recorded outcome being quite small, leading to having the variables
-being split up into
+being split up into two groups for the logistical regression.
+Furthermore, the original plan was to divide RR, SBP and GCS into
+revised trauma score (RTS) categories, from 1-4. However, many of the
+RTS variables had few to no patients with the observed outcome. This led
+to separation in the logistical regression in these variables, referring
+to when an outcome variable separates a predictor variable completely,
+which occurs when the predictor is only associated with one outcome
+value. In the case it was RTS 3 and 4, which left RTS 1 and 2 with
+confidence intervalls of 0.00 to infinity.
 

### 2023-11-24 manuscriptlongversion.Rmd
text

@@
 which occurs when the predictor is only associated with one outcome
-value. In the case it was RTS 3 and 4, which left RTS 1 and 2 with
-confidence intervalls of 0.00 to infinity.
+value [@OARC]. In the case of these variables, it was only RTS 3 and 4
+who had patients with the observed outcome, which left RTS 1 and 2 with
+confidence intervals of 0.00 to infinity. This way would have been the
+best way to examine these variables, but as that wasn't possible with
+this group of patients, we resorted to making these variables
+continuous.
 
-Discuss limitations of the study, taking into account sources of
-potential bias or imprecision. Discuss both direction and magnitude of
-any potential bias
+The biases that might have been present in this study would be selection
+bias and misclassification bias.
 

### 2023-11-28 manuscriptlongversion.Rmd
text

@@
 regression model, such as making them into factors, renaming outcomes in
-the database and removing unknown values.
+the database and removing unknown values. The variables were either
+binary or continuous.
 
@@
 For the missing hospital values in GCS, RR and SBP, values from EMT
-personnel were used
+personnel were used. This was accomplished by identifying missing data
+in the column for the hospital value of the variable and, if available,
+replacing the row in the dataset with the prehospital value. If a
+prehospital value is missing, the patient is excluded from the study.
 
-The dependent variable is OFI_delay, which is the presence of the ofi
-delayed treatment.
+The logistical regression is run through Rstudio using the glm command.
+To make the tables, the package gtsummary was used to visualize the
+adjusted and unadjusted regression, then merging the two tables into
+one.
 
@@
 The biases that might have been present in this study would be selection
-bias and misclassification bias.
+bias and misclassification bias. Our aim is to make this applicable to
+the whole trauma patient
 

### 2023-11-29 manuscriptlongversion.Rmd
text

@@
 
-The 7,051 patients included in the first logistical regression had
-characteristics as shown in table1. 69% of the individuals included were
-men, with the remaining 31% being women. The mean value for the ISS
-score was 11.30, with the standard deviation being 13.22. The majority
-of the patients had General care department as their highest care level,
-followed by ER, ICU, OR and lastly specialized care department. The
-majority of patients, 82%, were not intubated during their stay in the
-hospital, followed by 9.2% intubated at the hospital and 8.4% intubated
-prehospitaly. The mean age of the patients was 44.95 years old, with a
-standard deviation of 21.37. Most of the patients arrived during on-call
-hour, 60%, with the rest arriving during working hours. 67% of the
-patients arrived on a weekday, with the rest arriving on the weekend.
+Below are the characteristics of the group used in the logistical
+regression. Most of the patients were men, 68%, with that being a bit
+higher in the delayed treatment group, 74%. The mean value for the ISS
+score was 10.18, with that value being higher in the outcome group, at
+16.81. The mean age of the patients was 45.18 years old, with it being
+around the same in the delayed treatment group at 45.50. The mean number
+of injuries among the delayed treatment patients was at 6.33, which is
+higher than in the overall group at 4.76. The mean SBP was 137.33, with
+a standard deviation of 25.69. Mean respiratory rate was 18.59 with it
+being slightly higher in the outcome population at 20.39. Lastly, mean
+GCS was 13.93, being lower among patients with delayed treatment at
+12.56.
 
 ![](table1.png)
-
-In the following table, the results from the first regression analysis
-is displayed.
-
-The odds ratio is calculated, as well as a 95% confidence interval. The
-p value is also shown.
-
-As the table below shows, most of the variables, when adjusted for, does
-not have an effect on the risk of receiving delayed treatment. Only when
-not adjusted for, not being intubated had a correlation with increased
-incidence of delayed treatment, while having higher ISS had the inverse
-effect with lowering the rate of delayed treatment. Only two values in
-one of the categories had an effect, having ICU or OR compared to having
-ER as your highest care level. It suggests that patients who were in the
-ICU or OR as their highest care level had a lower chance of receiving
-delayed treatment compared with patients who had ER as their highest.
-
-![](results1.png)
-
-Below are the characteristics of the group used in the second logistical
-regression. In this group, fewer patients had missing data, leading to
-fewer exclusions. This resulted in different values in the same
-variables as previously used in the first logistical regression. 68% of
-the individuals included were men, with the remaining 32% being women.
-The mean value for the ISS score was 10.07, with the standard deviation
-being 10.27. The mean age of the patients was 44.20 years old, with a
-standard deviation of 21.41. The majority of the patients had no
-emergency measures performed, with 76% of the patients belonging to this
-group. Following this, the most common emergency procedures were other
-action, laparotomy, craniotomy, radiological intervention, intracranial
-pressure measurement, and lastly, revascularization. The mean SBP was
-137.60, with a standard deviation of 25.09. Mean respiratory rate was
-18.59 with a standard deviation of 5.10. Lastly, mean GCS was 13.95,
-with a standard deviation of 2.58.
-
-![](table2.png)
 
@@
 treatment among the patients, compared with not receiving any emergency
-measures performed.
+measures performed
 
-![](results2.png)
+![](results1.png)
 

### 2023-11-29 functions/create_factors2.R

@@ -20,21 +20,10 @@ create.factors2 <- function(dataset){
   
   dataset$ISS <- as.numeric(dataset$ISS)
   
-  ##Fixing inital procedure into categories
-   dataset$ed_emerg_proc <- ifelse(dataset$ed_emerg_proc == 1, NA,
-                                    ifelse(dataset$ed_emerg_proc == 2, "Laparotomy – hemostasis",
-                                           ifelse(dataset$ed_emerg_proc == 3, NA,
-                                                  ifelse(dataset$ed_emerg_proc == 4, "Revascularization (including surgery for pulseless limb",
-                                                         ifelse(dataset$ed_emerg_proc == 5, "Radiological intervention (Endovascular=embolization, stent, stentgraft)",
-                                                                ifelse(dataset$ed_emerg_proc == 6, "Craniotomy",
-                                                                       ifelse(dataset$ed_emerg_proc == 7, "Intracranial pressure measurement as the only measure",
-                                                                              ifelse(dataset$ed_emerg_proc == 8, "Other action",
-                                                                                     ifelse(dataset$ed_emerg_proc == 99, "ANo emergency measures performed",
-                                                                                        ifelse(dataset$ed_emerg_proc == 999, NA, NA))))))))))
-   
-   dataset <- subset(dataset, !(is.na(ed_emerg_proc)))
-   
-   dataset$Emergency_procedure <- as.factor(dataset$ed_emerg_proc)
+  ##Number of injuries removing missing values and making it continuous
+  dataset <- subset(dataset, !(is.na(NumberOfInjuries) | NumberOfInjuries == 999))
+  
+  dataset$Number_of_injuries <- as.numeric(dataset$NumberOfInjuries)
   
   ## SBP into rts and replacing missing values
   
@@ -86,27 +75,17 @@ create.factors2 <- function(dataset){
   dataset <-
     dataset[,!(
       names(dataset) %in% c(
-        "res_survival",
-        "host_care_level",
+
+
         "ed_gcs_sum",
         "ed_rr_value",
         "ed_sbp_value",
-        "dt_ed_first_ct",
-        "pre_intubated",
         "pt_age_yrs",
         "pre_gcs_sum",
-        "ed_rr_rtscat",
-        "ed_sbp_rtscat",
-        "DateTime_ArrivalAtHospital",
-        "Date",
-        "Time",
         "pre_rr_value",
         "pre_sbp_value",
         "ofi",
[…]

### 2023-11-29 functions/new_subset.R

@@ -10,25 +10,17 @@ clean.dataset <- function(dataset){
   dataset <-
     dataset[, c(
       "Gender",
-      "res_survival",
-      "host_care_level",
       "ed_gcs_sum",
       "ed_rr_value",
       "ed_sbp_value",
-      "dt_ed_first_ct",
-      "pre_intubated",
       "pt_age_yrs",
       "ISS",
       "ofi",
       "OFI_delay",
       "pre_gcs_sum",
-      "ed_rr_rtscat",
-      "ed_sbp_rtscat",
-      "DateTime_ArrivalAtHospital",
       "pre_rr_value",
       "pre_sbp_value",
-      "ed_intubated",
-      "ed_emerg_proc"
+      "NumberOfInjuries"
     )]
   
   ## remove rows without OFI

### 2023-11-29 functions/view_data.R

@@ -42,13 +42,14 @@ fancy_table1 <-
 ## regression model nr 2
 model2 <-
   glm(
-    OFI_delay ~ Gender + ISS + Age + Emergency_procedure + Systolic_blood_pressure + Respiratory_rate + Total_GCS,
+    OFI_delay ~ Gender + ISS + Age + Number_of_injuries + Systolic_blood_pressure + Respiratory_rate + Total_GCS,
     data = factors.data2,
     family = binomial)
 
 adjusted2 <-
   tbl_regression(model2, exponentiate = TRUE,
                  pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
+  
   bold_p() %>%
   bold_labels()
 
@@ -98,12 +99,11 @@ factors.data2 %>%
   tbl_summary(  
     by = OFI_delay,
     label = list(
+      Number_of_injuries = "Number of injuries",
       Age = "Age (years)",
       Systolic_blood_pressure = "Systolic blood pressure",
-      Emergency_procedure = "Emergency procedure",
       Respiratory_rate = "Respiratory rate",
-      Total_GCS = "GCS",
-      work_hours = "Work hours"),
+      Total_GCS = "GCS"),
     statistic = list(all_continuous() ~ "{mean} ± {sd}"),
     digits = list(all_continuous() ~ c(2, 2))
   ) %>%

### 2023-11-30 manuscriptlongversion.Rmd
text

@@
 review board and recorded in the trauma care quality database. The
-independent variables are gender, highest hospital care level, GCS on
-arrival, respiratory rate on arrival, systolic blood pressure on
-arrival, if the patient was intubated on arrival, whether they arrived
-during weekend or working hours, ISS and age. Of these variables, GCS,
-respiratory rate, systolic blood pressure, intubation, ISS and age are
-used as continuous variables in the analysis
+independent variables are gender, GCS on arrival, respiratory rate on
+arrival, systolic blood pressure on arrival, number of injuries, ISS and
+age. Of these variables, GCS, respiratory rate, systolic blood pressure,
+number of injuries, ISS and age are used as continuous variables in the
+analysis. Gender is used as a categorical variable.
 
@@
 patients were screened for OFIs, but starting 2017 all patients that are
-included in the database are screened for OFIs. During this period, 72
-patients were deemed to have received delayed treatment.
+included in the database are screened for OFIs. Out of the 14,022
+patients in the database, 7,378 were included in the study, with the
+rest being excluded due to either not being screened for ofi or missing
+values in the examined variables. 72 patients were deemed to have
+received delayed treatment. However, only 70 patients with the
+registered outcome were included in the study due to missing values
+among two of the patients.
 
@@
 the database and removing unknown values. The variables were either
-binary or continuous.
+binary or continuous. This code was used on scrambled data at first so
+to avoid selection bias as much as possible. When the code was finished,
+it was run on the unscrambled data.
 
@@
 we have 72 patients with the examined outcome, we can have 7 variables.
-As we had more variables we wanted to use in the regression model, we
-decided to run two logistical regression models. One with gender, age,
-ISS, emergency procedure, SBP, RR and GCS. The other one with gender,
-age, ISS, highest care level, intubation, whether they arrived during
-work hours and whether they arrived on a weekday.
 
@@
 To make the tables, the package gtsummary was used to visualize the
-adjusted and unadjusted regression, then merging the two tables into
-one.
+adjusted and unadjusted regression.
 
@@
 
-As were running two logistical regression models with different
-variables and a missing value in a variable led to exclusion, the two
-groups did not have the same amount of participants. In the first
-logistical regression, 1,261 patients had missing data in one or more of
-the analysed parameters which left 7,051 individuals group, 69 of which
-received delayed treatment. In the second logistical regression, 1,038
-patients had missing data in one or more of the analysed parameters
-which left 7,274 individuals group, 70 of which received delayed
-treatment.
+As we were running the logistical regression, further were excluded from
+the study due to missing data. Among the 8,312 patients screened for
+ofi, 7,378 were included in the study. 934 patients were excluded from
+the study due to missing values in one or more variables.
 
@@
 
-The analysis shows that when adjusted for, most variables do not have a
-correlation with receiving delayed treatment. Three variables, ISS, RR
-and GCS, had a correlation with receiving delayed treatment when not
-adjusted for. However, when adjusted for, the correlation dissapeared.
-In the case of emergency procedure, there was a correlation for all
-emergency procedures performed and having a lowered incidence of delayed
-treatment among the patients, compared with not receiving any emergency
-measures performed
+The analysis shows that some variables seem to have a correlation with
+patients receiving delayed treatment. ISS and respiratory rate both have
+a trend that patient with higher values seem to have a lower incidence
+of delayed treatment, both unadjusted and adjusted. GCS had the opposite
+correlation, with lower GCS often having lower incidence of delayed
+treatment, both unadjusted and adjusted. Number of injuries had the
+correlation of higher number of injuries having lower incidence of
+delayed treatment unadjusted, but that correlation was not seen when
+adjusted for.
 
@@
 
-## Discussion
+# Discussion
 

### 2023-12-01 manuscriptlongversion.Rmd
text

@@
 
-The results show that generally, patients do not systemically receive
-delayed treatment because of a patient level factor. In both groups, ISS
-had an unadjusted odds ratio of 0.98 and 0.96 respectively. The
-disparity between these ORs is because of previously mentioned slightly
-different group sizes and patients because of the different patients
-being excluded.
-
-As we see in the results, there were two variables that, when adjusted,
-had a correlation with the risk for a patient of receiving delayed
-treatment, one being the highest hospital care level. Compared with
-having the ER as the highest care level, patients who were in the ICU or
-the OR had a substantially lowered chance of receiving delayed
-treatment. This can be attributed to
-
-The other variable that when adjusted for had a correlation with
-receiving delayed treatment was getting a emergency procedure. Compared
-to not receiving an emergency measure, all observed emergency procedures
-had a correlation with lower chance of delayed treatment. This can be
-attributed to
+These results show that there seems to be a correlation between some
+variables and the presence of delayed treatment. A higher GCS had a
+correlation with delayed treatment while a lower ISS had the same
+correlation. This suggests that a less injured and more awake patient
+has a higher risk of receiving delayed treatment. Compared to previous a
+previous study analysing the same data but with ofi as the outcome, this
+result is surprising as that study came to the opposite result
+[@Pat_factors]. This can possibly be explained by that the study looked
+at all ofi, whereas this study concentrates on delayed treatment.
+Delayed treatment might have this correlation because an injured and
+unconscious patient might receive more attention at the hospital,
+leading to fewer instances delays compared to a patient that seems less
+injured.
 
@@
 
-The above shown results should be seen as a good thing, as it suggests
-that the delayed treatment happens more at random than because of a
-factor on a patient level.
-
 Give a cautious overall interpretation of results considering

### 2023-12-02 manuscriptlongversion.Rmd
text

@@
 
-## Results
+# Results
 
@@
 In this study, one of the limitations was the amount of people having
-the recorded outcome being quite small, leading to having the variables
-being split up into two groups for the logistical regression.
-Furthermore, the original plan was to divide RR, SBP and GCS into
-revised trauma score (RTS) categories, from 1-4. However, many of the
-RTS variables had few to no patients with the observed outcome. This led
-to separation in the logistical regression in these variables, referring
-to when an outcome variable separates a predictor variable completely,
-which occurs when the predictor is only associated with one outcome
-value [@OARC]. In the case of these variables, it was only RTS 3 and 4
-who had patients with the observed outcome, which left RTS 1 and 2 with
-confidence intervals of 0.00 to infinity. This way would have been the
-best way to examine these variables, but as that wasn't possible with
-this group of patients, we resorted to making these variables
-continuous.
+the recorded outcome being quite small, leading us into having to limit
+the amount of variables we wanted to examine. Furthermore, the original
+plan was to divide RR, SBP and GCS into revised trauma score (RTS)
+categories, from 1-4. However, many of the RTS variables had few to no
+patients with the observed outcome. This led to separation in the
+logistical regression in these variables, referring to when an outcome
+variable separates a predictor variable completely, which occurs when
+the predictor is only associated with one outcome value [@OARC]. In the
+case of these variables, it was only RTS 3 and 4 who had patients with
+the observed outcome, which left RTS 1 and 2 with confidence intervals
+of 0.00 to infinity. This way would have been the best way to examine
+these variables, but as that wasn't possible with this group of
+patients, we resorted to making these variables continuous.
 
 The biases that might have been present in this study would be selection
-bias and misclassification bias. Our aim is to make this applicable to
-the whole trauma patient
+bias and misclassification bias.
 
@@
 patients at the equivalent of a level 1 trauma center, making this study
-generalisable.
+generalisable. This is in line with our aim, as it was to make it
+applicable to the whole trauma patient population
 

### 2023-12-04 manuscriptlongversion.Rmd
text

@@
 
-The outcome is delayed treatment, as identified by the multidisciplinary
-review board and recorded in the trauma care quality database. The
-outcome we'll be using is "long time untill CT" and "long time until
-operation", which will be referred to as delayed treatment from now on.
+The outcome is the ofi delayed treatment. An ofi is any failure in care,
+e.g. clinical judgement error, any preventable or potentially
+preventable death, missed diagnosis, technical error and delayed
+treatment as identified and decided my the MoM conference. The outcomes
+we'll be using is "long time until CT" and "long time until operation",
+which will be referred to as delayed treatment from now on.
 
@@
 
-The outcome is delayed treatment as identified by the multidisciplinary
-review board and recorded in the trauma care quality database. The
-independent variables are gender, GCS on arrival, respiratory rate on
-arrival, systolic blood pressure on arrival, number of injuries, ISS and
-age. Of these variables, GCS, respiratory rate, systolic blood pressure,
-number of injuries, ISS and age are used as continuous variables in the
-analysis. Gender is used as a categorical variable.
+The independent variables are gender, GCS on arrival, respiratory rate
+on arrival, systolic blood pressure on arrival, number of injuries, ISS
+and age. Of these variables, GCS, respiratory rate, systolic blood
+pressure, number of injuries, ISS and age are used as continuous
+variables in the analysis. Gender is used as a categorical variable.
+
+For missing hospital values in GCS, RR and SBP, values from EMT
+personnel were used. This was accomplished by identifying missing data
+in the column for the hospital value of the variable and, if available,
+replacing the row in the dataset with the prehospital value. If a
+prehospital value is missing, the patient is excluded from the study.
 
@@
 regression model, such as making them into factors, renaming outcomes in
-the database and removing unknown values. The variables were either
-binary or continuous. This code was used on scrambled data at first so
-to avoid selection bias as much as possible. When the code was finished,
-it was run on the unscrambled data.
+the database and replacing or removing unknown values. The variables
+were either binary or continuous. This code was used on scrambled data
+at first so to avoid selection bias as much as possible. When the code
+was finished, it was run on the unscrambled data.
 
@@
 we have 72 patients with the examined outcome, we can have 7 variables.
-
-For the missing hospital values in GCS, RR and SBP, values from EMT
-personnel were used. This was accomplished by identifying missing data
-in the column for the hospital value of the variable and, if available,
-replacing the row in the dataset with the prehospital value. If a
-prehospital value is missing, the patient is excluded from the study.
 
@@
 ofi, 7,378 were included in the study. 934 patients were excluded from
-the study due to missing values in one or more variables.
+the study due to missing values in one or more variables. This left us
+with 70 patients in the delayed treatment group and 7,308 patients in
+the no delay to treatment group. The flowchart below illustrates the
+exclusion and inclusion of patients.
 
@@
 
+The problem mentioned above with too few patients in the outcome group
+made this into a post hoc analysis. As we didn't know the amount of
+people who received delayed treatment when we handled the scrambled
+data, we ran the code on the real data when we had too many variables
+for the dataset. Therefore, we had to change the variables after having
+seen the real data, making it a post hoc analysis. This could in theory
+open up for possible selection bias, as we might choose variables which
+give us results we are looking for. However, all the variables included
+in the final analysis were already included in the first one. Variables
+were only removed, not added after having seen the real data.
+
 The biases that might have been present in this study would be selection
-bias and misclassification bias.
+bias and misclassification bias. Selection bias I've explained above but
+misclassification bias is something that we can't control in this study.
+As previously mentioned, the possible existence of an ofi is decided
+during the MoM conferences, where they then add it to the database.
+However, they might not always be right in their judgement and
+misclassify from time to time, leading to patients landing in the wrong
+group in the analysis. This bias should not be big enough to have a
+significant impact on the results, but should be kept in mind.
 
@@
 Generalisability
-
-21
 

### 2023-12-05 manuscriptlongversion.Rmd
text

@@
 
-Key results
-
 These results show that there seems to be a correlation between some
@@
 injured.
+
+Respiratory rate also had the correlation that lower respiratory rate
+correlated with lower rates of delayed treatment among the patients.
+
+Number of injuries was significant when unadjusted, but the correlation
+dissapeared when adjusted for with other variables. Our guess as to why
+this is due to confounding due to it's shared variability with the ISS
+variable. Number of injuries and ISS is often correlated as a more
+severely injured patient will both score higher on ISS and number of
+injuries. Therefore, we hypothesize that the effect of number of
+injuries become insignificant when adjusting for ISS.
 
@@
 in the final analysis were already included in the first one. Variables
-were only removed, not added after having seen the real data.
+were only removed, not added after having seen the real data. The best
+way to have gone about this would have been to just look at the amount
+of people who received delayed treatment before choosing the variables.
+Then, choosing an amount of variables that would work with the amount of
+patients. This would have avoided making it post hoc and changing the
+variables after seeing the data.
 
@@
 significant impact on the results, but should be kept in mind.
+
+Another limitation with this study is that logistical regressions
+assumes that the data is binomial, which the data used is not. The
+analysis assumes a linear correlation between the dependent variable and
+the independent variable, which is best analysed in a normal
+distribution.
 

### 2023-12-06 manuscriptlongversion.Rmd
text

@@
 Respiratory rate also had the correlation that lower respiratory rate
-correlated with lower rates of delayed treatment among the patients.
+correlated with lower rates of delayed treatment among the patients. As
+the interval is \>10 to \<29 in the triage guideline, both too low and
+too high respiratory rate is not desirable[@EHCP2015]. This might have a
+similar explanation to ISS and GCS, that bradypnea in a patient might
+attract more attention in the hospital and lower the incidence of
+delayed treatment. Another possible explanation might be that patients
+with tachypnea might receive higher rates of delayed treatment because
+they are more severely injured, therefore a higher risk of missing a
+possible injury. Another similar study, which focused on ofi generally,
+showed the same correlation, although the significance dissapeared in
+the adjusted analysis.
 
@@
 
-The biases that might have been present in this study would be selection
-bias and misclassification bias. Selection bias I've explained above but
-misclassification bias is something that we can't control in this study.
-As previously mentioned, the possible existence of an ofi is decided
-during the MoM conferences, where they then add it to the database.
-However, they might not always be right in their judgement and
+Further discussing possible biases, misclassification bias might also be
+present in this study, though it is something that we can't control in
+this study. As previously mentioned, the possible existence of an ofi is
+decided during the MoM conferences, where they then add it to the
+database. However, they might not always be right in their judgement and
 misclassify from time to time, leading to patients landing in the wrong
-group in the analysis. This bias should not be big enough to have a
-significant impact on the results, but should be kept in mind.
+group in the analysis. Also, some patients might have been missed even
+before reaching the MoM conference. The patients included in the
+conference are selected through audit filters and two specialized
+nurses, and patients might be missed in this stage. This bias should not
+be big enough to have a significant impact on the results, but should be
+kept in mind.
 
@@
 the independent variable, which is best analysed in a normal
-distribution.
+distribution. However, Fischers test showed an association between 4 of
+the 7 variables, all of them being the ones who showed an increased or
+decreased risk of delayed treatment, showing that these results still
+are significant.
 
@@
 generalisable. This is in line with our aim, as it was to make it
-applicable to the whole trauma patient population
-
-Discuss the generalisability (external validity) of the study results
+applicable to the whole trauma patient population. However, as it is a
+single center study, where it's result only illustrates the situation in
+Stockholm, it is less applicable nationally or globally. This does not
+make it only applicable to the population of Stockholm, but is something
+to keep in mind.

### 2023-12-07 manuscriptlongversion.Rmd
text

@@
 specialized nurses determine whether they are. Several audit filters are
-set up to flag for patients with possible OFIs. Flagged patients are
-then reviewed by these two nurses who decide if there is a possible OFI,
-in which case their case is sent to a MoM conference for further review.
+set up to flag for patients with possible OFIs. Audit filters may
+include low SBP, high RR, EMS scene time greater than 20 mins etc.
+Flagged patients are then reviewed by these two nurses who decide if
+there is a possible OFI, in which case their case is sent to a MoM
+conference for further review.
 
@@
 
-Selection bias as this is a retrospective cohort study where the outcome
-has occurred.
+Selection bias as this is a retrospective, pots hoc cohort study where
+the outcome has occurred.
 
 Misclassification bias as the conference may reach a consensus that is
-wrong.
+wrong, alternatively that the two specialized nurses going through the
+patients flagged for audit filters might missclassify patients or that
+the audit filter misses a patient with ofi.
 
@@
 The logistical regression is run through Rstudio using the glm command.
-To make the tables, the package gtsummary was used to visualize the
-adjusted and unadjusted regression.
+It was ran several times to make an ajusted table and 7 unadjusted, one
+for each variable To make the tables, the package gtsummary was used to
+visualize the adjusted and unadjusted regression.
+
+This process was the only time AI was used in the writing process. It
+was used to make snippets of code for a specific purpose, revise and
+optimize code and find solutions when written code was not functioning
+properly.
 
@@
 
-In the following table, the results from the second regression analysis
-is displayed.
+In the following table, the results from the logistical regression
+analysis is displayed.
 
-The odds ratio is calculated in the same way as the previous regression
-model, as well as a 95% confidence interval. The p value is also shown.
+The odds ratio is calculated both unadjusted and adjusted for other
+variables, as well as a 95% confidence interval. The p value is also
+shown.
 
@@
 a trend that patient with higher values seem to have a lower incidence
-of delayed treatment, both unadjusted and adjusted. GCS had the opposite
+of delayed treatment, both unadjusted (ISS, OR 0.96, 95% CI
+0.95-0.98)(RR, OR 0.95, 95% CI 0.91-0.98) and adjusted (ISS, OR 0.97,
+95% CI 0.95-0.99)(RR, OR 0.96, 95% CI 0.93-0.99). GCS had the opposite
 correlation, with lower GCS often having lower incidence of delayed
-treatment, both unadjusted and adjusted. Number of injuries had the
-correlation of higher number of injuries having lower incidence of
-delayed treatment unadjusted, but that correlation was not seen when
-adjusted for.
+treatment, both unadjusted (OR 1.13, 95% CI 1.07-1.20) and adjusted (OR
+1.08, 95% CI 1.00-1.15). Number of injuries had the correlation of
+higher number of injuries having lower incidence of delayed treatment
+unadjusted (OR 0.94, 95% CI 0.91-0.98), but that correlation was not
+seen when adjusted for. The three remaining variables, age, gender and
+SBP, showed no correlation with the observed outcome.
 
@@
 at all ofi, whereas this study concentrates on delayed treatment.
-Delayed treatment might have this correlation because an injured and
-unconscious patient might receive more attention at the hospital,
-leading to fewer instances delays compared to a patient that seems less
-injured.
+Possible explanations to the correlation include the idea that
+unconsious and more severely injured patients receive more immediate
+attention at the hospital, leading to fewer delays.
 
@@
 showed the same correlation, although the significance dissapeared in
-the adjusted analysis.
+the adjusted analysis [@Pat_factors].
 
@@
 
-Limitations
+The fact that age and gender did not show an increased odds ratio should
+be seen as a positive sign. It suggests that among trauma patients,
+there is no systematical delay based on gender or age.
 
@@
 these variables, but as that wasn't possible with this group of
-patients, we resorted to making these variables continuous.
+patients, we resorted to making these variables continuous. We
+considered expanding the outcome group to include more ofi, but opted
+for the solution used in the study instead as we were solely interested
+in this ofi for this study.
 
@@
 
-Interpretation
+Interpretation and clinical applications
 
-Give a cautious overall interpretation of results considering
-objectives, limitations, multiplicity of analyses, results from similar
-studies, and other relevant evidence
+Further discussing the results and discussing it's clinical
+applications, the results can be interpreted as that it is mostly
+uninjured and awake patients that receive delayed treatment. The
+clinical applications for this are not clear, but might suggest that
+personnel working with trauma patients should be aware of that people
+that seem uninjured might be more likely to receive delayed treatment
+and trying to lower that risk.
 
@@
 to keep in mind.
+
+##

### 2023-12-08 manuscriptlongversion.Rmd
text

@@
 
-##
+In future studies building upon this one, various methods to further
+examine the correlation between delayed treatment and various variables,
+alternatively using another ofi as the outcome variable. Examining
+different variables such as highest hospital care lever, intubation
+status or first emergency procedure performed might be interesting
+variables, among others. To be able to do this, more patients in the
+outcome group would be needed, either by expanding the register to
+several hospitals, or expanding the outcome group by adding different
+ofi. This however would make the study less specific for delayed
+treatment. Further studies might also look into different ofi other than
+delayed treatment, such as
+
+## Conclusion
+
+# Abstract

### 2023-12-11 manuscriptlongversion.Rmd
text

@@
 ---
+
+# Abstract
+
+*Introduction:* Trauma remains a global health challenge, leading to
+millions of deaths each year. The biggest impact can be seen in low to
+middle income countries, but in high income countries, trauma systems
+have been developed to lower the mortality and morbidity associated with
+trauma. To further improve on these systems, conferences are held to
+better the care patients receive and to identify errors made, called
+opportunities for improvement, e.g. delayed treatment. *Aim:* This
+retrospective cohort study aims to identify patient level factors that
+are linked with receiving delayed treatment among trauma patients.
+*Material and Methods:* This study uses the database of Karolinska
+university hospital, the equivalent of a level 1 trauma center, to run a
+logistical regression linking the outcome, delayed treatment, to
+variables, observed patient level factors. Patients treated for trauma
+at the hospital between 2012 and 2022 were included, with ISS\>9 and/or
+trauma code activation. Exclusion criterias were not being screened for
+ofi and missing values in the used variables, age, gender, ISS, SBP, RR,
+GCS and number of injuries. *Results:* Higher ISS and RR was linked to
+higher rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR,
+OR 1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
+treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
+correlation to higher rates of delay, but not when adjusted for. The
+remaining variables showed no correlation. *Conclusion:* Our study
+suggests that it is mostly more severely injured and less awake patients
+that receive delayed treatment.
+
+*Inledning:* Trauma är fortfarande en global hälsoutmaning som skördar
+miljontals liv varje år. Den största effekten kan ses i låg till
+medelinkomstländer, men i höginkomstländer har traumasystem utvecklats
+för att sänka morbiditeten och mortaliteten relaterat till trauma. För
+att ytterliggare förbättra dessa system hålls konferenser för att
+förbättra vården patienter får och identifiera misstag som gjorts, så
+kallade förbättringsmöjligheter, t.ex. försenad vård. *Syfte:* Denna
+retrospektiva kohortstudie syftar till att identifiera patientfaktorer
+kopplade till försenad vård bland traumapatienter. *Material och Metod:*
+Denna studie använder Karolinska universitetssjukhusets databas,
+motsvarande ett traumacenter på nivå 1, för att köra en logistisk
+regression som kopplar utfallet, försenad vård, till variabler,
+observerade patientnivåfaktorer. Patienter som behandlades för trauma på
+sjukhuset mellan 2012 och 2022 inkluderades, med ISS\>9 och/eller
+traumakodaktivering. Exklusionskriterier var om patient inte screenades
+för ofi samt saknade värden i de använda variablerna, ålder, kön, ISS,
+SBT, AF, GCS och antal skador. *Resultat:* Högre ISS och AF var kopplade
+till högre grad av försenad vård (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR
+1.04, 95% CI 1.01-1.08), medan ett lägre GCS också var kopplad till
+försenad vård (OR 0.93, 95% CI 0.87-1.00). Antalet skador visade först
+ett samband med högre förseningsfrekvens, men inte i den justerade
+analysen. De återstående variablerna visade ingen korrelation.
+*Slutsats*: Vår studie tyder på att det mestadels är svårare skadade och
+mindre vakna patienter som får försenad behandling.
 
@@
 wrong, alternatively that the two specialized nurses going through the
-patients flagged for audit filters might missclassify patients or that
+patients flagged for audit filters might misclassify patients or that
 the audit filter misses a patient with ofi.
@@
 patients receiving delayed treatment. ISS and respiratory rate both have
-a trend that patient with higher values seem to have a lower incidence
-of delayed treatment, both unadjusted (ISS, OR 0.96, 95% CI
-0.95-0.98)(RR, OR 0.95, 95% CI 0.91-0.98) and adjusted (ISS, OR 0.97,
-95% CI 0.95-0.99)(RR, OR 0.96, 95% CI 0.93-0.99). GCS had the opposite
-correlation, with lower GCS often having lower incidence of delayed
-treatment, both unadjusted (OR 1.13, 95% CI 1.07-1.20) and adjusted (OR
-1.08, 95% CI 1.00-1.15). Number of injuries had the correlation of
-higher number of injuries having lower incidence of delayed treatment
-unadjusted (OR 0.94, 95% CI 0.91-0.98), but that correlation was not
+a trend that patient with lower values seem to have a lower incidence of
+delayed treatment, both unadjusted (ISS, OR 1.04, 95% CI 1.02-1.05)(RR,
+OR 1.06, 95% CI 1.02-1.09) and adjusted (ISS, OR 1.03, 95% CI
+1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the opposite
+correlation, with higher GCS often having lower incidence of delayed
+treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and adjusted (OR
+0.93, 95% CI 0.87-1.00). Number of injuries had the correlation of
+higher number of injuries having higher incidence of delayed treatment
+unadjusted (OR 1.06, 95% CI 1.02-1.10), but that correlation was not
 seen when adjusted for. The three remaining variables, age, gender and
@@
 These results show that there seems to be a correlation between some
-variables and the presence of delayed treatment. A higher GCS had a
-correlation with delayed treatment while a lower ISS had the same
-correlation. This suggests that a less injured and more awake patient
-has a higher risk of receiving delayed treatment. Compared to previous a
-previous study analysing the same data but with ofi as the outcome, this
-result is surprising as that study came to the opposite result
-[@Pat_factors]. This can possibly be explained by that the study looked
-at all ofi, whereas this study concentrates on delayed treatment.
-Possible explanations to the correlation include the idea that
-unconsious and more severely injured patients receive more immediate
-attention at the hospital, leading to fewer delays.
+variables and the presence of delayed treatment. A lower GCS had a
+correlation with delayed treatment while a higher ISS had the same
+correlation. This suggests that a more heavily injured and less awake
+patient is more prone to receiving delayed treatment. Compared to
+previous a previous study analysing the same data but with ofi as the
+outcome, this result is similar to what they found using the same
+database [@Pat_factors]. As the mentioned study suggests, this could be
+explained by that lower GCS indicates severe trauma, aswell as higher
+ISS. More injured patients often entails a more complex management and
+require more interventions. More interventions leave more room for delay
+to these interventions, why they are more prone to delay in treatment.
 
@@
 the independent variable, which is best analysed in a normal
-distribution. However, Fischers test showed an association between 4 of
-the 7 variables, all of them being the ones who showed an increased or
-decreased risk of delayed treatment, showing that these results still
-are significant.
+distribution. However, Fischers test showed a significant difference
+between the groups in 4 of the 7 variables, all of them being the ones
+who showed an increased or decreased risk of delayed treatment, showing
+that these results still are significant.
 
@@
 applications, the results can be interpreted as that it is mostly
-uninjured and awake patients that receive delayed treatment. The
-clinical applications for this are not clear, but might suggest that
-personnel working with trauma patients should be aware of that people
-that seem uninjured might be more likely to receive delayed treatment
-and trying to lower that risk.
+injured and less awake patients that receive delayed. The clinical
+applications for this are not clear, but caregivers should already be
+aware that the more injuries a patients has, the more room for error
+exists. It is good though to receive a reminder that it's important to
+always work towards minimizing the delay for these injured patients.
 
@@
 treatment. Further studies might also look into different ofi other than
-delayed treatment, such as
+delayed treatment, such as missed injury, faulty communication or
+triage.
 
@@
 
-# Abstract
+In conclusion, based on this small population, the results suggests that
+it is mostly more injured and less awake trauma patients that receive
+delayed treatment. However, due to the limitations of this study, the
+result should only be seen as something to keep in mind when treating
+patients as further studies need to be conducted on this subject.
+
+#

### 2023-12-11 functions/create_factors2.R

@@ -65,11 +65,12 @@ create.factors2 <- function(dataset){
   
   
   ##Replacing NA in OFI_dealy with no delay
-  dataset$OFI_delay[is.na(dataset$OFI_delay)] <- "No delay to treatment"
+  dataset$OFI_delay[is.na(dataset$OFI_delay)] <- "ANo delay to treatment"
   
   dataset$OFI_delay <- as.factor(dataset$OFI_delay)
   
-  dataset$ofi <- as.factor(dataset$ofi)
+
+
  
   ##removing redundant columns
   dataset <-

### 2023-12-11 functions/new_subset.R

@@ -4,7 +4,7 @@ clean.dataset <- function(dataset){
   ## create new colum for outcome to be observed
   dataset$OFI_delay <- as.factor(ifelse(dataset$Problemomrade_.FMP == "Lång tid till DT", "Delay to treatment",
                                                ifelse(dataset$Problemomrade_.FMP == "Lång tid till op", "Delay to treatment",
-                                                      "No delay to treatment")))
+                                                      "ANo delay to treatment")))
   
   ## remove all columns except the analyzed ones
   dataset <-

### 2023-12-11 functions/view_data.R

@@ -2,41 +2,7 @@ library(gtsummary)
 library(magrittr)
 library(dplyr)
 library(purrr)
-
-model.data <- function(dataset){
   
-  ## regression model nr 1
-  model1 <-
-    glm(
-      OFI_delay ~ Gender + Highest_care_level + Intubated + ISS + Age + Weekday + work_hours,
-      data = factors.data,
-      family = binomial)
-      
-      adjusted1 <-
-        tbl_regression(model1, exponentiate = TRUE,
-                       pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
-        bold_p() %>%
-        bold_labels()
-}
-  
-  unadjust1 <- tbl_uvregression(data = factors.data,
-                                exponentiate = TRUE,
-                                method = glm,
-                                y = OFI_delay,
-                                method.args = list(family = binomial),
-                                pvalue_fun = ~ style_pvalue(.x, digits = 2),
-                                hide_n = TRUE
-  ) %>%
-    bold_p(t = 0.05) %>%
-    bold_labels()
-
-    
-
-fancy_table1 <-
-    tbl_merge(
-      tbls        = list(adjusted1, unadjust1),
-      tab_spanner = c("Adjusted", "Unadjusted")
-    )
 
   
 ## regression model nr 2
@@ -67,29 +33,11 @@ unadjust2 <- tbl_uvregression(data = factors.data2,
 
 fancy_table2 <-
   tbl_merge(
-    tbls        = list(adjusted2, unadjust2),
-    tab_spanner = c("Adjusted", "Unadjusted")
+    tbls        = list(unadjust2, adjusted2),
+    tab_spanner = c("Unadjusted", "Adjusted")
   )
   ## view model
 
-    
-
-##table1
-table1 <-
-factors.data %>%
-  tbl_summary(  
-    by = OFI_delay,
-    label = list(
-      Age = "Age (years)",
-      Gender = "Gender",
-      work_hours = "Work hours"),
-    statistic = list(all_continuous() ~ "{mean} ± {sd}"),
-    digits = list(all_continuous() ~ c(2, 2))
-  ) %>%
-  add_p(list(all_continuous() ~ "t.test",
-             all_categorical() ~ "fisher.test")) %>%
-    add_overall("**Overall (N = {N})**") %>%
-   bold_labels()
 
 
 

### 2023-12-12 manuscriptlongversion.Rmd
text

@@
 output:
-  word_document: default
+  pdf_document: default
   html_document:
     df_print: paged
-  pdf_document: default
+  word_document: default
 bibliography: bibliography.bib
@@
 ---
+
+```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
+source("main.R")
+library(rofi)
+library(dplyr)
+library(ggplot2)
+library(gtsummary)
+library(gt)
+library(knitr)
+
+```
 
@@
 
-![](table1.png)
+```{r message = FALSE}
+table1
+```
 
@@
 
-![](results1.png)
+```{r message = FALSE}
+result1
+```
 

### 2023-12-12 Flowchart.R

@@ -5,6 +5,8 @@ library(grid)
 library(magrittr)
 library(gtsummary)
 
+flow.chart <- function(factors.data2){
+
 treated <- boxGrob(glue("Treated at KS between 2012 and 2022",
                         "n = {pop}",
                         pop = txtInt(14022),
@@ -15,16 +17,16 @@ screened <- boxGrob(glue("Screened for ofi",
                          .sep = "\n"))
 included <- boxGrob(glue("Included in study",
                          "n = {incl}",
-                         incl = txtInt(6489),
+                         incl = txtInt(7378),
                          .sep = "\n"))
 delay <- boxGrob(glue("Delay in treatment",
                       "n = {recr}",
-                      recr = txtInt(69),
+                      recr = txtInt(70),
                       .sep = "\n"))
 
 no_delay <- boxGrob(glue("No delay in treatment",
                          "n = {recr}",
-                         recr = txtInt(6420),
+                         recr = txtInt(7308),
                          .sep = "\n"))
 
 not_screened <- boxGrob(glue("Not screened for ofi",
@@ -34,7 +36,7 @@ not_screened <- boxGrob(glue("Not screened for ofi",
 
 missing_data <- boxGrob(glue("Missing data",
                              "n = {recr}",
-                             recr = txtInt(8312-6489),
+                             recr = txtInt(934),
                              .sep = "\n"))
 grid.newpage()
 vert <- spreadVertical(treated = treated,
@@ -69,3 +71,5 @@ grps
 not_screened
 missing_data
 
+
+}

### 2023-12-12 functions/result1.R

@@ -1,0 +1,41 @@
+library(gtsummary)
+library(magrittr)
+library(dplyr)
+library(purrr)
+
+result.1 <- function(factors.data2){
+
+## regression model nr 2
+model1 <-
+  glm(
+    OFI_delay ~ Gender + ISS + Age + Number_of_injuries + Systolic_blood_pressure + Respiratory_rate + Total_GCS,
+    data = factors.data2,
+    family = binomial)
+
+adjusted <-
+  tbl_regression(model1, exponentiate = TRUE,
+                 pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
+  
+  bold_p() %>%
+  bold_labels()
+
+
+unadjust <- tbl_uvregression(data = factors.data2,
+                              exponentiate = TRUE,
+                              method = glm,
+                              y = OFI_delay,
+                              method.args = list(family = binomial),
+                              pvalue_fun = ~ style_pvalue(.x, digits = 2),
+                              hide_n = TRUE
+) %>%
+  bold_p(t = 0.05) %>%
+  bold_labels()
+
+fancy_table <-
+  tbl_merge(
+    tbls        = list(unadjust, adjusted),
+    tab_spanner = c("Unadjusted", "Adjusted")
+  )
+return(fancy_table)
+
+}

### 2023-12-12 functions/view_data.R

@@ -4,7 +4,7 @@ library(dplyr)
 library(purrr)
   
 
-  
+table.1 <- function(factors.data2){
 ## regression model nr 2
 model2 <-
   glm(
@@ -41,8 +41,8 @@ fancy_table2 <-
 
 
 
-##table2
-table2 <-
+##table1
+table1 <-
 factors.data2 %>%
   tbl_summary(  
     by = OFI_delay,
@@ -60,6 +60,7 @@ factors.data2 %>%
 add_overall("**Overall (N = {N})**") %>%
   bold_labels()
   
-  return(model)
+  return(table1)
+}
 
 

### 2023-12-12 main.R

@@ -9,17 +9,14 @@ prepared.data <- prepare.data()
 ##clean data
 cleaned.data <- clean.dataset(prepared.data)
 
-## Create subset 1
-factors.data <- create.factors(cleaned.data)
-
 ## Create subset 2
 factors.data2 <- create.factors2(cleaned.data)
 
-## view data
-view.data <- model.data(factors.data)
+## table1
+table1 <- table.1(factors.data2)
 
-
-print(view.data)
+## result1
+result1 <- result.1(factors.data2)
 
 ## Welcome!
 

### 2023-12-12 manuscriptlongversion.Rmd
text

@@
 Karolinska University hospital, which is the equivalent of a trauma
-level 1 hospital, admits 1500 trauma patients each year. To be added to
-the trauma registry, a patient must be over 15 years of age, had ISS\>9
-and/or had trauma code activation. They get added to both the Karolinska
-Trauma registry as well as the national trauma registry (SweTrau). The
-registry includes data on vital signs, injuries, interventions and
-patient information. Another database, the trauma care quality database
-includes information relevant for MoM conferences. At Karolinska
-University Hospital, patients who die during hospital stay gets their
-case sent to a MoM everytime. However, for the rest of the patients, two
-specialized nurses determine whether they are. Several audit filters are
-set up to flag for patients with possible OFIs. Audit filters may
-include low SBP, high RR, EMS scene time greater than 20 mins etc.
-Flagged patients are then reviewed by these two nurses who decide if
-there is a possible OFI, in which case their case is sent to a MoM
-conference for further review.
+level 1 hospital, admits 1500 trauma patients each year. To have your
+information added to the trauma registry, a patient must be over 15
+years of age, had ISS\>9 and/or had trauma code activation. They get
+added to both the Karolinska Trauma registry as well as the national
+trauma registry (SweTrau). The registry includes data on vital signs,
+injuries, interventions and patient information. Another database, the
+trauma care quality database includes information relevant for MoM
+conferences. At Karolinska University Hospital, patients who die during
+hospital stay gets their case sent to a MoM everytime. However, for the
+rest of the patients, two specialized nurses determine whether they are.
+Several audit filters are set up to flag for patients with possible
+OFIs. Audit filters may include low SBP, high RR, EMS scene time greater
+than 20 mins etc. Flagged patients are then reviewed by these two nurses
+who decide if there is a possible OFI, in which case their case is sent
+to a MoM conference for further review.
 
@@
 subset of the trauma registry and includes about 8000 patients selected
-for review between 2013 and 2022. This project will link the two
-databases and assess how different patient level factors, such as age,
-sex, blood pressure, and injury severity, are associated with delayed
+for review between 2013 and 2022. This project linked the two databases
+and assessed how different patient level factors, such as age, sex,
+blood pressure, and injury severity, are associated with delayed
 treatment using logistic regression.
@@
 
-For missing hospital values in GCS, RR and SBP, values from EMT
-personnel were used. This was accomplished by identifying missing data
-in the column for the hospital value of the variable and, if available,
-replacing the row in the dataset with the prehospital value. If a
-prehospital value is missing, the patient is excluded from the study.
+For missing hospital values in GCS, RR and SBP, values collected from
+EMT personnel before arrival to hospital were used. This was
+accomplished by identifying missing data in the column for the hospital
+value of the variable and, if available, replacing the row in the
+dataset with the prehospital value. If a prehospital value is missing,
+the patient is excluded from the study.
 
@@
 The data comes from the trauma registry and trauma care quality
-database.
+database,
 
@@
 
-Selection bias as this is a retrospective, pots hoc cohort study where
-the outcome has occurred.
+Selection bias as this is a retrospective post hoc cohort study where
+the outcome has occurred before the study began.
 
@@
 Using the one in ten rule, 7 independent variables were used for the
-logistical regression given that we have 72 patients with delayed
+logistical regression given that we have 70 patients with delayed
 treatment. The one in ten rule states that for every variable examined,
 10 events are needed in the group. In this case, this means that because
-we have 72 patients with the examined outcome, we can have 7 variables.
+we have 70 patients with the examined outcome, we can have 7 variables.
 
@@
 It was ran several times to make an ajusted table and 7 unadjusted, one
-for each variable To make the tables, the package gtsummary was used to
-visualize the adjusted and unadjusted regression.
+for each variable. To make the tables, the package gtsummary was used to
+visualize the adjusted and unadjusted regression individually, then
+merging them together into one table.
 
@@
 
-These results show that there seems to be a correlation between some
-variables and the presence of delayed treatment. A lower GCS had a
-correlation with delayed treatment while a higher ISS had the same
-correlation. This suggests that a more heavily injured and less awake
-patient is more prone to receiving delayed treatment. Compared to
-previous a previous study analysing the same data but with ofi as the
-outcome, this result is similar to what they found using the same
-database [@Pat_factors]. As the mentioned study suggests, this could be
-explained by that lower GCS indicates severe trauma, aswell as higher
-ISS. More injured patients often entails a more complex management and
-require more interventions. More interventions leave more room for delay
-to these interventions, why they are more prone to delay in treatment.
+These results show a possible correlation between some variables and the
+presence of delayed treatment. A lower GCS had a correlation with
+delayed treatment while a higher ISS had the same correlation. This
+suggests that a more heavily injured and less awake patient is more
+prone to receiving delayed treatment. Compared to previous a previous
+study analysing the same data but with ofi as the outcome, this result
+is similar to what they found using the same database [@Pat_factors]. As
+the mentioned study suggests, this could be explained by that lower GCS
+indicates severe trauma, aswell as higher ISS. More injured patients
+often entails a more complex management and require more interventions.
+More interventions leave more room for delay to these interventions, why
+they are more prone to delay in treatment.
 
@@
 too high respiratory rate is not desirable[@EHCP2015]. This might have a
-similar explanation to ISS and GCS, that bradypnea in a patient might
-attract more attention in the hospital and lower the incidence of
-delayed treatment. Another possible explanation might be that patients
-with tachypnea might receive higher rates of delayed treatment because
-they are more severely injured, therefore a higher risk of missing a
-possible injury. Another similar study, which focused on ofi generally,
-showed the same correlation, although the significance dissapeared in
-the adjusted analysis [@Pat_factors].
+similar explanation to ISS and GCS, that patients with tachypnea might
+receive higher rates of delayed treatment because they are more severely
+injured, therefore a higher risk of missing a possible injury. Another
+possible explanation might be bradypnea in a patient might attract more
+attention in the hospital and lower the incidence of delayed treatment.
+Another similar study, which focused on ofi generally, showed the same
+correlation, although the significance dissapeared in the adjusted
+analysis [@Pat_factors].
 
@@
 be seen as a positive sign. It suggests that among trauma patients,
-there is no systematical delay based on gender or age.
+there is no systematic delay based on gender or age.
 
@@
 
-Further discussing possible biases, misclassification bias might also be
+On discussing possible biases, misclassification bias might also be
 present in this study, though it is something that we can't control in
@@
 nurses, and patients might be missed in this stage. This bias should not
-be big enough to have a significant impact on the results, but should be
+be big enough to have a significant impact on the results, but can be
 kept in mind.
@@
 
-Interpretation and clinical applications
-
-Further discussing the results and discussing it's clinical
-applications, the results can be interpreted as that it is mostly
-injured and less awake patients that receive delayed. The clinical
-applications for this are not clear, but caregivers should already be
-aware that the more injuries a patients has, the more room for error
-exists. It is good though to receive a reminder that it's important to
-always work towards minimizing the delay for these injured patients.
-
-Generalisability
+Further analysing the results and it's clinical applications, the
+results can be interpreted as that it is mostly injured and less awake
+patients that receive delayed. The clinical applications for this are
+not clear, but caregivers should already be aware that the more injuries
+a patients has, the more room for error exists. It is good though to
+receive a reminder that it's important to always work towards minimizing
+the delay for these injured patients.
 

### 2023-12-12 manuscriptlongversion.Rmd
text

@@
 registry and trauma care quality database at the Karolinska University
-Hospital in Solna. The trauma registry includes about 14,022 patients
-treated between 2012 and 2022. The trauma care quality database is a
-subset of the trauma registry and includes about 8000 patients selected
-for review between 2013 and 2022. This project linked the two databases
-and assessed how different patient level factors, such as age, sex,
-blood pressure, and injury severity, are associated with delayed
-treatment using logistic regression.
+Hospital in Solna. The trauma registry includes 14,022 patients treated
+between 2012 and 2022. The trauma care quality database is a subset of
+the trauma registry and includes about 8000 patients selected for review
+between 2013 and 2022. This project linked the two databases and
+assessed how different patient level factors, such as age, sex, blood
+pressure, and injury severity, are associated with delayed treatment
+using logistic regression.
 
@@
 The database only includes patients 15 years old and above, ISS \>9
-and/or trauma team activation. Inclusion in the study further requires
-that the patients have been assessed for OFIs. Patients were excluded if
+and/or trauma team activation. Inclusion in the study further required
+that the patients has been assessed for OFIs. Patients were excluded if
 they were missing information in one of the examined patient factors.
@@
 
-The fact that age and gender did not show an increased odds ratio should
-be seen as a positive sign. It suggests that among trauma patients,
-there is no systematic delay based on gender or age.
+When looking at the patient demographic, it is clear that the majority
+admitted to trauma care are men, with it being slightly higher in the
+delay to treatment group. The fact that age and gender did not show an
+increased odds ratio should be seen as a positive sign. It suggests that
+among trauma patients, there is no systematic delay based on gender or
+age.
+
+#### Limitations
 
@@
 
+#### Clinical applications
+
 Further analysing the results and it's clinical applications, the
@@
 the delay for these injured patients.
+
+#### Generalisability
 
@@
 to keep in mind.
+
+#### Future studies
 

### 2023-12-12 manuscriptlongversion.Rmd
text

@@
 the mentioned study suggests, this could be explained by that lower GCS
-indicates severe trauma, aswell as higher ISS. More injured patients
+indicates severe trauma, as well as higher ISS. More injured patients
 often entails a more complex management and require more interventions.
 More interventions leave more room for delay to these interventions, why
-they are more prone to delay in treatment.
+they are more prone to delay in treatment. Respiratory rate also had the
+correlation that lower respiratory rate correlated with lower rates of
+delayed treatment among the patients. As the interval is \>10 to \<29 in
+the triage guideline, both too low and too high respiratory rate is not
+desirable[@EHCP2015]. This might have a similar explanation to ISS and
+GCS, that patients with tachypnea might receive higher rates of delayed
+treatment because they are more severely injured, therefore a higher
+risk of missing a possible injury. A previous study also found this, but
+the correlation dissapeared when adjusted for, which was hypothesised
+being due to it's shared variability with the ISS
+variable[@Pat_factors]. In this study, this was the case for number of
+injuries, as it and ISS is often correlated as a more severely injured
+patient will both score higher on ISS and number of injuries. Therefore,
+we hypothesize that the effect of number of injuries become
+insignificant when adjusting for ISS.
 
-Respiratory rate also had the correlation that lower respiratory rate
-correlated with lower rates of delayed treatment among the patients. As
-the interval is \>10 to \<29 in the triage guideline, both too low and
-too high respiratory rate is not desirable[@EHCP2015]. This might have a
-similar explanation to ISS and GCS, that patients with tachypnea might
-receive higher rates of delayed treatment because they are more severely
-injured, therefore a higher risk of missing a possible injury. Another
-possible explanation might be bradypnea in a patient might attract more
-attention in the hospital and lower the incidence of delayed treatment.
-Another similar study, which focused on ofi generally, showed the same
-correlation, although the significance dissapeared in the adjusted
-analysis [@Pat_factors].
-
-Number of injuries was significant when unadjusted, but the correlation
-dissapeared when adjusted for with other variables. Our guess as to why
-this is due to confounding due to it's shared variability with the ISS
-variable. Number of injuries and ISS is often correlated as a more
-severely injured patient will both score higher on ISS and number of
-injuries. Therefore, we hypothesize that the effect of number of
-injuries become insignificant when adjusting for ISS.
+Not many studies regarding trauma and patient level factors have had
+delayed treatment as their outcome, with many analysing mortality. Some
+likeness exist with which patient level factors predict the outcome,
+with higher mortality being seen in patients with lower GCS and higher
+ISS [@Mort_sam][@Mort_hef]. Increased age was also a factor linked to
+mortality, which was not seen to be linked to delayed treatment in this
+study. This
 
@@
 admitted to trauma care are men, with it being slightly higher in the
-delay to treatment group. The fact that age and gender did not show an
-increased odds ratio should be seen as a positive sign. It suggests that
-among trauma patients, there is no systematic delay based on gender or
-age.
+delay to treatment group. However, not high enough to be statistically
+significant. The fact that gender did not show an increased odds ratio
+should be seen as a positive sign. It suggests that among trauma
+patients, there is no systematic delay based on this patient level
+factor. The same goes for age, as
 

### 2023-12-14 flowchart.R
Remove flowchart.R because of weird error

@@ -1,221 +1,0 @@
-################################
-## Data/numbers for flowchart ##
-################################
-## Load functions
-source("functions/functions.R")
-
-## Import data
-datasets <- rofi::import_data()
-
-## Merge data
-combined.datasets <- rofi::merge_data(datasets)
-## SIC(!) Temp fix
-combined.datasets[combined.datasets$tra_id == "92386","tra_DodsfallsanalysGenomford"] <- 1
-## Create OFI column
-combined.datasets$ofi <- create_ofi2(combined.datasets)
-
-#Remove documentation OFI
-combined.datasets$ofi <- ifelse(grepl("Dokumetation|dokumentation|Dokumentation", combined.datasets$Problemomrade_.FMP), "No", combined.datasets$ofi)
-
-dataset.clean.af <- clean_audit_filters(combined.datasets)
-
-## Separate and store cases without known outcome
-missing.outcome <- is.na(dataset.clean.af$ofi)
-n.missing.outcome <- sum(missing.outcome)
-dataset.clean.af <- dataset.clean.af[!missing.outcome, ]
-
-## sep and store cases with age < 15
-age15 <- dataset.clean.af[dataset.clean.af$pt_age_yrs <= 14, ]
-n.under.15 <- nrow(age15)
-
-dataset.comp.age <- dataset.clean.af[dataset.clean.af$pt_age_yrs > 14, ]
-
-## Fix formating and remove wrong values like 999
-clean.dataset <- clean_all_predictors(dataset.comp.age)
-
-n.not.ofi.screened <- sum(is.na(combined.datasets$ofi))
-
-
-df <- subset(clean.dataset, clean.dataset$tra_DodsfallsanalysGenomford == 1)
-n.df <- nrow(df)
-df2 <- subset(clean.dataset, clean.dataset$Fr1.14 == 2 | clean.dataset$Fr1.14 == 3)
-#df3 <- subset(clean.dataset,clean.dataset$tra_id == 77521)
-df2 <- df2[,c("VK_avslutad","tra_DodsfallsanalysGenomford","Problemomrade_.FMP","Fr1.14","tra_id","ofi","Deceased")]
-### Yes and No from mortality conferance
-n.df.ofi <- sum(df$ofi == "Yes", na.rm = TRUE)
-n.df.no.ofi <- sum(df$ofi == "No", na.rm = TRUE)
-
[…]

### 2023-12-14 manuscriptlongversion.Rmd

@@
 
-Rstudio was used to perform the logistical regression. The code was
-written to clean and prepare the data from the database. The data was
-processed to have the variables customized to fit into the logistical
-regression model, such as making them into factors, renaming outcomes in
-the database and replacing or removing unknown values. The variables
-were either binary or continuous. This code was used on scrambled data
-at first so to avoid selection bias as much as possible. When the code
-was finished, it was run on the unscrambled data.
+Rstudio was used to clean and prepare the data from the database. The
+data was processed to have the variables customized to fit into the
+logistical regression model, such as making them into factors, renaming
+outcomes in the database and replacing or removing unknown values. The
+variables were either binary or continuous. While writing the code, it
+was tested on scrambled data so to avoid selection bias as much as
+possible. When the code was finished, it was run on the unscrambled data
+for a result.
 
@@
 The logistical regression is run through Rstudio using the glm command.
-It was ran several times to make an ajusted table and 7 unadjusted, one
+It was ran several times to make an adjusted table and 7 unadjusted, one
 for each variable. To make the tables, the package gtsummary was used to
@@
 mortality, which was not seen to be linked to delayed treatment in this
-study. This
+study. An older patient is more likely to have co-morbidities that might
+affect their mortality but not increase their likelihood of receiving
+delayed treatment the same way ISS and GCS does[@Gioffre2018].
 
@@
 patients, there is no systematic delay based on this patient level
-factor. The same goes for age, as
+factor.
 

### 2023-12-14 manuscriptlongversion.Rmd
Comment on manuscript

@@
 library(knitr)
-
 ```
@@
 better the care patients receive and to identify errors made, called
+<!-- Consider clarifying that delayed treatment may be one of the most 
+common OFI-->
 opportunities for improvement, e.g. delayed treatment. *Aim:* This
 retrospective cohort study aims to identify patient level factors that
+<!-- I would use "associated" rather than "linked"-->
 are linked with receiving delayed treatment among trauma patients.
+<!-- The first sentence is a bit complicated. Try to simplify, break 
+into multiple sentences -->
 *Material and Methods:* This study uses the database of Karolinska
@@
 trauma code activation. Exclusion criterias were not being screened for
-ofi and missing values in the used variables, age, gender, ISS, SBP, RR,
+OFI and missing values in any of the variables, age, gender, ISS, SBP, RR,
 GCS and number of injuries. *Results:* Higher ISS and RR was linked to
@@
 the body's associated response [@gerdin_risk_2015], is and has long been
-a major cause of death around the world [@WHO_2021]. Within modern
+a major cause of death around the world [@WHO_2021]. 
+<!-- The sentence below is long and overly complicated. Are you saying 
+that trauma systems is an important part of healthcare and its implementation
+has reduced mortality and morbidity?-->
+Within modern
 hospital care, the trauma system stands as an important part in
@@
 ## Trauma Statistics
-
+<!-- I suggest you name this section something else, like "Trauma globally" 
+or something like that. Also, divide the second paragraph into two or 
+three paragraphs-->
 Trauma is often divided into two subgroups, blunt force trauma and
@@
 ## Trauma systems
-
+<!-- Split long paragraphs -->
 Trauma systems are infrastructures that exist to provide and optimize
@@
 trauma centers is associated with a 15-30% decrease in mortality
-[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012] . During the
+[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. During the
 last three decades, the introduction of trauma systems has contributed
-to lowering the incidence of preventable death . This is attributed to
+to lowering the incidence of preventable death. This is attributed to
 the improvements in care for acute brain injuries and bleeding control.
@@
 nursing and support staff[@västra_G]. For the trauma team to be
-mobilized, trauma code has to be activated, often by an emergency nurse.
+mobilized, a trauma code has to be activated, often by an emergency nurse.
 The nurse uses information gathered by the first responders to assess if
@@
 ## Mortality and morbidity (MoM) conferences
-
+<!-- Consider renaming this section as "Opportunities for improvement" -->
 Many hospitals have a trauma registry where they log the patients
@@
 ## Aim
-
+<!-- Make sure the aim here is exactly the same as in the abstract. Also,
+I suggest that your aim is to determine the associations between different 
+patient level factors and receiving delayed treatment -->
 The aim of this paper is to determine what different patient level
@@
 
-Karolinska University hospital, which is the equivalent of a trauma
-level 1 hospital, admits 1500 trauma patients each year. To have your
-information added to the trauma registry, a patient must be over 15
+The Karolinska University hospital, which is the equivalent of a trauma
+level 1 hospital, admits 1500 trauma patients each year. To be included in
+ the trauma registry, a patient must be over 15
 years of age, had ISS\>9 and/or had trauma code activation. They get
@@
 
-The outcome is the ofi delayed treatment. An ofi is any failure in care,
+The outcome is the OFI delayed treatment. An OFI is any failure in care,
 e.g. clinical judgement error, any preventable or potentially
@@
 treatment as identified and decided my the MoM conference. The outcomes
-we'll be using is "long time until CT" and "long time until operation",
+we used were "long time until CT" and "long time until operation",
 which will be referred to as delayed treatment from now on.
@@
 The data comes from the trauma registry and trauma care quality
-database,
+database.
 
@@
 
-This process was the only time AI was used in the writing process. It
+This process was the only time generative AI was used in the writing process. It
 was used to make snippets of code for a specific purpose, revise and
@@
 
-As we were running the logistical regression, further were excluded from
+As we were running the logistic regression, additional patients were excluded from
 the study due to missing data. Among the 8,312 patients screened for
@@
 with 70 patients in the delayed treatment group and 7,308 patients in
-the no delay to treatment group. The flowchart below illustrates the
+the no delay to treatment group. Figure 1 below illustrates the
 exclusion and inclusion of patients.
@@
 
-Below are the characteristics of the group used in the logistical
+<!-- You need to refer to tables and figures by numbers in the text, 
+i.e. Figure 1, Table 1 and so on -->
+Table 1 shows the characteristics of the group used in the logistic
 regression. Most of the patients were men, 68%, with that being a bit
@@
 
-In the following table, the results from the logistical regression
+In the following table<!-- See comments above-->, the results from the logistical regression
 analysis is displayed.
@@
 The odds ratio is calculated both unadjusted and adjusted for other
-variables, as well as a 95% confidence interval. The p value is also
-shown.
+variables<!-- What does "other" mean here?-->, as well as a 95% confidence interval. 
+The p value is also shown.
 
+<!-- Use "associated" instead of correlated or linked etc-->
 The analysis shows that some variables seem to have a correlation with
-patients receiving delayed treatment. ISS and respiratory rate both have
+patients receiving delayed treatment. ISS and respiratory rate (RR) both have
 a trend that patient with lower values seem to have a lower incidence of
-delayed treatment, both unadjusted (ISS, OR 1.04, 95% CI 1.02-1.05)(RR,
-OR 1.06, 95% CI 1.02-1.09) and adjusted (ISS, OR 1.03, 95% CI
+delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI 1.02-1.05)(RR,
+OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR 1.03, 95% CI
 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the opposite
@@
 suggests that a more heavily injured and less awake patient is more
-prone to receiving delayed treatment. Compared to previous a previous
-study analysing the same data but with ofi as the outcome, this result
+prone to receiving delayed treatment. Compared to a previous
+study analysing the same data but with OFI as the outcome, this result
 is similar to what they found using the same database [@Pat_factors]. As
-the mentioned study suggests, this could be explained by that lower GCS
-indicates severe trauma, as well as higher ISS. More injured patients
-often entails a more complex management and require more interventions.
+the mentioned study suggests, this could be explained by that lower GCS as well as higher ISS
+indicates severe trauma. More injured patients
+often requires a more complex management and more interventions.
 More interventions leave more room for delay to these interventions, why
@@
 treatment because they are more severely injured, therefore a higher
-risk of missing a possible injury. A previous study also found this, but
+risk of missing a possible injury. The previous study also found this, but
 the correlation dissapeared when adjusted for, which was hypothesised
-being due to it's shared variability with the ISS
+being due to its shared variability with the ISS
 variable[@Pat_factors]. In this study, this was the case for number of
@@
 Not many studies regarding trauma and patient level factors have had
-delayed treatment as their outcome, with many analysing mortality. Some
-likeness exist with which patient level factors predict the outcome,
+delayed treatment as their outcome, while many analysing mortality. Some
+similarities exist with which patient level factors predict the outcome,
 with higher mortality being seen in patients with lower GCS and higher
-ISS [@Mort_sam][@Mort_hef]. Increased age was also a factor linked to
+ISS [@Mort_sam][@Mort_hef]<!-- Are these two references?-->. Increased age was also a factor linked to
 mortality, which was not seen to be linked to delayed treatment in this
@@
 
-When looking at the patient demographic, it is clear that the majority
+When looking at the patient demographics, it is clear that the majority
 admitted to trauma care are men, with it being slightly higher in the
@@
 
-In this study, one of the limitations was the amount of people having
+In this study, one of the limitations was the number of people having
 the recorded outcome being quite small, leading us into having to limit
-the amount of variables we wanted to examine. Furthermore, the original
+the number of variables we wanted to examine. Furthermore, the original
 plan was to divide RR, SBP and GCS into revised trauma score (RTS)
@@
 patients with the observed outcome. This led to separation in the
-logistical regression in these variables, referring to when an outcome
+logistic regression in these variables, referring to when an outcome
 variable separates a predictor variable completely, which occurs when
@@
 of 0.00 to infinity. This way would have been the best way to examine
-these variables, but as that wasn't possible with this group of
+these variables, but as that was not possible with this group of
 patients, we resorted to making these variables continuous. We
-considered expanding the outcome group to include more ofi, but opted
+considered expanding the outcome group to include more OFI, but opted
 for the solution used in the study instead as we were solely interested
-in this ofi for this study.
+in this OFI for this study.
 
 The problem mentioned above with too few patients in the outcome group
-made this into a post hoc analysis. As we didn't know the amount of
+made this into a post hoc analysis. As we did not know the number of
 people who received delayed treatment when we handled the scrambled
@@
 On discussing possible biases, misclassification bias might also be
-present in this study, though it is something that we can't control in
-this study. As previously mentioned, the possible existence of an ofi is
+<!-- Do not use contractions, i.e. can't, doesn't etc -->
+present in this study, though it is something that we cannot control in
+this study. As previously mentioned, the possible existence of an OFI is
 decided during the MoM conferences, where they then add it to the
@@
 
-Another limitation with this study is that logistical regressions
+<!-- Not so sure about this section. What do you mean that the logistic
+regression assumes that the data is binomial? Do you mean that it requires 
+a binary outcome? I think you discuss potential limitations well enough in 
+the previous paragraphs, so you don't really need this -->
+Another limitation with this study is that logistic regressions
 assumes that the data is binomial, which the data used is not. The
@@
 #### Clinical applications
+
+<!-- Revise this section now after our meeting -->
 
@@
 applicable to the whole trauma patient population. However, as it is a
-single center study, where it's result only illustrates the situation in
+single center study, where its result only illustrates the situation in
 Stockholm, it is less applicable nationally or globally. This does not
@@
 to keep in mind.
+
+<!-- Think about what components of your study is generalisible and to 
+where. I think that your results are generalisible to  a large extent, in 
+the sense that severly injured patients are at higher risk of delayed 
+treatment, also in other contexts -->
 
@@
 examine the correlation between delayed treatment and various variables,
-alternatively using another ofi as the outcome variable. Examining
+alternatively using another OFI as the outcome variable. Examining
 different variables such as highest hospital care lever, intubation
@@
 In conclusion, based on this small population, the results suggests that
+<!-- I would say "with affected consciousness" instead of "less awake" -->
 it is mostly more injured and less awake trauma patients that receive

### 2023-12-14 manuscriptlongversion.Rmd
text

@@
 better the care patients receive and to identify errors made, called
-<!-- Consider clarifying that delayed treatment may be one of the most 
-common OFI-->
-opportunities for improvement, e.g. delayed treatment. *Aim:* This
-retrospective cohort study aims to identify patient level factors that
-<!-- I would use "associated" rather than "linked"-->
-are linked with receiving delayed treatment among trauma patients.
+opportunities for improvement, one of the more common being delayed
+treatment. *Aim:* This retrospective cohort study aims to identify
+patient level factors that are associated with receiving delayed
+treatment among trauma patients.
 <!-- The first sentence is a bit complicated. Try to simplify, break 
-into multiple sentences -->
-*Material and Methods:* This study uses the database of Karolinska
-university hospital, the equivalent of a level 1 trauma center, to run a
-logistical regression linking the outcome, delayed treatment, to
-variables, observed patient level factors. Patients treated for trauma
-at the hospital between 2012 and 2022 were included, with ISS\>9 and/or
-trauma code activation. Exclusion criterias were not being screened for
-OFI and missing values in any of the variables, age, gender, ISS, SBP, RR,
-GCS and number of injuries. *Results:* Higher ISS and RR was linked to
-higher rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR,
-OR 1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
-treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
-correlation to higher rates of delay, but not when adjusted for. The
-remaining variables showed no correlation. *Conclusion:* Our study
-suggests that it is mostly more severely injured and less awake patients
-that receive delayed treatment.
+into multiple sentences --> *Material and Methods:* This study uses the
+database of Karolinska university hospital, the equivalent of a level 1
+trauma center, to run a logistical regression linking the outcome,
+delayed treatment, to variables, observed patient level factors.
+Patients treated for trauma at the hospital between 2012 and 2022 were
+included, with ISS\>9 and/or trauma code activation. Exclusion criterias
+were not being screened for OFI and missing values in any of the
+variables, age, gender, ISS, SBP, RR, GCS and number of injuries.
+*Results:* Higher ISS and RR was linked to higher rates of delay in
+treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI
+1.01-1.08), while a lower GCS also was linked to delayed treatment (OR
+0.93, 95% CI 0.87-1.00). Number of injuries showed a correlation to
+higher rates of delay, but not when adjusted for. The remaining
+variables showed no correlation. *Conclusion:* Our study suggests that
+it is mostly more severely injured and less awake patients that receive
+delayed treatment.
 
@@
 förbättra vården patienter får och identifiera misstag som gjorts, så
-kallade förbättringsmöjligheter, t.ex. försenad vård. *Syfte:* Denna
-retrospektiva kohortstudie syftar till att identifiera patientfaktorer
-kopplade till försenad vård bland traumapatienter. *Material och Metod:*
-Denna studie använder Karolinska universitetssjukhusets databas,
-motsvarande ett traumacenter på nivå 1, för att köra en logistisk
-regression som kopplar utfallet, försenad vård, till variabler,
-observerade patientnivåfaktorer. Patienter som behandlades för trauma på
-sjukhuset mellan 2012 och 2022 inkluderades, med ISS\>9 och/eller
-traumakodaktivering. Exklusionskriterier var om patient inte screenades
-för ofi samt saknade värden i de använda variablerna, ålder, kön, ISS,
-SBT, AF, GCS och antal skador. *Resultat:* Högre ISS och AF var kopplade
-till högre grad av försenad vård (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR
-1.04, 95% CI 1.01-1.08), medan ett lägre GCS också var kopplad till
-försenad vård (OR 0.93, 95% CI 0.87-1.00). Antalet skador visade först
-ett samband med högre förseningsfrekvens, men inte i den justerade
-analysen. De återstående variablerna visade ingen korrelation.
-*Slutsats*: Vår studie tyder på att det mestadels är svårare skadade och
-mindre vakna patienter som får försenad behandling.
+kallade förbättringsmöjligheter, där försenad vård är en av de
+vanligaste. *Syfte:* Denna retrospektiva kohortstudie syftar till att
+identifiera patientfaktorer kopplade till försenad vård bland
+traumapatienter. *Material och Metod:* Denna studie använder Karolinska
+universitetssjukhusets databas, motsvarande ett traumacenter på nivå 1,
+för att köra en logistisk regression som kopplar utfallet, försenad
+vård, till variabler, observerade patientnivåfaktorer. Patienter som
+behandlades för trauma på sjukhuset mellan 2012 och 2022 inkluderades,
+med ISS\>9 och/eller traumakodaktivering. Exklusionskriterier var om
+patient inte screenades för ofi samt saknade värden i de använda
+variablerna, ålder, kön, ISS, SBT, AF, GCS och antal skador. *Resultat:*
+Högre ISS och AF var kopplade till högre grad av försenad vård (ISS, OR
+1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08), medan ett lägre
+GCS också var kopplad till försenad vård (OR 0.93, 95% CI 0.87-1.00).
+Antalet skador visade först ett samband med högre förseningsfrekvens,
+men inte i den justerade analysen. De återstående variablerna visade
+ingen korrelation. *Slutsats*: Vår studie tyder på att det mestadels är
+svårare skadade och mindre vakna patienter som får försenad behandling.
 
@@
 that trauma systems is an important part of healthcare and its implementation
-has reduced mortality and morbidity?-->
-Within modern
-hospital care, the trauma system stands as an important part in
-healthcare and is crucial to lowering mortality and morbidity in injured
-patients through many means, such as hospital care, patient follow up
-and prevention programs [@Choi_2021].
+has reduced mortality and morbidity?--> Within modern hospital care, the
+trauma system stands as an important part in healthcare and is crucial
+to lowering mortality and morbidity in injured patients through many
+means, such as hospital care, patient follow up and prevention programs
+[@Choi_2021].
 
-## Trauma Statistics
+## Trauma globally
+
+```{=html}
 <!-- I suggest you name this section something else, like "Trauma globally" 
@@
 three paragraphs-->
+```
 Trauma is often divided into two subgroups, blunt force trauma and
@@
 ## Trauma systems
+
 <!-- Split long paragraphs -->
+
 Trauma systems are infrastructures that exist to provide and optimize
@@
 trauma centers is associated with a 15-30% decrease in mortality
-[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. During the
-last three decades, the introduction of trauma systems has contributed
-to lowering the incidence of preventable death. This is attributed to
-the improvements in care for acute brain injuries and bleeding control.
-The incidence of late death because of sepsis and multiple organ failure
-has also been lowered, possibly a result of better and earlier
-resuscitation [@Asensio_2008]. Functioning trauma systems saves lives in
-the hospital, but their work outside of the hospital is just as
-important. Data collection from injured patients, such as mechanism of
-injury or mortality, are essential for creating databases that can be
-used for research. In turn, that research can be used for planning
-injury prevention programs that target the most common injuries in the
-most efficient ways, e.g. teen drivers, children, specific occupations
-etc. These injury prevention programs can be planned on trauma center,
+[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. During the last
+three decades, the introduction of trauma systems has contributed to
+lowering the incidence of preventable death. This is attributed to the
+improvements in care for acute brain injuries and bleeding control. The
+incidence of late death because of sepsis and multiple organ failure has
+also been lowered, possibly a result of better and earlier resuscitation
+[@Asensio_2008]. Functioning trauma systems saves lives in the hospital,
+but their work outside of the hospital is just as important. Data
+collection from injured patients, such as mechanism of injury or
+mortality, are essential for creating databases that can be used for
+research. In turn, that research can be used for planning injury
+prevention programs that target the most common injuries in the most
+efficient ways, e.g. teen drivers, children, specific occupations etc.
+These injury prevention programs can be planned on trauma center,
 organizational or government level[@ACS].
@@
 nursing and support staff[@västra_G]. For the trauma team to be
-mobilized, a trauma code has to be activated, often by an emergency nurse.
-The nurse uses information gathered by the first responders to assess if
-the patient fits any of the criteria for trauma code. Most healthcare
-facilities have established criteria or guidelines that trigger trauma
-code activation. These criteria typically include mechanisms of injury,
-physiological criteria, anatomical criteria and other specific
-indicators of severe trauma. There are different levels of trauma code
-with different criteria correlating with severity of injury, with level
-1 mobilizing the most personnel to the trauma room. Once in the trauma
-room, the team works systematically to manage the patients injuries.
-They handle the most urgent problems first, such as airways and
-breathing, with the aim to rapidly assess and stabilise the patient,
-prioritise their injuries and arrange for site of definitive care.
-[@Hedberg_2020]
+mobilized, a trauma code has to be activated, often by an emergency
+nurse. The nurse uses information gathered by the first responders to
+assess if the patient fits any of the criteria for trauma code. Most
+healthcare facilities have established criteria or guidelines that
+trigger trauma code activation. These criteria typically include
+mechanisms of injury, physiological criteria, anatomical criteria and
+other specific indicators of severe trauma. There are different levels
+of trauma code with different criteria correlating with severity of
+injury, with level 1 mobilizing the most personnel to the trauma room.
+Once in the trauma room, the team works systematically to manage the
+patients injuries. They handle the most urgent problems first, such as
+airways and breathing, with the aim to rapidly assess and stabilise the
+patient, prioritise their injuries and arrange for site of definitive
+care. [@Hedberg_2020]
 
 ## Mortality and morbidity (MoM) conferences
+
 <!-- Consider renaming this section as "Opportunities for improvement" -->
+
 Many hospitals have a trauma registry where they log the patients
@@
 ## Aim
+
+```{=html}
 <!-- Make sure the aim here is exactly the same as in the abstract. Also,
@@
 patient level factors and receiving delayed treatment -->
+```
 The aim of this paper is to determine what different patient level
@@
 The Karolinska University hospital, which is the equivalent of a trauma
-level 1 hospital, admits 1500 trauma patients each year. To be included in
- the trauma registry, a patient must be over 15
-years of age, had ISS\>9 and/or had trauma code activation. They get
-added to both the Karolinska Trauma registry as well as the national
-trauma registry (SweTrau). The registry includes data on vital signs,
-injuries, interventions and patient information. Another database, the
-trauma care quality database includes information relevant for MoM
-conferences. At Karolinska University Hospital, patients who die during
-hospital stay gets their case sent to a MoM everytime. However, for the
-rest of the patients, two specialized nurses determine whether they are.
-Several audit filters are set up to flag for patients with possible
-OFIs. Audit filters may include low SBP, high RR, EMS scene time greater
-than 20 mins etc. Flagged patients are then reviewed by these two nurses
-who decide if there is a possible OFI, in which case their case is sent
-to a MoM conference for further review.
+level 1 hospital, admits 1500 trauma patients each year. To be included
+in the trauma registry, a patient must be over 15 years of age, had
+ISS\>9 and/or had trauma code activation. They get added to both the
+Karolinska Trauma registry as well as the national trauma registry
+(SweTrau). The registry includes data on vital signs, injuries,
+interventions and patient information. Another database, the trauma care
+quality database includes information relevant for MoM conferences. At
+Karolinska University Hospital, patients who die during hospital stay
+gets their case sent to a MoM everytime. However, for the rest of the
+patients, two specialized nurses determine whether they are. Several
+audit filters are set up to flag for patients with possible OFIs. Audit
+filters may include low SBP, high RR, EMS scene time greater than 20
+mins etc. Flagged patients are then reviewed by these two nurses who
+decide if there is a possible OFI, in which case their case is sent to a
+MoM conference for further review.
 
@@
 treatment as identified and decided my the MoM conference. The outcomes
-we used were "long time until CT" and "long time until operation",
-which will be referred to as delayed treatment from now on.
+we used were "long time until CT" and "long time until operation", which
+will be referred to as delayed treatment from now on.
 
@@
 
-This process was the only time generative AI was used in the writing process. It
-was used to make snippets of code for a specific purpose, revise and
-optimize code and find solutions when written code was not functioning
-properly.
+This process was the only time generative AI was used in the writing
+process. It was used to make snippets of code for a specific purpose,
+revise and optimize code and find solutions when written code was not
+functioning properly.
 
@@
 
-As we were running the logistic regression, additional patients were excluded from
-the study due to missing data. Among the 8,312 patients screened for
-ofi, 7,378 were included in the study. 934 patients were excluded from
-the study due to missing values in one or more variables. This left us
-with 70 patients in the delayed treatment group and 7,308 patients in
-the no delay to treatment group. Figure 1 below illustrates the
-exclusion and inclusion of patients.
+As we were running the logistic regression, additional patients were
+excluded from the study due to missing data. Among the 8,312 patients
+screened for ofi, 7,378 were included in the study. 934 patients were
+excluded from the study due to missing values in one or more variables.
+This left us with 70 patients in the delayed treatment group and 7,308
+patients in the no delay to treatment group. Figure 1 below illustrates
+the exclusion and inclusion of patients.
 
@@
 
+```{=html}
 <!-- You need to refer to tables and figures by numbers in the text, 
 i.e. Figure 1, Table 1 and so on -->
+```
 Table 1 shows the characteristics of the group used in the logistic
@@
 
-In the following table<!-- See comments above-->, the results from the logistical regression
-analysis is displayed.
+In the following table<!-- See comments above-->, the results from the
+logistical regression analysis is displayed.
 
 The odds ratio is calculated both unadjusted and adjusted for other
-variables<!-- What does "other" mean here?-->, as well as a 95% confidence interval. 
-The p value is also shown.
+variables<!-- What does "other" mean here?-->, as well as a 95%
+confidence interval. The p value is also shown.
 
 <!-- Use "associated" instead of correlated or linked etc-->
+
 The analysis shows that some variables seem to have a correlation with
-patients receiving delayed treatment. ISS and respiratory rate (RR) both have
-a trend that patient with lower values seem to have a lower incidence of
-delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI 1.02-1.05)(RR,
-OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR 1.03, 95% CI
-1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the opposite
-correlation, with higher GCS often having lower incidence of delayed
-treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and adjusted (OR
-0.93, 95% CI 0.87-1.00). Number of injuries had the correlation of
-higher number of injuries having higher incidence of delayed treatment
-unadjusted (OR 1.06, 95% CI 1.02-1.10), but that correlation was not
-seen when adjusted for. The three remaining variables, age, gender and
-SBP, showed no correlation with the observed outcome.
+patients receiving delayed treatment. ISS and respiratory rate (RR) both
+have a trend that patient with lower values seem to have a lower
+incidence of delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI
+1.02-1.05)(RR, OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR
+1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the
+opposite correlation, with higher GCS often having lower incidence of
+delayed treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and
+adjusted (OR 0.93, 95% CI 0.87-1.00). Number of injuries had the
+correlation of higher number of injuries having higher incidence of
+delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10), but that
+correlation was not seen when adjusted for. The three remaining
+variables, age, gender and SBP, showed no correlation with the observed
+outcome.
 
@@
 suggests that a more heavily injured and less awake patient is more
-prone to receiving delayed treatment. Compared to a previous
-study analysing the same data but with OFI as the outcome, this result
-is similar to what they found using the same database [@Pat_factors]. As
-the mentioned study suggests, this could be explained by that lower GCS as well as higher ISS
-indicates severe trauma. More injured patients
-often requires a more complex management and more interventions.
-More interventions leave more room for delay to these interventions, why
-they are more prone to delay in treatment. Respiratory rate also had the
+prone to receiving delayed treatment. Compared to a previous study
+analysing the same data but with OFI as the outcome, this result is
+similar to what they found using the same database [@Pat_factors]. As
+the mentioned study suggests, this could be explained by that lower GCS
+as well as higher ISS indicates severe trauma. More injured patients
+often requires a more complex management and more interventions. More
+interventions leave more room for delay to these interventions, why they
+are more prone to delay in treatment. Respiratory rate also had the
 correlation that lower respiratory rate correlated with lower rates of
@@
 treatment because they are more severely injured, therefore a higher
-risk of missing a possible injury. The previous study also found this, but
-the correlation dissapeared when adjusted for, which was hypothesised
-being due to its shared variability with the ISS
+risk of missing a possible injury. The previous study also found this,
+but the correlation dissapeared when adjusted for, which was
+hypothesised being due to its shared variability with the ISS
 variable[@Pat_factors]. In this study, this was the case for number of
@@
 with higher mortality being seen in patients with lower GCS and higher
-ISS [@Mort_sam][@Mort_hef]<!-- Are these two references?-->. Increased age was also a factor linked to
-mortality, which was not seen to be linked to delayed treatment in this
-study. An older patient is more likely to have co-morbidities that might
-affect their mortality but not increase their likelihood of receiving
-delayed treatment the same way ISS and GCS does[@Gioffre2018].
+ISS [@Mort_sammy][@Mort_hefny]. Increased age was also a factor linked
+to mortality, which was not seen to be linked to delayed treatment in
+this study. An older patient is more likely to have co-morbidities that
+might affect their mortality but not increase their likelihood of
+receiving delayed treatment the same way ISS and GCS does[@Gioffre2018].
 
@@
 factor.
+
+It is important to keep in mind that delay in treatment does not have a
+set limit of time as a where under it is within the accepted timeframe
+and if you go over it, it is considered delayed treatment. The audit
+filters flag for a wait for CT over 30 minutes and a wait for operation
+over 60 minutes, which then gets reviewed by previously mentioned two
+specialized nurses. However, if a patient has a longer wait, it does not
+necessarily mean that there was a delay. If it takes a longer time for a
+trauma patient to receive a CT or operation, it might be because there
+was more urgent matters who had to be taken care of, alternatively, the
+MoM conference might deem it as the delay was justified for other
+reasons. The point is that the mean waiting time for CT and operation is
+higher than 30 and 60 minutes respectively when looking at the database.
+Using this audit filter as the definition for delay would not mean
+anything, as the majority of patients would be classified as having
+waited for too long. Instead, it is the MoM conferences that decide if
+the delay had any adverse effects for the patient, such as an injury
+worsening during the time spent waiting without a clear motivation to
+why it had to wait. It is these cases that they label it as the ofi "too
+long until CT" or "Too long until operation".
 
@@
 On discussing possible biases, misclassification bias might also be
-<!-- Do not use contractions, i.e. can't, doesn't etc -->
-present in this study, though it is something that we cannot control in
-this study. As previously mentioned, the possible existence of an OFI is
-decided during the MoM conferences, where they then add it to the
-database. However, they might not always be right in their judgement and
+<!-- Do not use contractions, i.e. can't, doesn't etc --> present in
+this study, though it is something that we cannot control in this study.
+As previously mentioned, the possible existence of an OFI is decided
+during the MoM conferences, where they then add it to the database.
+However, they might not always be right in their judgement and
 misclassify from time to time, leading to patients landing in the wrong
@@
 
+```{=html}
 <!-- Not so sure about this section. What do you mean that the logistic
@@
 the previous paragraphs, so you don't really need this -->
-Another limitation with this study is that logistic regressions
-assumes that the data is binomial, which the data used is not. The
-analysis assumes a linear correlation between the dependent variable and
-the independent variable, which is best analysed in a normal
-distribution. However, Fischers test showed a significant difference
-between the groups in 4 of the 7 variables, all of them being the ones
-who showed an increased or decreased risk of delayed treatment, showing
-that these results still are significant.
+```
+Another limitation with this study is that logistic regressions assumes
+that the data is binomial, which the data used is not. The analysis
+assumes a linear correlation between the dependent variable and the
+independent variable, which is best analysed in a normal distribution.
+However, Fischers test showed a significant difference between the
+groups in 4 of the 7 variables, all of them being the ones who showed an
+increased or decreased risk of delayed treatment, showing that these
+results still are significant.
 
@@
 
+```{=html}
 <!-- Think about what components of your study is generalisible and to 
@@
 treatment, also in other contexts -->
-
+```
 #### Future studies
@@
 In conclusion, based on this small population, the results suggests that
-<!-- I would say "with affected consciousness" instead of "less awake" -->
-it is mostly more injured and less awake trauma patients that receive
-delayed treatment. However, due to the limitations of this study, the
-result should only be seen as something to keep in mind when treating
-patients as further studies need to be conducted on this subject.
+it is mostly more injured and patients with affected consciousness that
+receive delayed treatment. However, due to the limitations of this
+study, the result should only be seen as something to keep in mind when
+treating patients as further studies need to be conducted on this
+subject.
 

### 2023-12-14 lab-book.Rmd
little
code
text

@@
 
+as.gt till word
+
+kommentera att best clinical practice är dt inom 30 min. Många får inte
+det, bara liten andel får det som största problem
+
 etiska tillstånd för detta projekt, skilt från swetrau. Etiska tillstånd
@@
 
-New.subset\$dt_ed_first_ct \<- as.factor(New.subset\$dt_ed_first_ct)
+delay over 30/60 mins not always delay, only if bad
+
+OR 1.03 for each step over 0
+
+We can do better for injured and less conscious, clinicians need to be
+more aware.

### 2023-12-15 manuscriptlongversion.Rmd
text

@@
 opportunities for improvement, one of the more common being delayed
-treatment. *Aim:* This retrospective cohort study aims to identify
-patient level factors that are associated with receiving delayed
-treatment among trauma patients.
-<!-- The first sentence is a bit complicated. Try to simplify, break 
-into multiple sentences --> *Material and Methods:* This study uses the
-database of Karolinska university hospital, the equivalent of a level 1
-trauma center, to run a logistical regression linking the outcome,
-delayed treatment, to variables, observed patient level factors.
-Patients treated for trauma at the hospital between 2012 and 2022 were
-included, with ISS\>9 and/or trauma code activation. Exclusion criterias
-were not being screened for OFI and missing values in any of the
-variables, age, gender, ISS, SBP, RR, GCS and number of injuries.
-*Results:* Higher ISS and RR was linked to higher rates of delay in
-treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI
-1.01-1.08), while a lower GCS also was linked to delayed treatment (OR
-0.93, 95% CI 0.87-1.00). Number of injuries showed a correlation to
-higher rates of delay, but not when adjusted for. The remaining
-variables showed no correlation. *Conclusion:* Our study suggests that
-it is mostly more severely injured and less awake patients that receive
-delayed treatment.
+treatment. *Aim:* This retrospective cohort study aimed to determine the
+associations between different patient level factors and receiving
+delayed treatment. *Material and Methods:* This study used the database
+of Karolinska university hospital to run a logistical regression linking
+the outcome, delayed treatment, to variables, observed patient level
+factors. Patients treated for trauma at the hospital between 2012 and
+2022 were included, with ISS\>9 and/or trauma code activation. Exclusion
+criterias were not being screened for OFI and missing values in any of
+the observed variables. *Results:* Higher ISS and RR was linked to
+higher rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR,
+OR 1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
+treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
+correlation to higher rates of delay, but not when adjusted for. The
+remaining variables showed no correlation. *Conclusion:* Our study
+suggests that it is mostly more severely injured and less awake patients
+that receive delayed treatment.
 
@@
 kallade förbättringsmöjligheter, där försenad vård är en av de
-vanligaste. *Syfte:* Denna retrospektiva kohortstudie syftar till att
-identifiera patientfaktorer kopplade till försenad vård bland
-traumapatienter. *Material och Metod:* Denna studie använder Karolinska
-universitetssjukhusets databas, motsvarande ett traumacenter på nivå 1,
-för att köra en logistisk regression som kopplar utfallet, försenad
-vård, till variabler, observerade patientnivåfaktorer. Patienter som
-behandlades för trauma på sjukhuset mellan 2012 och 2022 inkluderades,
-med ISS\>9 och/eller traumakodaktivering. Exklusionskriterier var om
-patient inte screenades för ofi samt saknade värden i de använda
-variablerna, ålder, kön, ISS, SBT, AF, GCS och antal skador. *Resultat:*
-Högre ISS och AF var kopplade till högre grad av försenad vård (ISS, OR
-1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08), medan ett lägre
-GCS också var kopplad till försenad vård (OR 0.93, 95% CI 0.87-1.00).
-Antalet skador visade först ett samband med högre förseningsfrekvens,
-men inte i den justerade analysen. De återstående variablerna visade
-ingen korrelation. *Slutsats*: Vår studie tyder på att det mestadels är
-svårare skadade och mindre vakna patienter som får försenad behandling.
+vanligaste. *Syfte:* Denna retrospektiva kohortstudie syftade till att
+identifiera associationen mellan patientfaktorer och försenad vård.
+*Material och Metod:* Denna studie använde Karolinska
+universitetssjukhusets databas för att köra en logistisk regression som
+kopplar utfallet, försenad vård, till variabler, observerade
+patientnivåfaktorer. Patienter som behandlades för trauma på sjukhuset
+mellan 2012 och 2022 inkluderades, med ISS\>9 och/eller
+traumakodaktivering. Exklusionskriterier var om patient inte screenades
+för ofi samt om de saknade värden i de undersökta variablerna.
+*Resultat:* Högre ISS och AF var kopplade till högre grad av försenad
+vård (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08),
+medan ett lägre GCS också var kopplad till försenad vård (OR 0.93, 95%
+CI 0.87-1.00). Antalet skador visade först ett samband med högre
+förseningsfrekvens, men inte i den justerade analysen. De återstående
+variablerna visade ingen korrelation. *Slutsats*: Vår studie tyder på
+att det mestadels är svårare skadade och mindre vakna patienter som får
+försenad behandling.
+
+Keywords: Trauma, Delay to treatment, Patient factors.
 
@@
 the body's associated response [@gerdin_risk_2015], is and has long been
-a major cause of death around the world [@WHO_2021].
-<!-- The sentence below is long and overly complicated. Are you saying 
-that trauma systems is an important part of healthcare and its implementation
-has reduced mortality and morbidity?--> Within modern hospital care, the
-trauma system stands as an important part in healthcare and is crucial
-to lowering mortality and morbidity in injured patients through many
-means, such as hospital care, patient follow up and prevention programs
-[@Choi_2021].
+a major cause of death around the world [@WHO_2021]. The trauma system
+plays an important part in modern healthcare and has shown to lower
+mortality and morbidity in injured patients through many means, such as
+hospital care, patient follow up and prevention programs [@Choi_2021].
 
@@
 41,500\$ per patient in high income countries (HIC), depending on the
-country [@Willenberg_2012]. However, it has been shown that this burden
-is not evenly distributed between or within countries. Many social
-factors such as age, sex and social status play a major role in the risk
-of dying from trauma, with young men with low socio-economic status
-being at most risk [@WHO_2021]. But it is not only patient level factors
-that affect what effect trauma has on people. About 90% of all trauma
-related deaths occur in low- and middle-income countries (LMIC), with
-death rates by trauma also being higher than in HIC. Even within these
-LMIC, people of poorer socio-economic status have higher death rates
-from trauma. Problems identified within these countries that contribute
-to these statisitics were infrastructure, education and training,
-attributed to lack of funding, brain drain to HIC and lack of
+country [@Willenberg_2012].However, it has been shown that this burden
+is not evenly distributed between or within countries.
+
+Many social factors such as age, sex and social status play a major role
+in the risk of dying from trauma, with young men with low socio-economic
+status being at most risk [@WHO_2021]. But it is not only patient level
+factors that affect what effect trauma has on people. About 90% of all
+trauma related deaths occur in low- and middle-income countries (LMIC),
+with death rates by trauma also being higher than in HIC. Even within
+these LMIC, people of poorer socio-economic status have higher death
+rates from trauma. Problems identified within these countries that
+contribute to these statisitics were infrastructure, education and
+training, attributed to lack of funding, brain drain to HIC and lack of
 availability of basic amenities [@Shanthakumar_2021]. In HIC where
@@
 ## Trauma systems
-
-<!-- Split long paragraphs -->
 
@@
 trauma centers is associated with a 15-30% decrease in mortality
-[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. During the last
-three decades, the introduction of trauma systems has contributed to
-lowering the incidence of preventable death. This is attributed to the
-improvements in care for acute brain injuries and bleeding control. The
-incidence of late death because of sepsis and multiple organ failure has
-also been lowered, possibly a result of better and earlier resuscitation
-[@Asensio_2008]. Functioning trauma systems saves lives in the hospital,
-but their work outside of the hospital is just as important. Data
-collection from injured patients, such as mechanism of injury or
-mortality, are essential for creating databases that can be used for
-research. In turn, that research can be used for planning injury
-prevention programs that target the most common injuries in the most
-efficient ways, e.g. teen drivers, children, specific occupations etc.
-These injury prevention programs can be planned on trauma center,
-organizational or government level[@ACS].
+[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012].
+
+During the last three decades, the introduction of trauma systems has
+contributed to lowering the incidence of preventable death. This is
+attributed to the improvements in care for acute brain injuries and
+bleeding control. The incidence of late death because of sepsis and
+multiple organ failure has also been lowered, possibly a result of
+better and earlier resuscitation [@Asensio_2008]. Functioning trauma
+systems saves lives in the hospital, but their work outside of the
+hospital is just as important. Data collection from injured patients,
+such as mechanism of injury or mortality, are essential for creating
+databases that can be used for research. In turn, that research can be
+used for planning injury prevention programs that target the most common
+injuries in the most efficient ways, e.g. teen drivers, children,
+specific occupations etc. These injury prevention programs can be
+planned on trauma center, organizational or government level[@ACS].
 
@@
 
-## Mortality and morbidity (MoM) conferences
-
-<!-- Consider renaming this section as "Opportunities for improvement" -->
+## Opportunities for improvement
 
@@
 
-```{=html}
-<!-- Make sure the aim here is exactly the same as in the abstract. Also,
-I suggest that your aim is to determine the associations between different 
-patient level factors and receiving delayed treatment -->
-```
-The aim of this paper is to determine what different patient level
-factors affect the risk of receiving delayed treatment at a trauma
-center.
+This retrospective cohort study aimed to determine the associations
+between different patient level factors and receiving delayed treatment.
 
@@
 
-Ethical permit is required and exists. The ethical review number is
-2021-02541 and 2021-03531. An ethical problem that might be brought up
-is the anonymity of the patients. In the registry, patients social
-security number is present and could theoretically be used to identify
-the patient in question. This will not be a problem though, as in the
-analysis, the social security number is removed from the data before the
-logistical regression is ran. Furthermore, the results will be presented
-with patients grouped together, not individually. This will make it
+An ethical problem that might be brought up is the anonymity of the
+patients. In the registry, patients social security number is present
+and could theoretically be used to identify the patient in question.
+This will not be a problem though, as in the analysis, the social
+security number is removed from the data before the logistical
+regression is ran. Furthermore, the results will be presented with
+patients grouped together, not individually. This will make it
 impossible for anyone to identify patients. During the analysis, a VPN
 was used to access the database to ensure that no data was leaked or
-used for other purposes.
+used for other purposes. Ethical permit is required and exists. The
+ethical review number is 2021-02541 and 2021-03531.
 
@@
 
-The odds ratio is calculated both unadjusted and adjusted for other
-variables<!-- What does "other" mean here?-->, as well as a 95%
-confidence interval. The p value is also shown.
+The odds ratio is calculated both unadjusted and adjusted. A 95%
+confidence interval was used with the p value shown in the table.
 
-<!-- Use "associated" instead of correlated or linked etc-->
-
-The analysis shows that some variables seem to have a correlation with
+The analysis shows that some variables seem to have a association with
 patients receiving delayed treatment. ISS and respiratory rate (RR) both
@@
 adjusted (OR 0.93, 95% CI 0.87-1.00). Number of injuries had the
-correlation of higher number of injuries having higher incidence of
+association of higher number of injuries having higher incidence of
 delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10), but that
-correlation was not seen when adjusted for. The three remaining
-variables, age, gender and SBP, showed no correlation with the observed
+association was not seen when adjusted for. The three remaining
+variables, age, gender and SBP, were not shown to be linked to the
 outcome.
@@
 delayed treatment while a higher ISS had the same correlation. This
-suggests that a more heavily injured and less awake patient is more
+suggests that a more heavily injured and less conscious patient is more
 prone to receiving delayed treatment. Compared to a previous study
@@
 On discussing possible biases, misclassification bias might also be
-<!-- Do not use contractions, i.e. can't, doesn't etc --> present in
-this study, though it is something that we cannot control in this study.
-As previously mentioned, the possible existence of an OFI is decided
-during the MoM conferences, where they then add it to the database.
-However, they might not always be right in their judgement and
+present in this study, though it is something that we cannot control in
+this study. As previously mentioned, the possible existence of an OFI is
+decided during the MoM conferences, where they then add it to the
+database. However, they might not always be right in their judgement and
 misclassify from time to time, leading to patients landing in the wrong
@@
 kept in mind.
-
-```{=html}
-<!-- Not so sure about this section. What do you mean that the logistic
-regression assumes that the data is binomial? Do you mean that it requires 
-a binary outcome? I think you discuss potential limitations well enough in 
-the previous paragraphs, so you don't really need this -->
-```
-Another limitation with this study is that logistic regressions assumes
-that the data is binomial, which the data used is not. The analysis
-assumes a linear correlation between the dependent variable and the
-independent variable, which is best analysed in a normal distribution.
-However, Fischers test showed a significant difference between the
-groups in 4 of the 7 variables, all of them being the ones who showed an
-increased or decreased risk of delayed treatment, showing that these
-results still are significant.
 
@@
 
+Further analysing the results and it's clinical applications, the
+results can be interpreted as that it is mostly injured patients with
+affected consciousness that receive delayed treatment.
+
 #### Generalisability
@@
 patients at the equivalent of a level 1 trauma center, making this study
-generalisable. This is in line with our aim, as it was to make it
-applicable to the whole trauma patient population. However, as it is a
-single center study, where its result only illustrates the situation in
-Stockholm, it is less applicable nationally or globally. This does not
-make it only applicable to the population of Stockholm, but is something
-to keep in mind.
+generalisable. While the population only represents Karolinska
+university hospital, the conclusion is should be applicable to the whole
+trauma population.
 
-```{=html}
-<!-- Think about what components of your study is generalisible and to 
-where. I think that your results are generalisible to  a large extent, in 
-the sense that severly injured patients are at higher risk of delayed 
-treatment, also in other contexts -->
-```
 #### Future studies

### 2023-12-15 manuscriptlongversion.Rmd
text

@@
 
-When looking at the patient demographics, it is clear that the majority
-admitted to trauma care are men, with it being slightly higher in the
-delay to treatment group. However, not high enough to be statistically
-significant. The fact that gender did not show an increased odds ratio
-should be seen as a positive sign. It suggests that among trauma
-patients, there is no systematic delay based on this patient level
-factor.
-
 It is important to keep in mind that delay in treatment does not have a
@@
 Further analysing the results and it's clinical applications, the
-results can be interpreted as that it is mostly injured and less awake
-patients that receive delayed. The clinical applications for this are
-not clear, but caregivers should already be aware that the more injuries
-a patients has, the more room for error exists. It is good though to
-receive a reminder that it's important to always work towards minimizing
-the delay for these injured patients.
+results can be interpreted as that it is mostly injured patients with
+affected consciousness that receive delayed treatment. The clinical
+applications of this can be seen as we need to do more to avoid this
+delay for this patient group. If clinicians are aware of that these
+patients receive delayed treatment more frequently, active steps can be
+taken towards reducing this risk and bettering the trauma care.
 
-Further analysing the results and it's clinical applications, the
-results can be interpreted as that it is mostly injured patients with
-affected consciousness that receive delayed treatment.
+Regarding health equity when looking at the patient demographics, it is
+clear that the majority admitted to trauma care are men, with it being
+slightly higher in the delay to treatment group. However, not high
+enough to be statistically significant. The fact that gender did not
+show an increased odds ratio should be seen as a positive sign. It
+suggests that among trauma patients, there is no systematic delay based
+on this patient level factor.
 

### 2023-12-15 functions/create_factors2.R

@@ -65,9 +65,13 @@ create.factors2 <- function(dataset){
   
   
   ##Replacing NA in OFI_dealy with no delay
-  dataset$OFI_delay[is.na(dataset$OFI_delay)] <- "ANo delay to treatment"
+  dataset$OFI_delay[is.na(dataset$OFI_delay)] <- "No delay to treatment"
   
   dataset$OFI_delay <- as.factor(dataset$OFI_delay)
+  
+  dataset$OFI_delay <- relevel(factors.data2$OFI_delay, ref = "No delay to treatment")
+  
+
   
 
 

### 2023-12-15 functions/new_subset.R

@@ -4,7 +4,7 @@ clean.dataset <- function(dataset){
   ## create new colum for outcome to be observed
   dataset$OFI_delay <- as.factor(ifelse(dataset$Problemomrade_.FMP == "Lång tid till DT", "Delay to treatment",
                                                ifelse(dataset$Problemomrade_.FMP == "Lång tid till op", "Delay to treatment",
-                                                      "ANo delay to treatment")))
+                                                      "No delay to treatment")))
   
   ## remove all columns except the analyzed ones
   dataset <-

### 2023-12-15 functions/view_data.R

@@ -14,8 +14,17 @@ model2 <-
 
 adjusted2 <-
   tbl_regression(model2, exponentiate = TRUE,
-                 pvalue_fun = ~ style_pvalue(.x, digits = 2),) %>%
-  
+                 pvalue_fun = ~ style_pvalue(.x, digits = 2),
+  rename = list(
+    Gender = "Gender",
+    ISS = "ISS",
+    Age = "Age",
+    Number_of_injuries = "Number of injuries",
+    Systolic_blood_pressure = "Systolic blood pressure",
+    Respiratory_rate = "Respiratory rate",
+    Total_GCS = "GCS"
+  )
+  ) %>%
   bold_p() %>%
   bold_labels()
 
@@ -26,7 +35,16 @@ unadjust2 <- tbl_uvregression(data = factors.data2,
                               y = OFI_delay,
                               method.args = list(family = binomial),
                               pvalue_fun = ~ style_pvalue(.x, digits = 2),
-                              hide_n = TRUE
+                              hide_n = TRUE,
+                              rename = list(
+                                Gender = "Gender",
+                                ISS = "ISS",
+                                Age = "Age",
+                                Number_of_injuries = "Number of injuries",
+                                Systolic_blood_pressure = "Systolic blood pressure",
+                                Respiratory_rate = "Respiratory rate",
+                                Total_GCS = "GCS"
+                              )
 ) %>%
   bold_p(t = 0.05) %>%
   bold_labels()

### 2023-12-18 manuscriptlongversion.Rmd
text

@@
 
-```{=html}
-<!-- I suggest you name this section something else, like "Trauma globally" 
-or something like that. Also, divide the second paragraph into two or 
-three paragraphs-->
-```
 Trauma is often divided into two subgroups, blunt force trauma and
@@
 ```
-Table 1 shows the characteristics of the group used in the logistic
+Table1 shows the characteristics of the group used in the logistic
 regression. Most of the patients were men, 68%, with that being a bit
@@
 
-In the following table<!-- See comments above-->, the results from the
-logistical regression analysis is displayed.
+In the following table, result1, the results from the logistical
+regression analysis is displayed.
 
@@
 have a trend that patient with lower values seem to have a lower
-incidence of delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI
-1.02-1.05)(RR, OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR
-1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the
-opposite correlation, with higher GCS often having lower incidence of
-delayed treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and
-adjusted (OR 0.93, 95% CI 0.87-1.00). Number of injuries had the
-association of higher number of injuries having higher incidence of
-delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10), but that
-association was not seen when adjusted for. The three remaining
+incidence of delayed treatment, in both unadjusted
+
+(ISS, OR 1.04, 95% CI 1.02-1.05)(RR, OR 1.06, 95% CI 1.02-1.09) and
+adjusted analyses (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI
+1.01-1.08). GCS had the opposite correlation, with higher GCS often
+having lower incidence of delayed treatment, both unadjusted (OR 0.88,
+95% CI 0.83-0.94) and adjusted (OR 0.93, 95% CI 0.87-1.00). Number of
+injuries had the association of higher number of injuries having higher
+incidence of delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10),
+but that association was not seen when adjusted for. The three remaining
 variables, age, gender and SBP, were not shown to be linked to the

### 2023-12-18 CopyOfcreate_factors1.R

@@ -151,3 +151,32 @@ create.factors <- function(dataset){
   return(dataset)
 }
 
+
+
+
+
+library(gtsummary)
+library(broom)
+
+# Fit the logistic regression model
+model1 <- glm(
+  OFI_delay ~ Gender + ISS + Age + Number_of_injuries + Systolic_blood_pressure + Respiratory_rate + Total_GCS,
+  data = factors.data2,
+  family = binomial
+)
+
+adjusted <- tbl_regression(model1, exponentiate = TRUE,
+                           pvalue_fun = ~style_pvalue(.x, digits = 2)) %>%
+  bold_p() %>%
+  bold_labels() %>%
+  modify_table_styling(
+    columns = var_label(
+      Gender = "New Gender Name",
+      ISS = "New ISS Name",
+      Age = "New Age Name",
+      Number_of_injuries = "New Injuries Name",
+      Systolic_blood_pressure = "New SBP Name",
+      Respiratory_rate = "New Respiratory Rate Name",
+      Total_GCS = "New GCS Name"
+    )
+  )

### 2023-12-18 functions/create_factors2.R

@@ -69,7 +69,7 @@ create.factors2 <- function(dataset){
   
   dataset$OFI_delay <- as.factor(dataset$OFI_delay)
   
-  dataset$OFI_delay <- relevel(factors.data2$OFI_delay, ref = "No delay to treatment")
+  dataset$OFI_delay <- relevel(dataset$OFI_delay, ref = "No delay to treatment")
   
 
   

### 2023-12-18 functions/prepare_data.R

@@ -5,7 +5,7 @@ prepare.data <- function(){
   ## Import data
   data <- rofi::import_data(
     user = "antonw",
-    password = "2-parts-Attack-applied")
+    password = "Analyzer-51-liberty-feared")
     
   ## Create variable combined datasets by merging
   combined.datasets <- rofi::merge_data(data)

### 2023-12-18 functions/table1.R

@@ -1,0 +1,31 @@
+table1 <- function(dataset){
+
+  factors.data2 %>%
+    tbl_summary(
+      by = OFI_delay,
+      label = list(
+        Age = "Age (years)",
+        Gender = "Gender",
+        Total_GCS = "GCS",
+        ISS = "ISS",
+        Number_of_injuries = "Number of injuries",
+        Systolic_blood_pressure = "Systolic blood pressure",
+        Respiratory_rate = "Respiratory rate"
+      ),
+      statistic = list(
+        Age ~ "{mean} ± {sd}",
+        Gender ~ "{mean} ± {sd}",
+        Total_GCS ~ "{median} [IQR]",
+        ISS ~ "{mean} ± {sd}",
+        Number_of_injuries ~ "{mean} ± {sd}",
+        Systolic_blood_pressure ~ "{mean} ± {sd}",
+        Respiratory_rate ~ "{mean} ± {sd}"
+      ),
+      digits = list(all_continuous() ~ c(2, 2))
+    ) %>%
+    add_p()
+  
+  
+  return(table1)
+}
+

### 2023-12-18 table1.R

@@ -1,18 +1,0 @@
-table1 <- function(dataset){
-  
-  factors.data %>%
-    tbl_summary(  # Create summary statistics for different groups
-      by = OFI_delay,
-      label = list(
-        Age = "Age (years)",
-        Gender = "Gender",
-        Total_GCS = "GCS"
-      ),
-      statistic = list(all_continuous() ~ "{mean} ± {sd}"),
-      digits = list(all_continuous() ~ c(2, 2))
-    ) %>%
-    add_p()  # Add p-values
-  
-  
-  return(table1)
-}

### 2023-12-19 manuscriptlongversion.Rmd
seminar version

@@
 
-# Abstract
+# **Trauma and the patient level factors linked with delayed treatment**
 
-*Introduction:* Trauma remains a global health challenge, leading to
+Introduction: Trauma remains a global health challenge, leading to
 millions of deaths each year. The biggest impact can be seen in low to
@@
 opportunities for improvement, one of the more common being delayed
-treatment. *Aim:* This retrospective cohort study aimed to determine the
+treatment. Aim: This retrospective cohort study aimed to determine the
 associations between different patient level factors and receiving
-delayed treatment. *Material and Methods:* This study used the database
-of Karolinska university hospital to run a logistical regression linking
-the outcome, delayed treatment, to variables, observed patient level
+delayed treatment. Material and Methods: This study used the database of
+Karolinska university hospital to run a logistic regression linking the
+outcome, delayed treatment, to variables, observed patient level
 factors. Patients treated for trauma at the hospital between 2012 and
@@
 criterias were not being screened for OFI and missing values in any of
-the observed variables. *Results:* Higher ISS and RR was linked to
-higher rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR,
-OR 1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
+the observed variables. Results: Higher ISS and RR was linked to higher
+rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR
+1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
 treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
 correlation to higher rates of delay, but not when adjusted for. The
-remaining variables showed no correlation. *Conclusion:* Our study
+remaining variables showed no correlation. Conclusion: Our study
 suggests that it is mostly more severely injured and less awake patients
@@
 
-*Inledning:* Trauma är fortfarande en global hälsoutmaning som skördar
+**Trauma och patientnivåfaktorerna kopplade till försenad behandling**
+
+Inledning: Trauma är fortfarande en global hälsoutmaning som skördar
 miljontals liv varje år. Den största effekten kan ses i låg till
 medelinkomstländer, men i höginkomstländer har traumasystem utvecklats
-för att sänka morbiditeten och mortaliteten relaterat till trauma. För
-att ytterliggare förbättra dessa system hålls konferenser för att
+för att minska morbiditeten och mortaliteten relaterat till trauma. För
+att ytterligare förbättra dessa system hålls konferenser för att
 förbättra vården patienter får och identifiera misstag som gjorts, så
 kallade förbättringsmöjligheter, där försenad vård är en av de
-vanligaste. *Syfte:* Denna retrospektiva kohortstudie syftade till att
+vanligaste. Syfte: Denna retrospektiva kohortstudie syftade till att
 identifiera associationen mellan patientfaktorer och försenad vård.
-*Material och Metod:* Denna studie använde Karolinska
+Material och Metod: Denna studie använde Karolinska
 universitetssjukhusets databas för att köra en logistisk regression som
@@
 traumakodaktivering. Exklusionskriterier var om patient inte screenades
-för ofi samt om de saknade värden i de undersökta variablerna.
-*Resultat:* Högre ISS och AF var kopplade till högre grad av försenad
-vård (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08),
-medan ett lägre GCS också var kopplad till försenad vård (OR 0.93, 95%
-CI 0.87-1.00). Antalet skador visade först ett samband med högre
-förseningsfrekvens, men inte i den justerade analysen. De återstående
-variablerna visade ingen korrelation. *Slutsats*: Vår studie tyder på
-att det mestadels är svårare skadade och mindre vakna patienter som får
-försenad behandling.
+för ofi samt om de saknade värden i de undersökta variablerna. Resultat:
+Högre ISS och AF var kopplade till högre grad av försenad vård (ISS, OR
+1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08), medan ett lägre
+GCS också var kopplad till försenad vård (OR 0.93, 95% CI 0.87-1.00).
+Antalet skador visade först ett samband med högre förseningsfrekvens,
+men inte i den justerade analysen. De återstående variablerna visade
+ingen korrelation. Slutsats: Vår studie tyder på att det mestadels är
+svårare skadade och mindre vakna patienter som får försenad behandling.
 
@@
 
-# Introduction
+# **Introduction**
 
-## Definition
+## **Definition**
 
 Trauma, defined as the clinical entity composed of physical injury and
-the body's associated response [@gerdin_risk_2015], is and has long been
-a major cause of death around the world [@WHO_2021]. The trauma system
-plays an important part in modern healthcare and has shown to lower
-mortality and morbidity in injured patients through many means, such as
-hospital care, patient follow up and prevention programs [@Choi_2021].
+the body\'s associated response (1), is and has long been a major cause
+of death around the world (2). The trauma system plays an important part
+in modern healthcare and has shown to lower mortality and morbidity in
+injured patients through many means, such as hospital care, patient
+follow up and prevention programs (3).
 
-## Trauma globally
+## **Trauma globally**
 
@@
 and creates an open wound. Examples of penetrating trauma are gunshot
-wounds and stab wounds [@NIGMS_2020]. Differences exist between the two
-groups, as blunt trauma patients tend to be more injured on arrival to
-the hospital and as such, require more resources and are hospitalized
-for a longer periods [@FITCH2019].
+wounds and stab wounds (4). Differences exist between the two groups, as
+blunt trauma patients tend to be more injured on arrival to the hospital
+and as such, require more resources and are hospitalised for a longer
+period (5).
 
 Today, trauma takes the lives of around 4,4 million people each year,
-almost 8% of all deaths [@WHO_2021]. In the United States, trauma is the
-4th leading cause of death among the general population and the leading
-cause for people between the ages of 1 and 44 [@AAST_2020]. It is also
-an important cause of hospitalization and morbidity among all age
-groups, including seniors, and is responsible for an estimated 10% of
-all years lived with a disability globally. This has a significant
-burden on social and economic level, costing countries billions of US
-dollars each year in healthcare and lost productivity [@connolly_2018].
-Studies estimate the cost of trauma care being between 18,500\$ and
-41,500\$ per patient in high income countries (HIC), depending on the
-country [@Willenberg_2012].However, it has been shown that this burden
-is not evenly distributed between or within countries.
+almost 8% of all deaths (2). In the United States, trauma is the 4th
+leading cause of death among the general population and the leading
+cause for people between the ages of 1 and 44 (6). It is also an
+important cause of hospitalisation and morbidity among all age groups,
+including seniors, and is responsible for an estimated 10% of all years
+lived with a disability globally. This has a significant burden on
+social and economic level, costing countries billions of US dollars each
+year in healthcare and lost productivity (7). Studies estimate the cost
+of trauma care being between 18,500\$ and 41,500\$ per patient in high
+income countries (HIC), depending on the country (8).However, it has
+been shown that this burden is not evenly distributed between or within
+countries.
 
@@
 in the risk of dying from trauma, with young men with low socio-economic
-status being at most risk [@WHO_2021]. But it is not only patient level
-factors that affect what effect trauma has on people. About 90% of all
-trauma related deaths occur in low- and middle-income countries (LMIC),
-with death rates by trauma also being higher than in HIC. Even within
-these LMIC, people of poorer socio-economic status have higher death
-rates from trauma. Problems identified within these countries that
-contribute to these statisitics were infrastructure, education and
-training, attributed to lack of funding, brain drain to HIC and lack of
-availability of basic amenities [@Shanthakumar_2021]. In HIC where
-funding and governance over the healthcare system is better designed,
-functioning trauma systems and dedicated trauma centers exist that have
-been shown to lower mortality but also improve functional outcome in
-trauma patients [@Nirula_2006].
+status being at most risk (2). But it is not only patient level factors
+that affect what effect trauma has on people. About 90% of all trauma
+related deaths occur in low- and middle-income countries (LMIC), with
+death rates by trauma also being higher than in HIC. Even within these
+LMIC, people of poorer socio-economic status have higher death rates
+from trauma. Problems identified within these countries that contribute
+to these statistics were infrastructure, education and training,
+attributed to lack of funding, brain drain to HIC and lack of
+availability of basic amenities (9). In HIC where funding and governance
+over the healthcare system is better designed, functioning trauma
+systems and dedicated trauma centers exist that have been shown to lower
+mortality but also improve functional outcome in trauma patients (10).
 
-## Trauma systems
+## **Trauma systems**
 
-Trauma systems are infrastructures that exist to provide and optimize
+Trauma systems are infrastructures that exist to provide and optimise
 care for injured patients starting with injury recognition and triage,
@@
 functioning trauma system requires strong leadership and engagement at
-the trauma center, regional and national level [@connolly_2018]. This
-system is crucial to provide care for trauma patients, both in reducing
-morbidity and mortality in this patient group. Earlier studies from
-Sweden have shown that treating severely injured patients at a trauma
-center is associated with a 41% lower adjusted 30-day mortality rate
-compared to being treated at a non trauma care center due to them being
-more capable of treating these patients, with potential survival benefit
-increasing with higher injury severity [@Candefjord_2020]. Other studies
-have shown similar results, showing that treating injured patients at
-trauma centers is associated with a 15-30% decrease in mortality
-[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012].
+the trauma center, regional and national level (7). This system is
+crucial to provide care for trauma patients, both in reducing morbidity
+and mortality in this patient group. Earlier studies from Sweden have
+shown that treating severely injured patients at a trauma center is
+associated with a 41% lower adjusted 30-day mortality rate compared to
+being treated at a non trauma care center due to them being more capable
+of treating these patients, with potential survival benefit increasing
+with higher injury severity (11). Other studies have shown similar
+results, showing that treating injured patients at trauma centers is
+associated with a 15-30% decrease in mortality (12--15).
 
@@
 multiple organ failure has also been lowered, possibly a result of
-better and earlier resuscitation [@Asensio_2008]. Functioning trauma
-systems saves lives in the hospital, but their work outside of the
-hospital is just as important. Data collection from injured patients,
-such as mechanism of injury or mortality, are essential for creating
-databases that can be used for research. In turn, that research can be
-used for planning injury prevention programs that target the most common
-injuries in the most efficient ways, e.g. teen drivers, children,
-specific occupations etc. These injury prevention programs can be
-planned on trauma center, organizational or government level[@ACS].
+better and earlier resuscitation (16). Functioning trauma systems saves
+lives in the hospital, but their work outside of the hospital is just as
+important. Data collection from injured patients, such as mechanism of
+injury or mortality, are essential for creating databases that can be
+used for research. In turn, that research can be used for planning
+injury prevention programs that target the most common injuries in the
+most efficient ways, e.g. teen drivers, children, specific occupations
+etc. These injury prevention programs can be planned on trauma center,
+organisational or government level(17).
 
-Trauma teams are multidisciplinary and operate in these centers. They
-play a pivotal role in the treatment of the trauma patient, as they
+Trauma teams are multidisciplinary teams and operate in these centers.
+They play a pivotal role in the treatment of the trauma patient, as they
 provide the initial care in the critical stage of trauma. In Sweden,
-trauma teams are lead by a team leader who is a surgeon, and include
-practitioners from the specialities of intensive care, orthopedics,
-nursing and support staff[@västra_G]. For the trauma team to be
-mobilized, a trauma code has to be activated, often by an emergency
-nurse. The nurse uses information gathered by the first responders to
-assess if the patient fits any of the criteria for trauma code. Most
-healthcare facilities have established criteria or guidelines that
-trigger trauma code activation. These criteria typically include
-mechanisms of injury, physiological criteria, anatomical criteria and
-other specific indicators of severe trauma. There are different levels
-of trauma code with different criteria correlating with severity of
-injury, with level 1 mobilizing the most personnel to the trauma room.
-Once in the trauma room, the team works systematically to manage the
-patients injuries. They handle the most urgent problems first, such as
-airways and breathing, with the aim to rapidly assess and stabilise the
-patient, prioritise their injuries and arrange for site of definitive
-care. [@Hedberg_2020]
+trauma teams are led by a team leader who is a surgeon, and include
+practitioners from the specialities of intensive care, orthopaedics,
+nursing and support staff(18). For the trauma team to be mobilised, a
+trauma code has to be activated, often by an emergency nurse. The nurse
+uses information gathered by the first responders to assess if the
+patient fits any of the criteria for trauma code. Most healthcare
+facilities have established criteria or guidelines that trigger trauma
+code activation. These criteria typically include mechanisms of injury,
+physiological criteria, anatomical criteria and other specific
+indicators of severe trauma. There are different levels of trauma code
+with different criteria correlating with severity of injury, with level
+1 mobilising the most personnel to the trauma room. Once in the trauma
+room, the team works systematically to manage the patient's injuries.
+They handle the most urgent problems first, such as airways and
+breathing, with the aim to rapidly assess and stabilise the patient,
+prioritise their injuries and arrange for a site of definitive care.
+(19)
 
-## Opportunities for improvement
+## **Opportunities for improvement**
 
@@
 registry SweTrau, which functions as a national database designed to
-enable scientific processing with epidemiological technology
-[@SweTrau_2021] This logging of information is important work, as a
-cornerstone of trauma quality improvement programs is multidisciplinary
-MoM conferences. The MoM conference is a meeting where different
-specialities who work with trauma care sit down and discuss deaths and
-complications in order to look for preventable factors. These
+enable scientific processing with epidemiological technology (20). This
+logging of information is important work, as a cornerstone of trauma
+quality improvement programs is multidisciplinary mortality and
+morbidity (MoM) conferences. The MoM conference is a meeting where
+different specialties who work within trauma care sit down and discuss
+deaths and complications in order to look for preventable factors. These
 conferences are performed in many hospitals globally, almost everywhere
 where there are formal medical specialty departments and sometimes in
-smaller hospitals as well [@WHO_2009]. In Sweden, many hospitals have
-mortality conferences on deceased patient cases, but only one has MoM
-conferences to better their trauma care on patients who lived
-[@SweTrau_2023]. The reason being that this expansion of MoM conferences
-takes time and resources, which is why all hospitals don't implement
-this.
+smaller hospitals as well (21). In Sweden, many hospitals have mortality
+conferences on deceased patient cases, but few has MoM conferences to
+better their trauma care on patients who lived (22). The reason being
+that this expansion of MoM conferences takes time and resources, which
+is why all hospitals don\'t implement this.
 
@@
 process is effective, as it has been shown that this review is
-associated with high-quality trauma care [@Santana_2014]. Examples of
-OFI may include lack of resources and management errors. One common OFI
-is delayed treatment. According to previous studies, among the
-preventable or possibly preventable deaths in trauma patients, delay in
-treatment has been identified as a major error contributing to death,
-found in up to 52.9% of patients in said group [@Teixeira_2007]. Delayed
-treatment has been shown to have adverse effects on patients, showing
-why it is such an important issue and the need to develop strategies to
-combat [@Sampalis_1995; @Sampalis_1994]. Although it is such a common
-OFI, the patient level factors associated with delayed treatment are
-poorly understood. There may be several factors that correlate with
-receiving delayed treatment, but they have yet to be identified. Mapping
-these factors may help in identifying patients that might be at risk for
-receiving delayed treatment before it happens.
+associated with high-quality trauma care (23). Examples of OFI may
+include lack of resources and management errors. One common OFI is
+delayed treatment. According to previous studies, among the preventable
+or possibly preventable deaths in trauma patients, delay in treatment
+has been identified as a major error contributing to death, found in up
+to 52.9% of patients in said group (24). Delayed treatment has been
+shown to have adverse effects on patients, showing why it is such an
+important issue and the need to develop strategies to combat (25,26).
+Although it is such a common OFI, the patient level factors associated
+with delayed treatment are poorly understood. There may be several
+factors that correlate with receiving delayed treatment, but they have
+yet to be identified. Mapping these factors may help in identifying
+patients that might be at risk for receiving delayed treatment before it
+happens.
 
-## Aim
+## **Aim**
 
@@
 
-# Methods
+# **Methods**
 
-## Setting
-
-The Karolinska University hospital, which is the equivalent of a trauma
-level 1 hospital, admits 1500 trauma patients each year. To be included
-in the trauma registry, a patient must be over 15 years of age, had
-ISS\>9 and/or had trauma code activation. They get added to both the
-Karolinska Trauma registry as well as the national trauma registry
-(SweTrau). The registry includes data on vital signs, injuries,
-interventions and patient information. Another database, the trauma care
-quality database includes information relevant for MoM conferences. At
-Karolinska University Hospital, patients who die during hospital stay
-gets their case sent to a MoM everytime. However, for the rest of the
-patients, two specialized nurses determine whether they are. Several
-audit filters are set up to flag for patients with possible OFIs. Audit
-filters may include low SBP, high RR, EMS scene time greater than 20
-mins etc. Flagged patients are then reviewed by these two nurses who
-decide if there is a possible OFI, in which case their case is sent to a
-MoM conference for further review.
-
-## Study design
+## **Study design**
 
@@
 
-## Outcome
+## **Setting**
+
+The Karolinska University hospital, which is the equivalent of a trauma
+level 1 hospital, admits 1500 trauma patients each year. To be included
+in the trauma registry, a patient must be over 15 years of age, had
+ISS\>9 and/or had trauma code activation. They are added to both the
+Karolinska Trauma registry as well as the national trauma registry
+(SweTrau). The registry includes data on vital signs, injuries,
+interventions and patient information. Another database, the trauma care
+quality database includes information relevant for MoM conferences. At
+Karolinska University Hospital, patients who die during hospital stay
+get their case sent to a MoM every time. However, for the rest of the
+patients, two specialised nurses determine whether they are. Several
+audit filters are set up to flag for patients with possible OFIs. Audit
+filters may include low SBP, high RR, EMS scene time greater than 20
+mins etc. Flagged patients are then reviewed by these two nurses who
+decide if there is a possible OFI, in which case their case is sent to a
+MoM conference for further review.
+
+## **Outcome**
 
@@
 treatment as identified and decided my the MoM conference. The outcomes
-we used were "long time until CT" and "long time until operation", which
-will be referred to as delayed treatment from now on.
+we used were \"long time until CT\" and \"long time until operation\",
+which will be referred to as delayed treatment from now on.
 
-## Participants
+## **Participants**
 
-The database only includes patients 15 years old and above, ISS \>9
-and/or trauma team activation. Inclusion in the study further required
-that the patients has been assessed for OFIs. Patients were excluded if
-they were missing information in one of the examined patient factors.
+The database, as previously mentioned, only includes patients 15 years
+old and above, ISS \>9 and/or trauma team activation. Inclusion in the
+study further required that the patients have been assessed for OFIs.
+Patients were excluded if they were missing information in one of the
+examined patient factors.
 
-## Variables
+## **Variables**
 
@@
 
-## Data sources/measurement
+## **Data sources/measurement**
 
-The data comes from the trauma registry and trauma care quality
-database.
+The data comes from the trauma registry and trauma care quality database
+and is accessed through the program rstudio.
 
-## Bias
+## **Bias**
 
 Selection bias as this is a retrospective post hoc cohort study where
-the outcome has occurred before the study began.
+the outcome has occurred before the study began. This might have
+affected how we chose variables, opting for variables with more
+interesting results.
 
 Misclassification bias as the conference may reach a consensus that is
-wrong, alternatively that the two specialized nurses going through the
+wrong, alternatively that the two specialised nurses going through the
 patients flagged for audit filters might misclassify patients or that
@@
 
-## Study size
+## **Study size**
 
-The study size is all of the eligible patients that was treated at the
+The study size is all of the eligible patients that were treated at the
 hospital between 2012 and 2022. Starting 2013 only a subgroup of
@@
 
-## Statistical methods
+## **Statistical methods**
 
-Multivariable and univariable logistical regression. A 5% significance
-level and 95% confidence levels will be used.
-
-Rstudio was used to clean and prepare the data from the database. The
-data was processed to have the variables customized to fit into the
-logistical regression model, such as making them into factors, renaming
-outcomes in the database and replacing or removing unknown values. The
-variables were either binary or continuous. While writing the code, it
-was tested on scrambled data so to avoid selection bias as much as
-possible. When the code was finished, it was run on the unscrambled data
-for a result.
+Multivariable and univariable logistic regression. A 5% significance
+level and 95% confidence level was used. Rstudio was used to clean and
+prepare the data from the database. The data was processed to have the
+variables customised to fit into the logistical regression model, such
+as making them into factors, renaming outcomes in the dataframe and
+replacing or removing unknown values. The variables were either binary
+or continuous. While writing the code, it was tested on scrambled data
+so as to avoid selection bias as much as possible. When the code was
+finished, it was run on the unscrambled data for a result.
 
@@
 
-The logistical regression is run through Rstudio using the glm command.
-It was ran several times to make an adjusted table and 7 unadjusted, one
+The logistical regression was run through Rstudio using the glm command.
+It was run several times to make an adjusted table and 7 unadjusted, one
 for each variable. To make the tables, the package gtsummary was used to
-visualize the adjusted and unadjusted regression individually, then
+visualise the adjusted and unadjusted regression individually, then
 merging them together into one table.
@@
 process. It was used to make snippets of code for a specific purpose,
-revise and optimize code and find solutions when written code was not
+revise and optimise code and find solutions when written code was not
 functioning properly.
 
-## Ethical considerations
+## **Ethical considerations**
 
 An ethical problem that might be brought up is the anonymity of the
-patients. In the registry, patients social security number is present
+patients. In the registry, a patient's social security number is present
 and could theoretically be used to identify the patient in question.
 This will not be a problem though, as in the analysis, the social
-security number is removed from the data before the logistical
-regression is ran. Furthermore, the results will be presented with
+security number is removed from the dataframe before the logistical
+regression was run. Furthermore, the results will be presented with
 patients grouped together, not individually. This will make it
@@
 
+**Figure 1. The flowchart illustrates how patients were included and
+excluded from the study, with what group they are in and the amount of
+patients in that group illustrated in the boxes**
+
 ![](Rplot.png)
 
-```{=html}
-<!-- You need to refer to tables and figures by numbers in the text, 
-i.e. Figure 1, Table 1 and so on -->
-```
-Table1 shows the characteristics of the group used in the logistic
+Figure 2 shows the characteristics of the group used in the logistic
 regression. Most of the patients were men, 68%, with that being a bit
@@
 
+**Figure 2. Characteristics of patients included in study. Shown is the
+amount of participants in each group and overall with mean values and
+standard deviations, except in gender, where the amount of participants
+in each category is shown instead.**
+
 ```{r message = FALSE}
@@
 
-In the following table, result1, the results from the logistical
+**Abbreviations: N - number of patients, ISS - Injury Severity Score,
+GCS - Glasgow Coma Scale.**
+
+**a P-value shows fisher's test for gender and Welch t-test for the rest
+of the variables.**
+
+In the following table, figure 3, the results from the logistical
 regression analysis is displayed.
@@
 
-The analysis shows that some variables seem to have a association with
+The analysis shows that some variables seem to have an association with
 patients receiving delayed treatment. ISS and respiratory rate (RR) both
 have a trend that patient with lower values seem to have a lower
-incidence of delayed treatment, in both unadjusted
-
-(ISS, OR 1.04, 95% CI 1.02-1.05)(RR, OR 1.06, 95% CI 1.02-1.09) and
-adjusted analyses (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI
-1.01-1.08). GCS had the opposite correlation, with higher GCS often
-having lower incidence of delayed treatment, both unadjusted (OR 0.88,
-95% CI 0.83-0.94) and adjusted (OR 0.93, 95% CI 0.87-1.00). Number of
-injuries had the association of higher number of injuries having higher
-incidence of delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10),
-but that association was not seen when adjusted for. The three remaining
+incidence of delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI
+1.02-1.05)(RR, OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR
+1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the
+opposite correlation, with higher GCS often having lower incidence of
+delayed treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and
+adjusted (OR 0.93, 95% CI 0.87-1.00). Number of injuries had the
+association of a higher number of injuries having higher incidence of
+delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10), but that
+association was not seen when adjusted for. The three remaining
 variables, age, gender and SBP, were not shown to be linked to the
 outcome.
+
+**Figure 3. Results of logistic regression with odds ratio, 95%
+confidence interval and p-value. Both unadjusted values and adjusted
+values are shown.**
 
@@
 
-# Discussion
+**Abbreviations: OR - Odds ratio, CI - Confidence interval, ISS - Injury
+Severity Score, GCS - Glasgow Coma Scale.**
+
+# **Discussion**
 
@@
 analysing the same data but with OFI as the outcome, this result is
-similar to what they found using the same database [@Pat_factors]. As
-the mentioned study suggests, this could be explained by that lower GCS
-as well as higher ISS indicates severe trauma. More injured patients
-often requires a more complex management and more interventions. More
+similar to what they found using the same database (27). As the
+mentioned study suggests, this could be explained by that lower GCS as
+well as higher ISS indicates severe trauma. More injured patients often
+require more complex management and more interventions. More
 interventions leave more room for delay to these interventions, why they
@@
 the triage guideline, both too low and too high respiratory rate is not
-desirable[@EHCP2015]. This might have a similar explanation to ISS and
-GCS, that patients with tachypnea might receive higher rates of delayed
+desirable(28). This might have a similar explanation to ISS and GCS,
+that patients with tachypnea might receive higher rates of delayed
 treatment because they are more severely injured, therefore a higher
 risk of missing a possible injury. The previous study also found this,
-but the correlation dissapeared when adjusted for, which was
+but the correlation disappeared when adjusted for, which was
 hypothesised being due to its shared variability with the ISS
-variable[@Pat_factors]. In this study, this was the case for number of
-injuries, as it and ISS is often correlated as a more severely injured
-patient will both score higher on ISS and number of injuries. Therefore,
-we hypothesize that the effect of number of injuries become
-insignificant when adjusting for ISS.
+variable(27). In this study, this was the case for number of injuries,
+with it and ISS often being linked, as a more severely injured patient
+will both score higher on ISS and number of injuries. Therefore, we
+hypothesise that the effect of number of injuries becomes insignificant
+when adjusting for ISS.
 
 Not many studies regarding trauma and patient level factors have had
-delayed treatment as their outcome, while many analysing mortality. Some
-similarities exist with which patient level factors predict the outcome,
-with higher mortality being seen in patients with lower GCS and higher
-ISS [@Mort_sammy][@Mort_hefny]. Increased age was also a factor linked
-to mortality, which was not seen to be linked to delayed treatment in
-this study. An older patient is more likely to have co-morbidities that
-might affect their mortality but not increase their likelihood of
-receiving delayed treatment the same way ISS and GCS does[@Gioffre2018].
+delayed treatment as their outcome, as many are analysing mortality.
+Some similarities exist with which patient level factors predict the
+outcome, with higher mortality being seen in patients with lower GCS and
+higher ISS [(29)](30). Increased age was also a factor linked to
+mortality, which was not seen to be linked to delayed treatment in this
+study. An older patient is more likely to have comorbidities that might
+affect their mortality but not increase their likelihood of receiving
+delayed treatment the same way ISS and GCS does(31).
 
@@
 over 60 minutes, which then gets reviewed by previously mentioned two
-specialized nurses. However, if a patient has a longer wait, it does not
+specialised nurses. However, if a patient has a longer wait, it does not
 necessarily mean that there was a delay. If it takes a longer time for a
 trauma patient to receive a CT or operation, it might be because there
-was more urgent matters who had to be taken care of, alternatively, the
+were more urgent matters who had to be taken care of, alternatively, the
 MoM conference might deem it as the delay was justified for other
-reasons. The point is that the mean waiting time for CT and operation is
-higher than 30 and 60 minutes respectively when looking at the database.
-Using this audit filter as the definition for delay would not mean
-anything, as the majority of patients would be classified as having
-waited for too long. Instead, it is the MoM conferences that decide if
-the delay had any adverse effects for the patient, such as an injury
-worsening during the time spent waiting without a clear motivation to
-why it had to wait. It is these cases that they label it as the ofi "too
-long until CT" or "Too long until operation".
+reasons. The point being, it is clear that the mean waiting time for CT
+and operation is higher than 30 and 60 minutes respectively when looking
+at the database. Using this audit filter as the definition for delay
+would not mean anything, as the majority of patients would be classified
+as having waited for too long. Instead, it is the MoM conferences that
+decide if the delay had any adverse effects for the patient, such as an
+injury worsening during the time spent waiting without a clear
+motivation to why it had to wait. It is these cases that they label it
+as the "too long until CT '' or "Too long until operation". Meaning, all
+delays in the database are actual assessed delays, not just a time limit
+that has been surpassed.
 
-#### Limitations
+The results of the study are expected as previous similar studies have
+come to similar conclusions. A goal to strive towards in trauma care
+through research and improvement programs is that no patient level
+factors should be linked with a higher risk of delayed treatment. The
+variables found in this study to be linked with delayed treatment are
+however factors that originate from the trauma, not previously existing
+in the patient such as age or gender. This implies that patients are not
+systematically discriminated against based on their gender or age, which
+is a positive sign seen from a health equity point of view.
+
+#### **Strengths and Limitations**
 
@@
 variable separates a predictor variable completely, which occurs when
-the predictor is only associated with one outcome value [@OARC]. In the
+the predictor is only associated with one outcome value (32). In the
 case of these variables, it was only RTS 3 and 4 who had patients with
@@
 considered expanding the outcome group to include more OFI, but opted
-for the solution used in the study instead as we were solely interested
-in this OFI for this study.
+for the solution used in the study instead as we felt that including
+more OFI would lead to less clinically applicable results.
 
@@
 before reaching the MoM conference. The patients included in the
-conference are selected through audit filters and two specialized
+conference are selected through audit filters and two specialised
 nurses, and patients might be missed in this stage. This bias should not
@@
 
-#### Clinical applications
+#### **Clinical applications**
 
-<!-- Revise this section now after our meeting -->
+Further analysing the results and its clinical applications, the results
+can be interpreted as that it is mostly injured patients with affected
+consciousness that receive delayed treatment. The clinical applications
+of this can be seen as we need to do more to avoid this delay for this
+patient group. The study highlights the urgency in these patients and
+the heightened risk for delay they face due to their situation. If
+clinicians are aware that these patients receive delayed treatment more
+frequently, active steps can be taken towards reducing this risk and
+bettering the trauma care.
 
-Further analysing the results and it's clinical applications, the
-results can be interpreted as that it is mostly injured patients with
-affected consciousness that receive delayed treatment. The clinical
-applications of this can be seen as we need to do more to avoid this
-delay for this patient group. If clinicians are aware of that these
-patients receive delayed treatment more frequently, active steps can be
-taken towards reducing this risk and bettering the trauma care.
+Further applying the study clinically, it is also important to educate
+clinicians on what steps went wrong in these patients. If clinicians
+know the most common mistakes made in this patient group, they might be
+less likely to make these mistakes themselves, possibly lowering the
+risk of receiving delayed treatment for these patients with these
+factors. The results of this study and similar ones should be
+communicated to trauma centers, as the goal is to always better the
+trauma care for patients.
 
-Regarding health equity when looking at the patient demographics, it is
-clear that the majority admitted to trauma care are men, with it being
-slightly higher in the delay to treatment group. However, not high
-enough to be statistically significant. The fact that gender did not
-show an increased odds ratio should be seen as a positive sign. It
-suggests that among trauma patients, there is no systematic delay based
-on this patient level factor.
+#### **Generalisability**
 
-#### Generalisability
-
-Discussing the generalisability of this study, we included the all the
+Discussing the generalisability of this study, we included all the
 trauma patients treated and Karolinska University Hospital and excluded
@@
 generalisable. While the population only represents Karolinska
-university hospital, the conclusion is should be applicable to the whole
+university hospital, the conclusion should be applicable to the whole
 trauma population.
 
-#### Future studies
+#### **Future studies**
 
-In future studies building upon this one, various methods to further
-examine the correlation between delayed treatment and various variables,
-alternatively using another OFI as the outcome variable. Examining
-different variables such as highest hospital care lever, intubation
-status or first emergency procedure performed might be interesting
-variables, among others. To be able to do this, more patients in the
-outcome group would be needed, either by expanding the register to
-several hospitals, or expanding the outcome group by adding different
-ofi. This however would make the study less specific for delayed
-treatment. Further studies might also look into different ofi other than
-delayed treatment, such as missed injury, faulty communication or
-triage.
+In future studies, various methods to further examine the association
+between delayed treatment and various variables, alternatively using
+another OFI as the outcome variable. Examining different variables such
+as highest hospital care level, intubation status or first emergency
+procedure performed might be interesting variables, among others. To be
+able to do this, more patients in the outcome group would be needed,
+either by expanding the register to several hospitals, or expanding the
+outcome group by adding different ofi. This however would make the study
+less specific for delayed treatment. Further studies might also look
+into different ofi other than delayed treatment, such as missed injury,
+faulty communication or triage.
 
-## Conclusion
+## **Conclusion**
 
-In conclusion, based on this small population, the results suggests that
-it is mostly more injured and patients with affected consciousness that
-receive delayed treatment. However, due to the limitations of this
-study, the result should only be seen as something to keep in mind when
-treating patients as further studies need to be conducted on this
-subject.
+In conclusion, based on this single trauma center, the results suggest
+that it is mostly more injured and patients with affected consciousness
+that receive delayed treatment. Clinicians should be aware of this as
+they can take steps to lower that risk for this patient group. Further
+studies need to be conducted on this subject to further examine the link
+between delayed treatment and other patient level factors.
 
-#
+# **Acknowledgements**
+
+I would like to extend my gratitude to my supervisor Martin Gerdin and
+my two co-supervisors Jonatan Attergrim and Kelvin Szolnoky for always
+being quick with providing valuable help and feedback during all stages
+of the writing process as well as guidance during the data analysis part
+of the project.

### 2023-12-19 manuscriptlongversion.Rmd
seminar version

@@
 Trauma, defined as the clinical entity composed of physical injury and
-the body\'s associated response (1), is and has long been a major cause
+the body's associated response (1), is and has long been a major cause
 of death around the world (2). The trauma system plays an important part
@@
 that this expansion of MoM conferences takes time and resources, which
-is why all hospitals don\'t implement this.
+is why all hospitals don't implement this.
 
@@
 treatment as identified and decided my the MoM conference. The outcomes
-we used were \"long time until CT\" and \"long time until operation\",
-which will be referred to as delayed treatment from now on.
+we used were "long time until CT" and "long time until operation", which
+will be referred to as delayed treatment from now on.
 
@@
 motivation to why it had to wait. It is these cases that they label it
-as the "too long until CT '' or "Too long until operation". Meaning, all
+as the "too long until CT '' or"Too long until operation". Meaning, all
 delays in the database are actual assessed delays, not just a time limit

### 2023-12-20 manuscriptlongversion.Rmd
text

@@
 
+#### **Clinical applications**
+
 Further analysing the results and its clinical applications, the results
@@
 of this can be seen as we need to do more to avoid this delay for this
-patient group. The study highlights the urgency in these patients and
-the heightened risk for delay they face due to their situation. If
-clinicians are aware that these patients receive delayed treatment more
-frequently, active steps can be taken towards reducing this risk and
-bettering the trauma care.
+patient group. It is important to remember that the delays are all
+patients that have received adverse outcomes because of their delay.
+Some might include injuries being missed or not being treated. The study
+highlights the urgency in these patients and the heightened risk for
+delay they face due to their situation. If clinicians are aware that
+these patient factors receive delayed treatment more frequently, active
+steps can be taken towards reducing this risk and bettering the trauma
+care.
+
+This also highlights the importance of adequate diagnostic imaging when
+the patient is stabilised. It occurs that patients are stabilised in the
+operating room (OR), gets taken to the CT then back to the OR. This is
+important as to not miss any injury, especially in severely injured
+patients where it can be difficult to identify all injuries without
+diagnostic imaging. When clinicians have access to these instruments
+they should be used whenever deemed necessary, as studies suggest it has
+a positive effect on patient outcome [@Hajibandeh_2014].
 
 Further applying the study clinically, it is also important to educate
-clinicians on what steps went wrong in these patients. If clinicians
-know the most common mistakes made in this patient group, they might be
-less likely to make these mistakes themselves, possibly lowering the
-risk of receiving delayed treatment for these patients with these
-factors. The results of this study and similar ones should be
-communicated to trauma centers, as the goal is to always better the
-trauma care for patients.
+clinicians on what steps went wrong in these patients and what could
+have been done to prevent it. If clinicians know the most common
+mistakes made in this patient group, they might be less likely to make
+these mistakes themselves, possibly lowering the risk of receiving
+delayed treatment for these patients with these factors. The results of
+this study and similar ones should be communicated to trauma centers, as
+the goal is to always better the trauma care for patients.
 

### 2023-12-20 CopyOfcreate_factors1.R

@@ -180,3 +180,26 @@ adjusted <- tbl_regression(model1, exponentiate = TRUE,
       Total_GCS = "New GCS Name"
     )
   )
+
+
+
+
+
+
+tbl_summary(
+  by = OFI_delay,
+  data = factors.data2,
+  label = list(
+    Age = "Age (years)",
+    Gender = "Gender",
+    Total_GCS = "GCS",
+    ISS = "ISS",
+    Number_of_injuries = "Number of injuries",
+    Systolic_blood_pressure = "Systolic blood pressure",
+    Respiratory_rate = "Respiratory rate"
+  ),
+  statistic = list(c("Total_GCS", "ISS") ~ "{median} ({p25}, {p75})"),
+  digits = list(all_continuous() ~ c(2, 2))
+) %>%
+  add_p() %>%
+  add_overall()

### 2024-01-09 manuscriptlongversion.Rmd
possible final draft

@@
 
-These results show a possible correlation between some variables and the
-presence of delayed treatment. A lower GCS had a correlation with
-delayed treatment while a higher ISS had the same correlation. This
-suggests that a more heavily injured and less conscious patient is more
+The aim was to determine the associations between patient level factors
+and receiving delayed treatment in adult trauma patients. A logistic
+regression on seven variables showed that higher ISS and RR and lower
+GCS was significantly associated with patients receiving delayed
+treatment.
+
+This suggests that a heavily injured and less conscious patient is more
 prone to receiving delayed treatment. Compared to a previous study
-analysing the same data but with OFI as the outcome, this result is
-similar to what they found using the same database (27). As the
-mentioned study suggests, this could be explained by that lower GCS as
-well as higher ISS indicates severe trauma. More injured patients often
-require more complex management and more interventions. More
-interventions leave more room for delay to these interventions, why they
-are more prone to delay in treatment. Respiratory rate also had the
-correlation that lower respiratory rate correlated with lower rates of
-delayed treatment among the patients. As the interval is \>10 to \<29 in
-the triage guideline, both too low and too high respiratory rate is not
-desirable(28). This might have a similar explanation to ISS and GCS,
-that patients with tachypnea might receive higher rates of delayed
-treatment because they are more severely injured, therefore a higher
-risk of missing a possible injury. The previous study also found this,
-but the correlation disappeared when adjusted for, which was
-hypothesised being due to its shared variability with the ISS
-variable(27). In this study, this was the case for number of injuries,
-with it and ISS often being linked, as a more severely injured patient
-will both score higher on ISS and number of injuries. Therefore, we
-hypothesise that the effect of number of injuries becomes insignificant
-when adjusting for ISS.
+analysing the same data but with OFI as the observed outcome, this
+result is similar to what they found[@Pat_factors]. As the mentioned
+study suggests, this could be explained by that lower GCS as well as a
+higher ISS indicates severe trauma. More injured patients often require
+more complex management and a higher number of interventions, increasing
+the risk of delayed treatment.
+
+Respiratory rate had the correlation that a lower respiratory rate was
+associated with lower rates of delayed treatment among the patients. As
+the interval is \>10 to \<29 in the triage guideline, both too low and
+too high respiratory rate is not desirable[@EHCP2015]. This could have a
+similar explanation to ISS and GCS, that patients with tachypnea might
+receive higher rates of delayed treatment because they are more severely
+injured, therefore a higher risk of missing a possible injury. The
+previous study also found this, but the correlation disappeared when
+adjusted for, which said study hypothesised being due to its shared
+variability with the ISS variable[@Pat_factors]. In this study, the same
+correlation could be seen with NOI. A more severely injured patient will
+score higher on both ISS and NOI. Therefore, our thought is that the
+effect of NOI becomes insignificant when adjusting for ISS.
 
@@
 delayed treatment as their outcome, as many are analysing mortality.
-Some similarities exist with which patient level factors predict the
+Some similarities exist between which patient level factors predict the
 outcome, with higher mortality being seen in patients with lower GCS and
-higher ISS [(29)](30). Increased age was also a factor linked to
-mortality, which was not seen to be linked to delayed treatment in this
-study. An older patient is more likely to have comorbidities that might
-affect their mortality but not increase their likelihood of receiving
-delayed treatment the same way ISS and GCS does(31).
+higher ISS [@Mort_sammy][@Mort_hefny]. Advanced age has also been
+identified as a contributory factor to increased mortality. However, it
+is noteworthy that this study did not establish a significant
+association between age and delayed treatment. Unlike ISS and GCS, which
+are indicative of the severity of injuries and the immediate treatment
+needs, advanced age is more likely to introduce comorbidities that
+affect overall mortality but may not necessarily heighten the
+probability of receiving delayed treatment[@Gioffre2018].
 
-It is important to keep in mind that delay in treatment does not have a
-set limit of time as a where under it is within the accepted timeframe
-and if you go over it, it is considered delayed treatment. The audit
-filters flag for a wait for CT over 30 minutes and a wait for operation
+When discussing the outcome delay in treatment used in this study, it is
+not a set time limit where if the wait exceeds it, it is considered
+delayed treatment. The audit filters used at Karolinska University
+Hospital flag for a wait for CT over 30 minutes and a wait for operation
 over 60 minutes, which then gets reviewed by previously mentioned two
 specialised nurses. However, if a patient has a longer wait, it does not
-necessarily mean that there was a delay. If it takes a longer time for a
-trauma patient to receive a CT or operation, it might be because there
-were more urgent matters who had to be taken care of, alternatively, the
-MoM conference might deem it as the delay was justified for other
-reasons. The point being, it is clear that the mean waiting time for CT
-and operation is higher than 30 and 60 minutes respectively when looking
-at the database. Using this audit filter as the definition for delay
-would not mean anything, as the majority of patients would be classified
-as having waited for too long. Instead, it is the MoM conferences that
-decide if the delay had any adverse effects for the patient, such as an
-injury worsening during the time spent waiting without a clear
-motivation to why it had to wait. It is these cases that they label it
-as the "too long until CT '' or"Too long until operation". Meaning, all
-delays in the database are actual assessed delays, not just a time limit
-that has been surpassed.
+necessarily mean that there was a delay. 30 and 60 minutes are time
+limits set to have a high sensitivity and low specificity in the audit
+filters. Whether it is a delay is decided by the MoM conferences. It is
+clear that the mean waiting time for CT and operation is higher than 30
+and 60 minutes respectively when looking at the database. Using this
+audit filter as the definition for delay would not mean anything, as the
+majority of patients would be classified as having waited for too long.
+Instead, the MoM conferences decide if the wait had any adverse effects
+on the patients, such as an injury worsening during this time. It is
+these cases that they label it as the \"too long until CT\" or \"Too
+long until operation\". All delays in the database are delays that
+adversely affected the patients.
+
+#### **Strengths and Limitations**
+
+In this study, one of the limitations was the number of patients having
+the recorded outcome being quite small, leading us into having to limit
+the number of variables we wanted to examine. Furthermore, the original
+plan was to divide RR, SBP and GCS into revised trauma score (RTS)
+categories from 1-4, since this would have allowed for a more detailed
+analysis. However, the sample size was too small in the outcome group
+which led us to resort to making these variables continuous. We
+considered expanding the outcome group to include more OFI, but opted
+for the solution used in the study instead as we felt that including
+more OFI would lead to less clinically applicable results.
+
+The problem mentioned above with too few patients in the outcome group
+made the study into a post hoc analysis. We were not aware of the number
+of patients who received delayed treatment when we handled the scrambled
+data as we wanted to minimise selection bias. When we ran the code on
+the real data it became clear that the sample size was too small for our
+approach. Therefore, we had to change the variables after having seen
+the real data, making it a post hoc analysis. This could in theory open
+up for selection bias, as we might have chosen variables which give us
+results we are looking for. We did our best to remove variables at
+random, but it is possible that we subconsciously removed variables that
+would change the result in a way benefiting us. The best way to have
+gone about this would have been to calculate the amount of people who
+received delayed treatment before choosing the variables. Then, choosing
+an amount of variables that would work with the amount of patients. This
+would have avoided making it post hoc and minimised selection bias.
+
+On discussing possible biases, misclassification bias might also be
+present in this study, though it is something that we cannot control in
+this study. As previously mentioned, the possible existence of an OFI is
+decided during the MoM conferences. However, they might not always be
+correct in their judgement and occasionally misclassify, leading to
+patients being in the wrong group in the analysis. Also, some patients
+might have been missed even before reaching the MoM conference. The
+patients included in the conference are selected through audit filters
+and two specialised nurses, and patients might be missed in this stage.
+This bias should not be big enough to have a significant impact on the
+results, but can be kept in mind.
+
+#### **Clinical applications**
+
+Analysing the results and its clinical applications, the results can be
+interpreted as that it is mostly injured patients with affected
+consciousness that receive delayed treatment. The clinical applications
+of this are that clinicians need to do more to avoid this delay for this
+patient group. All patients in the outcome group are patients that
+experienced adverse effects because of their delayed treatment. Some
+might include injuries being missed or not being treated, leading to
+unnecessary suffering for the patient. Clinicians being aware of the
+patient level factors at risk for delayed treatment might help with
+their decision making when caring for these patients.
+
+It is important to highlight the importance of adequate diagnostic
+imaging when the patient is stabilised, as a substantial part of the
+patients in the outcome group were patients who waited too long for
+diagnostic imaging. To lower this risk, it occurs that patients are
+stabilised in the operating room (OR), gets taken to the CT then back to
+the OR. This is done to minimise the risk of missing an injury,
+especially in severely injured patients where it can be difficult to
+identify all injuries without diagnostic imaging. When clinicians have
+access to these instruments they should be used whenever deemed
+necessary, as studies suggest it has a positive effect on patient
+outcome [@Hajibandeh_2014].
+
+#### Equity
 
@@
 variables found in this study to be linked with delayed treatment are
-however factors that originate from the trauma, not previously existing
-in the patient such as age or gender. This implies that patients are not
-systematically discriminated against based on their gender or age, which
-is a positive sign seen from a health equity point of view.
-
-#### **Strengths and Limitations**
-
-In this study, one of the limitations was the number of people having
-the recorded outcome being quite small, leading us into having to limit
-the number of variables we wanted to examine. Furthermore, the original
-plan was to divide RR, SBP and GCS into revised trauma score (RTS)
-categories, from 1-4. However, many of the RTS variables had few to no
-patients with the observed outcome. This led to separation in the
-logistic regression in these variables, referring to when an outcome
-variable separates a predictor variable completely, which occurs when
-the predictor is only associated with one outcome value (32). In the
-case of these variables, it was only RTS 3 and 4 who had patients with
-the observed outcome, which left RTS 1 and 2 with confidence intervals
-of 0.00 to infinity. This way would have been the best way to examine
-these variables, but as that was not possible with this group of
-patients, we resorted to making these variables continuous. We
-considered expanding the outcome group to include more OFI, but opted
-for the solution used in the study instead as we felt that including
-more OFI would lead to less clinically applicable results.
-
-The problem mentioned above with too few patients in the outcome group
-made this into a post hoc analysis. As we did not know the number of
-people who received delayed treatment when we handled the scrambled
-data, we ran the code on the real data when we had too many variables
-for the dataset. Therefore, we had to change the variables after having
-seen the real data, making it a post hoc analysis. This could in theory
-open up for possible selection bias, as we might choose variables which
-give us results we are looking for. However, all the variables included
-in the final analysis were already included in the first one. Variables
-were only removed, not added after having seen the real data. The best
-way to have gone about this would have been to just look at the amount
-of people who received delayed treatment before choosing the variables.
-Then, choosing an amount of variables that would work with the amount of
-patients. This would have avoided making it post hoc and changing the
-variables after seeing the data.
-
-On discussing possible biases, misclassification bias might also be
-present in this study, though it is something that we cannot control in
-this study. As previously mentioned, the possible existence of an OFI is
-decided during the MoM conferences, where they then add it to the
-database. However, they might not always be right in their judgement and
-misclassify from time to time, leading to patients landing in the wrong
-group in the analysis. Also, some patients might have been missed even
-before reaching the MoM conference. The patients included in the
-conference are selected through audit filters and two specialised
-nurses, and patients might be missed in this stage. This bias should not
-be big enough to have a significant impact on the results, but can be
-kept in mind.
-
-#### **Clinical applications**
-
-#### **Clinical applications**
-
-Further analysing the results and its clinical applications, the results
-can be interpreted as that it is mostly injured patients with affected
-consciousness that receive delayed treatment. The clinical applications
-of this can be seen as we need to do more to avoid this delay for this
-patient group. It is important to remember that the delays are all
-patients that have received adverse outcomes because of their delay.
-Some might include injuries being missed or not being treated. The study
-highlights the urgency in these patients and the heightened risk for
-delay they face due to their situation. If clinicians are aware that
-these patient factors receive delayed treatment more frequently, active
-steps can be taken towards reducing this risk and bettering the trauma
-care.
-
-This also highlights the importance of adequate diagnostic imaging when
-the patient is stabilised. It occurs that patients are stabilised in the
-operating room (OR), gets taken to the CT then back to the OR. This is
-important as to not miss any injury, especially in severely injured
-patients where it can be difficult to identify all injuries without
-diagnostic imaging. When clinicians have access to these instruments
-they should be used whenever deemed necessary, as studies suggest it has
-a positive effect on patient outcome [@Hajibandeh_2014].
-
-Further applying the study clinically, it is also important to educate
-clinicians on what steps went wrong in these patients and what could
-have been done to prevent it. If clinicians know the most common
-mistakes made in this patient group, they might be less likely to make
-these mistakes themselves, possibly lowering the risk of receiving
-delayed treatment for these patients with these factors. The results of
-this study and similar ones should be communicated to trauma centers, as
-the goal is to always better the trauma care for patients.
+patient factors that arise from the experienced trauma, not previously
+existing in the patient such as age or gender. This implies that
+patients are not systematically discriminated against based on their
+gender or age, which is a positive sign seen from a health equity point
+of view.
 
@@
 40.7% of that population based on that they had not been screened for
-ofi. Additional patients were excluded because of missing data, but no
+OFI. Additional patients were excluded because of missing data, but no
 distinct groups except the not screened patients were removed. On that
@@
 
-In future studies, various methods to further examine the association
-between delayed treatment and various variables, alternatively using
-another OFI as the outcome variable. Examining different variables such
-as highest hospital care level, intubation status or first emergency
-procedure performed might be interesting variables, among others. To be
-able to do this, more patients in the outcome group would be needed,
-either by expanding the register to several hospitals, or expanding the
-outcome group by adding different ofi. This however would make the study
-less specific for delayed treatment. Further studies might also look
-into different ofi other than delayed treatment, such as missed injury,
-faulty communication or triage.
+To further expand the knowledge on what patient level factors are
+associated with delayed treatment, future studies should aim for a
+larger sample size to be able to analyse more variables, such as highest
+hospital care level, intubation status or first emergency procedure
+performed. To be able to do this, more patients in the outcome group
+would be needed. This can be done by conducting a study which includes
+the registries of several hospitals.
+
+A question that arose during this study was what factors are associated
+with other OFI, such as missed injury, communication errors or faulty
+triage, and if they are similar to the ones found in this study. Future
+studies could use the same data but use another OFI as the outcome to
+compare how variables associated with each OFI differ.
 
@@
 
-In conclusion, based on this single trauma center, the results suggest
-that it is mostly more injured and patients with affected consciousness
-that receive delayed treatment. Clinicians should be aware of this as
-they can take steps to lower that risk for this patient group. Further
+Based on this single trauma center, the results suggest that it is
+mostly more injured and patients with affected consciousness that
+receive delayed treatment. Clinicians should be aware of this as they
+can take steps to lower that risk for this patient group. Further
 studies need to be conducted on this subject to further examine the link
-between delayed treatment and other patient level factors.
+between delayed treatment and patient level factors.
 

### 2024-01-11 manuscriptlongversion.Rmd

@@
 delayed treatment. Material and Methods: This study used the database of
-Karolinska university hospital to run a logistic regression linking the
-outcome, delayed treatment, to variables, observed patient level
+Karolinska university hospital to run a logistic regression associating the outcome, delayed treatment, to variables, observed patient level
 factors. Patients treated for trauma at the hospital between 2012 and
 2022 were included, with ISS\>9 and/or trauma code activation. Exclusion
-criterias were not being screened for OFI and missing values in any of
-the observed variables. Results: Higher ISS and RR was linked to higher
-rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR
-1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
+criteria were not being screened for OFI and missing values in any of
+the observed variables. Results: Higher ISS (OR 1.03, 95% CI 1.01-1.06)and RR (OR 1.04, 95% CI 1.01-1.08) was linked to higher
+rates of delay in treatment while a lower GCS also was linked to delayed
 treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
-correlation to higher rates of delay, but not when adjusted for. The
-remaining variables showed no correlation. Conclusion: Our study
-suggests that it is mostly more severely injured and less awake patients
-that receive delayed treatment.
+correlation to higher rates of delay, but not in the adjusted analysis.
+The remaining variables were not significantly associated with delayed treatment.
+Conclusion: Our study suggests that it is mostly more severely injured patients and patients with affected consciousness that receive delayed treatment.
 
 **Trauma och patientnivåfaktorerna kopplade till försenad behandling**
+
+<!-- See edits above -->
 
@@
 The aim was to determine the associations between patient level factors
-and receiving delayed treatment in adult trauma patients. A logistic
-regression on seven variables showed that higher ISS and RR and lower
+and receiving delayed treatment in adult trauma patients<!-- Make sure this aim is exactly the same as in the abstract and at the end of the introduction-->.
+The results indicate that higher ISS and RR and lower
 GCS was significantly associated with patients receiving delayed
@@
 
-Respiratory rate had the correlation that a lower respiratory rate was
-associated with lower rates of delayed treatment among the patients. As
-the interval is \>10 to \<29 in the triage guideline, both too low and
-too high respiratory rate is not desirable[@EHCP2015]. This could have a
-similar explanation to ISS and GCS, that patients with tachypnea might
-receive higher rates of delayed treatment because they are more severely
-injured, therefore a higher risk of missing a possible injury. The
-previous study also found this, but the correlation disappeared when
-adjusted for, which said study hypothesised being due to its shared
-variability with the ISS variable[@Pat_factors]. In this study, the same
-correlation could be seen with NOI. A more severely injured patient will
-score higher on both ISS and NOI. Therefore, our thought is that the
-effect of NOI becomes insignificant when adjusting for ISS.
+A lower respiratory rate was associated  with lower rates of delayed treatment among the patients.
+As the interval is \>10 to \<29 in the triage guideline, both too low and too high respiratory rate is not desirable[@EHCP2015].
+This could have a similar explanation to ISS and GCS, that patients with tachypnea might have higher rates of delayed treatment because they are more severely injured, therefore a higher risk of missing a possible injury.
+The previous study also found this, but the association was no longer significant in the adjusted analysis for, which said study hypothesised being due to its shared variability with the ISS variable[@Pat_factors].
+In this study, the same correlation could be seen with NOI.
+A more severely injured patient will score higher on both ISS and NOI.
+Therefore, our thought is that the effect of NOI becomes insignificant when adjusting for ISS.
 
@@
 outcome, with higher mortality being seen in patients with lower GCS and
-higher ISS [@Mort_sammy][@Mort_hefny]. Advanced age has also been
+higher ISS [@Mort_sammy; @Mort_hefny]. Advanced age has also been
 identified as a contributory factor to increased mortality. However, it
@@
 patients at the equivalent of a level 1 trauma center, making this study
-generalisable. While the population only represents Karolinska
-university hospital, the conclusion should be applicable to the whole
-trauma population.
+generalisable to similar hospitals in high resource settings.
+While the population only represents Karolinska university hospital, the conclusion should be applicable to the whole trauma population.<!-- Consider removing this last sentence-->
 
@@
 
-I would like to extend my gratitude to my supervisor Martin Gerdin and
-my two co-supervisors Jonatan Attergrim and Kelvin Szolnoky for always
-being quick with providing valuable help and feedback during all stages
-of the writing process as well as guidance during the data analysis part
-of the project.
+I would like to extend my gratitude to my supervisor Martin Gerdin Wärnberg and my two co-supervisors Jonatan Attergrim and Kelvin Szolnoky for always being quick with providing valuable help and feedback during all stages of the writing process as well as guidance during the data analysis part of the project.
